### 2021-03-10 src/manuscript_figures.Rmd
init project with scripts, settings, readme

---
title: "Pan-cancer analysis"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
theme_set(ggpubr::theme_pubclean())

inputDir <- '/data/local/buyar/arcas/subtyping_paper/pancancer/output/'
assay_dir <- '/data/local/buyar/arcas/subtyping_paper/pancancer/assays/'
surv <- do.call(rbind, readRDS('/data/local/buyar/pancancer_diagnostics/data/GDCdata_processed2/TCGA.surv.RDS'))
projectDict <- sapply(yaml::read_yaml('/data/local/buyar/arcas/subtyping_paper/data/projects.dict.yaml'), 
                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
tcga_subtypes <- process_tcga_subtypes()
```

```{r import_lfs}
LFs <- sapply(simplify = F, c('pca_factors', 'maui_factors', 'mofa_factors'), function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'factors.csv$', full.names = T)
  LFs <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- dt$V1
    return(M)
  })
  names(LFs) <- unlist(lapply(strsplit(basename(names(LFs)), split = '\\.'), function(x) paste(x[1:2], collapse = '.')))
  return(LFs)
})
```

```{r import_feature_weights}
FWs <- sapply(simplify = F, c('pca_factors', 'maui_factors', 'mofa_factors'), function(folder) {
  message('Importing ',folder)
  files <- dir(file.path(inputDir, folder), pattern = 'feature_weights.csv$', full.names = T)
  W <- pbapply::pbsapply(simplify = F, files, function(f) {
    dt <- data.table::fread(f)
    M <- as.matrix(dt[,-1])
    rownames(M) <- gsub("^assay: ", "", dt$V1) # required only for maui
    return(M)
  })
  names(W) <- unlist(lapply(strsplit(basename(names(W)), split = '\\.'), 
                            function(x) paste(x[1:2], collapse = '.')))
  return(W)
})
```

```{r}
# Define a function that takes as input LFs and some patient labels (clinical/cancer type/subtype)
# and evaluates predictability of the labels using the LFs
predict_labels <- function(LFs, surv, tcga_subtypes, 
                           project, datatype, label.type, ...) {
  source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
  predict_stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
    require(data.table)
    message("predicting ",label.type," using ",tool)
    experiment <- paste(project, datatype, sep = '.')
    M <- LFs[[tool]][[experiment]]
    if(label.type == 'cancer_type') {
      labels <- data.frame('cancer_type' = surv[match(rownames(M), bcr_patient_barcode)]$project, 
                           row.names = rownames(M))
    } else if (label.type == 'subtype') {
      labels <- data.frame('subtype' = tcga_subtypes[match(rownames(M),
                                                           bcr_patient_barcode)]$Subtype_Selected, 
                           row.names = rownames(M))
    } else if (label.type == 'clinical') {
      labels <- process_clinical_covariates(rownames(M))
    }
    dt <- do.call(rbind, lapply(colnames(labels), function(x) {
      l <- labels[,x]
      if(is.numeric(l)){
        # categorize samples by intervals
        l <- as.character(cut_interval(l, 10))
      }
      l[is.na(l)] <- 'Undefined'
      message("Predicting ",x, " for experiment ",experiment)
      dt <- suppressMessages(predict_cluster_labels(M, l, ...))
      dt$label.type <- x
      dt$project <- project
      dt$datatypes <- datatype
      dt$method <- tool
      return(dt)
    }))
    return(dt)
  }))
  return(predict_stats[!is.na(train_auc)])
}

plot_comp1 <- function(dt) {
  ggboxplot(dt, 
          x = 'method', y = 'test_auc', color = 'method', 
          add = 'jitter', legend = 'none') + 
    stat_compare_means(method = 'wilcox.test', paired = T, method.args = list('alternative' = 'greater'), 
                     comparisons = list(c('maui_factors', 'mofa_factors'), 
                                        c('maui_factors', 'pca_factors'))) +
    labs(y = 'AUC') + theme(axis.text = element_text(size = 10))
}

plot_comp2 <- function(dt) {
  dtc <- melt(dcast(dt, label ~ method, value.var = 'test_auc'), 
              id.vars = c('label', 'maui_factors'))
  dtc$difference <- dtc$maui_factors - dtc$value
  ggplot(dtc, 
         aes(x = maui_factors, y = value)) + 
    geom_point() + geom_abline(slope = 1) + 
    facet_wrap(~ variable, nrow = 1) + 
    labs(y = 'AUC')
}

get_top_by_label <- function(dt, label.type, topN = 5) {
  do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
}

# find percent contribution of different datatypes to the top factor specific variables 
# LFs: Latent factors; nested list
# FWs: Feature weights for each LF; nested list 
# tool: pca/maui/mofa
# project: e.g. pancancer
# datatype: e.g. cnv_gex_mut
# factors: factor vector (e.g. cancer types for each sample or subtype info for each sample)
# top_lfs: top LF to pick per factor
# top_featuers: top features to pick per latent factor
plot_datatype_contribution_by_factor <- function(LFs, FWs, tool, project, datatype, 
                                                 factors, top_lfs = 1, top_features = 100) {
  experiment <- paste0(project, '.', datatype)
  L <- LFs[[tool]][[experiment]]
  W <- FWs[[tool]][[experiment]]
  dt <- get_factor_specific_variables(L, factors)
  top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', topN = top_lfs)
  # for each factor, pick top lfs and pick top features per lf and plot composition
  freq <- do.call(rbind, lapply(split(top_vars, top_vars$ref_cl), function(x) {
    l <- unique(x$variable) # one or more lfs 
    # get top features per lf and concatenate
    f <- do.call(c, lapply(l, function(y) {
      head(sort(abs(W[,y]), decreasing = T), top_features)
    }))
    freq <- data.table(table(sub("^(.+?)\\..+$", "\\1", unique(names(f)))))
    freq$factor <- unique(x$ref_cl)
    colnames(freq)[1] <- 'omics'
    return(freq)
  }))

  ggplot(freq, aes(x = factor, y = N)) + 
    geom_bar(stat = 'identity', aes(fill = omics), 
             position = 'fill') + coord_flip() + 
    labs(x = '', y = 'Percentage of top features contributing to top latent factors')
}
```

```{r pancancer_prediction, fig.height=9, fig.width=12}
# Find out if cancer types are predictable using latent factors 
pancancer_stats <- predict_labels(LFs, surv, tcga_subtypes, 'pancancer', 
                                  'cnv_gex_meth_mut', label.type = 'cancer_type', num.trees = 1000)
p1 <- plot_comp1(pancancer_stats) + theme(axis.text.x = element_blank())
p2 <- plot_comp2(pancancer_stats)
# plot cancer specific factors
M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
p3 <- plot_label_specific_vars(M, ct)

# data type contribution to top LF per factor
p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
                                           'pancancer', 'cnv_gex_meth_mut', factors = ct, 
                                           top_lfs = 1, top_features = 100) 
p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
  theme(legend.position = 'right', legend.background = element_blank())

Fig1 <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B')), 
                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('C', 'D')), 
                  nrow = 2,
                  rel_heights = c(2,3))

print(Fig1)
ggsave('pancancer_cancer_type_prediction.pdf', plot = Fig1, units = 'in', width = 12, height = 9)

# predictability of subtypes using pancancer LFs
pancancer_subtype_stats <- predict_labels(LFs, surv, tcga_subtypes, 'pancancer', 
                                  'cnv_gex_meth_mut', label.type = 'subtype', num.trees = 1000)

p1 <- plot_comp1(pancancer_subtype_stats) 
p2 <- plot_comp2(pancancer_subtype_stats)
# plot subtype specific factors
M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# remove NA values and pick subtypes with at least 20 samples 
selected <- names(which(table(st) > 20))
M <- M[st %in% selected,]
labels <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected

p3 <- plot_label_specific_vars(M, labels)

Fig2 <- plot_grid(plot_grid(p1, p2, nrow = 2), p3, ncol = 2, rel_widths = c(2, 3), 
          labels = 'AUTO')
print(Fig2)

ggsave('pancancer_subtype_prediction.pdf', plot = Fig2, units = 'in', width = 12, height = 9)

# train/test accuracy correlation 
#ggscatter(pancancer_subtype_stats, x = 'train_auc', y = 'test_auc', 
 #         color = 'method', facet.by = 'method', add = 'reg.line')
```

# Unsupervised pan-cancer clustering

We do unsupervised clustering of samples and measure AMI to observe which tool's clusters correspond best to 
ground truth (cancer types)

```{r unsupervised_clustering, fig.height=12, fig.width=10}
get_plot <- function(M, plot.type = 'cancertype') {

  df <- plot_tsne(M)
  if(plot.type == 'cancertype') {
    df$tcga <- surv[match(rownames(df), bcr_patient_barcode)]$project
    df <- df[!is.na(df$tcga),]
  } else if (plot.type == 'subtype') {
    df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
    df <- df[!is.na(df$tcga),]
  }
  clust <- get_kmeans_subtypes(M[rownames(df),], k_vals = length(unique(df$tcga)), 0)
  df$cluster <- as.factor(clust[rownames(df),])

  p <- ggplot(df, aes(x = tSNE1, y = tSNE2))  
  p1 <- p + geom_point(aes(color = cluster)) +
    geom_text_repel(data = get_centroid(df[,1:2], df$cluster), 
              aes(x,y,label = text.label), size = 6) + 
    theme(legend.position = 'none')
  p2 <- p + geom_point(aes(color = tcga)) + 
          geom_text_repel(data = get_centroid(df[,1:2], df$tcga), 
              aes(x,y,label = gsub('TCGA-', '', text.label)), size = 3) +
    theme(legend.position = 'none')

  p <- cowplot::plot_grid(p1, p2)
  # also compute mutual information
  ami <- aricode::AMI(df$tcga, df$cluster)
  return(list('p' = p, 'ami' = ami))
}

pr <-  'pancancer' #'pancancer 99%' # 'nsclc 98%' pan-gastrointestinal 98% 'pan-gyn 98%'
e <- paste0(pr, '.cnv_gex_meth_mut')

#M <- LFs$pca_factors[[e]]
#ct <- surv[match(rownames(M), bcr_patient_barcode)]$project

# for maui, we drop redundant LFs 
p1 <- get_plot(remove_redundant_variables(LFs$maui_factors[[e]], perc = 99), 'cancertype') 
p2 <- get_plot(LFs$pca_factors[[e]], 'cancertype') # use all pcs, 
p3 <- get_plot(LFs$mofa_factors[[e]], 'cancertype') # mofa

# use AMI values to make another plot
p4 <- ggplot(melt(data.frame('MAUI' = p1$ami, 'PCA' = p2$ami, 'MOFA' = p3$ami)), 
       aes(x = variable, y = value)) + 
  geom_bar(aes(fill = variable), stat = 'identity') + 
  geom_text(aes(y = value + 0.05, label = round(value, 2))) + 
  labs(y = 'AMI', x = '') + 
  theme(legend.title = element_blank()) + coord_equal()

Fig <- cowplot::plot_grid(p1$p, p2$p, p3$p, p4, nrow = 4, labels = 'AUTO')
print(Fig)
ggsave("pancancer.unsupervised_clustering.pdf", Fig, units = 'in', width = 10, height = 12)
```

# clustering metrics of various cancers

Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
```{r clustering_metrics, fig.width=10, fig.height=8}
multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      Mr <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
      M <- t(subset_features_by_variance(t(Mr), topN = 10))
     } else {
      M <- M[,head(colnames(M), 10),drop = F]
    }
    val <- c('internal') 
    if(ncol(M) > 2) {val <- c('internal', 'stability')}
    res <- clValid::clValid(obj = M,
                          nClust = 3:8,
                          clMethods = 'kmeans',
                          validation = val,
                          maxitems = 10000,
                          nstart = 1000)
    memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']])))
    colnames(memberships) <- paste0('k', colnames(memberships))
    stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
    colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
    return(list('clustering_stats' = stats, 'cluster_memberships' = memberships))
  })
})
parallel::stopCluster(cl)

cluster.stats <- do.call(rbind, lapply(names(clusters), function(tool) {
  do.call(rbind, lapply(names(clusters[[tool]]), function(experiment) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      datatypes <- strsplit(experiment, "\\.")[[1]][2]
      stats <- clusters[[tool]][[experiment]][['clustering_stats']]
      stats$cancer_type <- cancer_type
      stats$datatypes <- datatypes
      stats$method <- tool
      return(stats)
  }))
}))

# plot silhouette/stability score comparison
dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb]
p1 <- ggboxplot(dt1, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = list(c('maui_factors', 'pca_factors'), 
                                        c('maui_factors', 'mofa_factors'))) + 
  labs(title = 'Silhouette score', x = '') + 
  theme(legend.position = 'none')

dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb]
p2 <- ggboxplot(dt2, 
          x = 'method', y = 'score', add = 'jitter', color = 'method') + 
  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                     comparisons = list(c('maui_factors', 'pca_factors'), 
                                        c('maui_factors', 'mofa_factors'))) + 
  labs(title = 'Cluster Stability Score (APN)', x = '') + 
  theme(legend.position = 'none')

# combine both scores and plot for different values of k
dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
            cancer_type + method + k ~ metric, value.var = 'score')
dt$kv <- paste0('kmeans(k=',dt$k,')')
p3 <- ggplot(dt,
       aes(x = Silhouette, y = APN)) + 
        geom_point(aes(color = method), size = 3) + 
        facet_wrap(~ kv, ncol = 2) +
        labs(title = 'Multi-omics clustering: Silhouette vs APN')

Fig4 <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
                  p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))
print(Fig4)
ggsave('cluster_stability_metrics.pdf', plot = Fig4, units = 'in', 
       width = 10, height = 8)

```

# survival-predictive factors

I look for survival-predictive maui LFs, show how much c-index improves on top of 
clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.

```{r survival_factors, fig.width=12, fig.height=8}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
                  'PFI', 'PFI.time')

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    require(data.table)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
    res <- get_prognostic_covariates(clin, M)
    if(!is.null(res)) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      res$project <- cancer_type
      v <- res[order(pval)][1]$variable
      dat <- merge(clin, M[,v,drop = F], by = 'row.names')
      rownames(dat) <- dat$Row.names
      dat$Row.names <- NULL
      # compute hazard ratios 
      cox1 <- coxph(Surv(PFI.time,PFI) ~ ., 
              data=dat, x = TRUE)
      hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
      hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
      colnames(hr) <- c('hr', 'hr_low', 'hr_high')
      rownames(hr) <- v
      p1 <- survminer::ggforest(cox1, data = dat)

      # split variable into high/low to compute survival adjusted curves
      cp <- find_survival_cutpoint(dat[,v,drop=F], surv)
      dat[,v] <- ifelse(dat[,v] > cp, paste(v, 'high'), paste(v, 'low'))
      cox2 <- coxph(Surv(PFI.time,PFI) ~ ., 
                    data=dat, x = TRUE)
      p2 <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')

      # compute Cindex with all survival-predictive LFs and without LFs 
      vars <- head(res[order(pval)]$variable, 10)
      rsf_stats <- fit_RSF_clin(clin, M[,vars], ntree = 1000)

      return(list('stats' = res, 'p1' = p1, 'p2' = p2, 'hr' = hr, 
                  'Cindex_base' = rsf_stats$C_wo, 'Cindex_with_LFs' = rsf_stats$C_with))
    }
    return(NULL)
  })
parallel::stopCluster(cl)

plots <- sapply(simplify = F, names(surv.stats), function(x) {   
  p <- surv.stats[[x]][['p2']] 
  if(!is.null(p)) { 
    p + theme(axis.text.x = element_text(size = 8), 
              axis.text.y = element_text(size = 8), 
              axis.title.x = element_blank(), 
              axis.title.y = element_blank(), 
              legend.title = element_blank(),
              legend.position = c(0.5, 0.9),
              legend.text = element_text(size = 8), 
              legend.background = element_blank(), 
              legend.direction = 'vertical') 
    }
})
plots <- plots[!sapply(plots, is.null)]
p1 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
                   labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
                   label_size = 8)
print(p1)
ggsave(filename = 'surv.adjustedcurves.prog_lfs.pdf', plot = p1, width = 12, height = 8, units = 'in')

# plot hazard ratios for top LF per cancer type 
hr <- do.call(rbind, lapply(names(surv.stats), function(x) {
  if(is.null(surv.stats[[x]])) {return(NULL)}
  message(x)
  df <- surv.stats[[x]][['hr']]
  df$project <- unlist(strsplit(x, '\\.'))[[1]]
  return(df)
}))
hr$latent_factor <- rownames(hr)
# remove hr with Inf values
hr <- hr[!is.infinite(hr$hr_high),]

p2 <- ggplot(hr, aes(x = reorder(paste0(project, ' (', latent_factor, ')'), hr), y = log(hr))) +
  geom_bar(stat = 'identity') + 
  geom_errorbar(aes(ymin = log(hr_low), ymax = log(hr_high)), width = 0.2) + 
  #geom_text(aes(label = latent_factor, y = ifelse(hr > 1, log(hr+0.005), log(hr-0.005))), 
   #         nudge_x = 0.25) + 
  labs(x = '', y = 'log(hazard-ratio) of top latent factor per cancer\n(adjusted for tumor stage + age + gender)')+
  coord_flip()

# make a plot to show improvement in C index with LFs on top of clinical factors
ci <- do.call(rbind, lapply(names(surv.stats), function(x) {
  if(is.null(surv.stats[[x]])) {return(NULL)}
  df <- data.frame('base' = surv.stats[[x]][['Cindex_base']], 
                   'with_lf' = surv.stats[[x]][['Cindex_with_LFs']], 
                   'project' = unlist(strsplit(x, '\\.'))[[1]])
  return(df)
}))

dt <- melt(ci, id.vars = 'project')
dt$variable <- ifelse(dt$variable == 'base', 'Only clinical factors', 'Clinical factors + MAUI LFs')

p3 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
  geom_segment(aes(x = reorder(project, with_lf), xend = project, 
                   y = base, yend = with_lf)) + 
  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 6) +
  coord_flip() + 
  labs(x = '', y = '') + 
  theme(legend.title = element_blank(), legend.position = 'bottom', 
        legend.background = element_blank(), 
        legend.margin = margin(), 
        legend.direction = 'vertical', legend.box.background = element_blank()) +
  scale_color_brewer(type = 'qual', palette = 6)

p <- cowplot::plot_grid(p3, p2, nrow = 1, labels = 'AUTO')

print(p)
ggsave(filename = 'surv.Cindex_improv.hazard_ratio.pdf', plot = p, width = 12, height = 8, units = 'in')

```

# clinical factor associated MAUI LFs

Find maui LFs associated to clinical factors and plot them

```{r clinical_factors, fig.width=12, fig.height=9}
experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
# find clinical-factor associated LFs 
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('experiments', 'LFs'))
clin.stats <- do.call(rbind, 
        pbapply::pbsapply(cl = cl, simplify = F, 
               experiments, function(experiment) {
               require(data.table)
               source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
               M <- LFs$maui_factors[[experiment]]
               clin <- process_clinical_covariates(rownames(M))
               dt <- do.call(rbind, 
                    pbapply::pbsapply(simplify = F, 
                                      colnames(clin), function(x) {
                      if(!is.numeric(clin[,x])) {
                        factors <- clin[,x]
                        m <- M[!is.na(factors),]
                        factors <- factors[!is.na(factors)]
                        dt <- get_factor_specific_variables(m, factors) #compute_anova(m, factors)
                        dt$label <- x
                        return(dt)
                      }
                      return(NULL)
                      }))
               dt$project <- unlist(strsplit(experiment, "\\."))[[1]]
               return(dt)
             }))
parallel::stopCluster(cl)

# find top factors per project per label and plot
dt <- clin.stats[padj < 0.05, .SD[which.min(padj)],by = c('project', 'label')]

require(ggrepel)
dt$label <- stringr::str_wrap(gsub('_', ' ', dt$label), width = 8)
p1 <- ggplot(dt, aes(x = project, y = label)) + 
  geom_point(aes(color = -log10(padj), size = -log10(padj))) + 
  geom_text_repel(aes(label = variable)) + 
  coord_flip() +
  scale_color_gradient(low = 'blue', high = 'red') + 
  labs(x = '', y = 'clinical factor')

# pick example cases and show more detailed LF activation 
# top examples by clinical factor type
# top 5 variables per clinical factor

top <- clin.stats[padj < 0.05,.SD[which.min(padj)], by = label][order(padj),.SD[which.min(padj)],by = project]
plots <- sapply(simplify = FALSE, 1:4, function(i) {
  pr <- top$project[i]
  l <- top$label[i]
  M <- LFs$maui_factors[[paste0(pr, '.cnv_gex_meth_mut')]]
  factors <- process_clinical_covariates(rownames(M))[[l]]
  M <- M[!is.na(factors),]
  factors <- factors[!is.na(factors)]
  plot_label_specific_vars(M, factors, string_width = 24)+ #, top_vars = top_vars) + 
    labs(title = pr, subtitle = l) + 
    theme(legend.position = 'none')
})

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
p <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
print(p)
ggsave(filename = 'clinical_factors.LFs.pdf', p, units = 'in', width = 12, height = 9)
```

# Predicting MSI status using latent factors
```{r msi_stats, fig.width=10, fig.height=8}
# predict MSI status for pan-gastrointestinal tumors
# read MSI data
dataDir <- '/data/local/buyar/pancancer_diagnostics/data/GDCdata_prepared'
MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
  f <- file.path(dataDir, paste0(pr, '.MSI.tsv'))
  dt <- data.table::fread(f)[,c(2,4)]
  colnames(dt)[2] <- 'MSI'
  dt$project <- pr
  return(dt)
}))
# remove samples with undertermined MSI status
MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]
msi_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
  y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, partition = 0.6, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  #res <- predict_labels_glmnet(M, y)
  res$method <- tool
  return(res)
}))
msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))
# with glmnet and error bars 
# for supplements (includes other tools)
# p1 <- ggplot(msi_stats, aes(x = method, y = auc)) + 
#   geom_bar(aes(fill = method), stat = 'identity') + 
#   geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2) + 
#   labs(y = 'AUC',x = '') + theme(legend.title = element_blank()) + 
#   facet_grid(~ label)

p1 <- ggplot(msi_stats[method == 'MAUI'], aes(x = label, y = auc)) +
  geom_bar(stat = 'identity', aes(fill = label)) + 
  labs(y = 'AUC', x = '') +
  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2)+
  theme(legend.title = element_blank())

# make some maui specific plots to explore MSI related LFs
M <- remove_redundant_variables(LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']], 
                                perc = 99) # make it consistent with pan-cancer
y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI 
M <- M[!is.na(y),]
y <- y[!is.na(y)]

# tsne plot of top LF related to MSI-H and MSI classes
dt <- get_factor_specific_variables(M, y)
# use top predictive lfs to guide tsne 
top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)[['variable']])

df <- plot_tsne(M[,top_vars], y, returnData = T)
# color tsne by top lf informative for MSI-H
top_lf <- dt[order(padj)][ref_cl == 'MSI-H']$variable[1]
df <- merge(df, data.frame(M[,top_lf,drop =F]), by = 'row.names')
df$cancertype <- surv[match(df$Row.names, bcr_patient_barcode)]$project
df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected

p2.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
  theme(legend.title = element_blank())
p2.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes_string(color = top_lf)) + 
  scale_color_gradient(low = 'gray', high = 'red')
p2.3 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(cancertype))) + 
  theme(legend.title = element_blank())
p2.4 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(subtype))) + 
  theme(legend.title = element_blank())

print(cowplot::plot_grid(p2.1, p2.2, p2.3, p2.4, ncol = 2, labels = 'AUTO'))
# top features that contribute to top factors relevant to MSI status prediction
f <- data.table::fread(file.path(inputDir, 'maui_factors', 
                               paste('pan-gastrointestinal', # project 
                                      'cnv_gex_meth_mut', #data type
                                      'maui_factors', # tool
                                      'feature_weights.csv', sep = '.')))
feat.M <- as.matrix(f[,-1])
rownames(feat.M) <- gsub('assay: ', '', f$V1)

# sort features by contribution to LF and plot by data type 
df <- data.frame(feat.M[,top_lf,drop = F])
df$data.type <- sub('^(.+?)\\..+$', '\\1', rownames(df))

# get top 100 features
df <- df[order(abs(df[,top_lf]), decreasing = T),]
freq <- do.call(rbind, lapply(0:29, function(i) {
  s <- (i*100+1)
  e <- (i+1)*100
  x <- data.frame(table(df[s:e,]$data.type))
  x$rank <- stringr::str_wrap(paste0("[",s,",",e,"]"),
                              width = 8)
  x$ind <- i
  return(x)
}))
p3 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
  geom_bar(aes(fill = Var1), stat='identity') + theme(legend.title = element_blank()) + 
  labs(x = paste0("top features contributing to ",top_lf))

p <- cowplot::plot_grid(p1, p3, p2.1, p2.2, labels = 'AUTO')
print(p)
ggsave(filename = 'MSI.status.pdf', p, units = 'in', width = 10, height = 8)

```

# Semi-supervised clustering of samples

1. Cluster samples by LFs associated to known TCGA subtypes or those that are associated to survival outcomes. 
2. Annotate each cluster using cancer state scores or hallmark gene set scores

## TCGA subtype informed clusters

### Use-case: pan-gastrointestinal cancers
```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
require(survminer)
require(cluster)
tcga_subtypes <- process_tcga_subtypes()
# subtype specific LFs
get_top_by_label <- function(dt, label.type, topN = 5) {
  do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
}

pr <- 'pan-gastrointestinal' #'TCGA-BRCA' # 'glioma' 
experiment <- paste0(pr, '.cnv_gex_meth_mut')
M <- LFs$maui_factors[[experiment]]
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# find subtype specific LFs 
dt <- get_factor_specific_variables(M, st)

# use gap statistic to get clusters
vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 3)[['variable']])
# use gap statistic to find optimum clusters 
#r1 <- cluster::clusGap(M[,vars], FUNcluster = kmeans, K.max = 20, iter.max = 50, B = 100)
#plot(r1)
#k <- cluster::maxSE(r1$Tab[,3], r1$Tab[,4], method = 'firstSEmax', SE.factor = 1)
k <- length(na.omit(unique(st)))
# cluster samples using only subtype specific factors 
clust <- data.frame(get_kmeans_subtypes(M[,vars], k_vals = k, min_cluster_size = 0))

# plots
df <- plot_tsne(M[,vars], as.factor(clust[,1]), returnData = T)
colnames(df)[3] <- 'maui'
df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
df$cancertype <- surv[match(rownames(df), bcr_patient_barcode)]$project

p <- ggplot(df, aes(x = tSNE1, y = tSNE2))  
p1 <- p + geom_point(aes(color = maui)) +
  geom_text_repel(data = get_centroid(df[,1:2], df$maui), 
            aes(x,y,label = text.label), size = 6)
p2 <- p + geom_point(aes(color = tcga)) + 
        geom_text_repel(data = get_centroid(df[,1:2], df$tcga), 
            aes(x,y,label = text.label), size = 5)
p3 <- plot_survival(surv, df[,'maui',drop =F])[['plot']] + theme(legend.position = 'none') + labs(y = 'Progression Free Survival')
p4 <- plot_survival(surv, df[,'tcga',drop =F])[['plot']] + theme(legend.position = 'none') + labs(y = 'Progression Free Survival')

# use hallmark gene sets to annotate cluster specific functions
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))
S <- scores[intersect(rownames(M), rownames(scores)),]
clust$tcga <- tcga_subtypes[match(rownames(clust), bcr_patient_barcode)]$Subtype_Selected

# plot top gene sets
h1 <- plot_label_specific_vars(S, clust[rownames(S), 1], 
                              as_heatmap = T, cellwidth = 10, 
                              cellheight = 12,
                              top_vars_per_label = 2)
h2 <- plot_label_specific_vars(S, clust[rownames(S), 2], 
                              as_heatmap = T, cellwidth = 10, 
                              cellheight = 12, 
                              top_vars_per_label = 2)

# altogether
P <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
                         labels = 'AUTO', ncol = 2)
print(P)
ggsave(filename = paste0('semi_supervised.clustering.', experiment,'.pdf'), 
       P, units = 'in', 
       width = 10, height = 12)

```

## Clustering informed by survival outcomes

Pick survival-predictive LFs and cluster samples based on those
```{r semi_supervised_survival, fig.width=10, fig.height=12}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
                  'PFI', 'PFI.time')

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
                                function(experiment){
    require(data.table)
    require(survminer)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
    res <- get_prognostic_covariates(clin, M)
    # now cluster samples based on only survival predictive factors
    prog_lfs <- unique(res$variable)
    if(length(prog_lfs) < 2) {
      return(NULL)
    }
    clust <- get_spectral_clusters(M[,prog_lfs], k_vals = 2:10)
    # find value of k which minimizes log-rank p-value
    dt <- do.call(rbind, lapply(colnames(clust), function(x) {
      message(x)
      pval <- plot_survival(surv, clust[,x,drop=F], return_pval = T)
      data.table('k' = x, 'pval' = pval)
    }))

    # return clustering with best result
    top <- dt[order(pval)][1]$k
    return(list('prog.lf' = M[,prog_lfs], 'prog.clust' = clust[,top,drop = F]))
})
parallel::stopCluster(cl)

# compare maui prognostic clusters with TCGA labels 
stats <- do.call(rbind, lapply(names(prog.clust), function(experiment) {
  message(experiment)
  if(is.null(prog.clust[[experiment]])) {return(NULL)}
  df <- prog.clust[[experiment]][['prog.clust']]
  colnames(df)[1] <- 'maui'
  df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
  do.call(rbind, lapply(colnames(df), function(x) {
    if(length(unique(df[,x])) < 2) {
      pval <- NA
    } else {
      pval <- plot_survival(surv, df[,x,drop = F], return_pval = T)
    }
    return(data.table('source' = x, 'pval' = pval, 'experiment' = experiment))
  }))
}))

stats$cancertype <- gsub('.cnv_gex_meth_mut', '', stats$experiment)
# which cancers can we improve upon tcga?
p1 <- ggplot(stats, aes(x = cancertype, y = -log10(pval))) +  
  geom_bar(stat = 'identity', aes(fill = source), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  coord_flip() + 
  scale_fill_brewer(type = 'qual', palette = 6)

# use hallmark gene sets to annotate cluster specific functions
#scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))

# use canonical pathways to annotate prognostic groups
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_canonical_pathways/ssgsea.canonical_pathways.RDS'))

# scores geneset scores
# M survival predictive LF matrix
# maui_clust: best survival prognostic clustering of M 
plot_clusts <- function(M, maui_clust, scores) {
  df <- plot_tsne(M, as.factor(maui_clust[,1]), returnData = T)
  colnames(df)[3] <- 'maui'
  #df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected

  p1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
    geom_point(aes(color = maui)) +
    geom_text_repel(data = get_centroid(df[,1:2], df$maui), 
            aes(x,y,label = text.label), size = 6)
  # plot survival for the worst prognostic group, and the rest
  ms <- get_median_survival_by_group(surv, df[,'maui',drop=F])

  # find clusters with worst/best prognosis (by median survival)
  worst <- names(sort(ms)[1])
  best <- names(sort(ms, T)[1])

  p2 <- plot_survival(surv, 
                      df[df$maui %in% c(best, worst), 
                         'maui', drop = F])[['plot']] +
    labs(y = 'PFI') +
    theme(legend.position = 'left', legend.title = element_blank()) + 
    scale_color_brewer(type = 'qual', palette = 6)

  # heatmap of hallmark scores for best/worst survivors
  selected <- intersect(rownames(df[df$maui %in% c(worst, best),]), rownames(S))
  dt <- get_factor_specific_variables(S[selected,], df[selected, 'maui'])
  # select factors 
  dt[padj < 0.05][order(padj)]
  h1 <- plot_label_specific_vars(S[selected,], 
                                 df[selected, 'maui'], 
                                 padj_threshold = 0.1,
                                as_heatmap = T, cellwidth = 15, 
                                top_vars_per_label = 5, cellheight = 15)
  return(list('p1' = p1, 'p2' = p2, 'h1' = h1))
}

# plot examples for which we improve upon tcga
dt <- dcast(stats, experiment ~ source, value.var = 'pval')
dt$improvement <- -log10(dt$maui) - -log10(dt$tcga)
dt <- dt[order(improvement, decreasing = T)]

pr <- 'TCGA-UCEC'
e <- paste0(pr, '.cnv_gex_meth_mut')
M <- prog.clust[[e]]$prog.lf
maui_clust <- prog.clust[[e]]$prog.clust
S <- scores[intersect(rownames(M), rownames(scores)),]
P <- plot_clusts(M, maui_clust, S)

Pc <- cowplot::plot_grid(p1, plot_grid(P$p1, P$p2, nrow = 1, labels = c('B', 'C')), 
                   P$h1$gtable, nrow = 3, labels = c('A', '', 'D'))

print(Pc)
ggsave(filename = paste0('semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
       Pc, units = 'in', 
       width = 10, height = 12)

```

# Classify metastatic samples

Use TCGA-SKCM as case study to see if we can predict which samples are metastatic and which ones are primary
We have multi-omics data from 360 metastatic tumors and 100 primary tumors (distinct group of patients)
```{r, eval = FALSE}
# read batch info to see which patient samples are metastatic
batches <- do.call(rbind, lapply(dir(assay_dir, "TCGA-SKCM.*info.tsv", full.names = T),
                  data.table::fread))
# get sample types from sample barcodes
st <- get_tcga_sample_types(batches$V1)
st$bcr_patient_barcode <- sub("^(.{12}).+$", "\\1", st$barcode)
# primary tumors: TP, metastatic tumors: TM
st <- unique(st[sampletype %in% c('TP', 'TM')][,2:3])

skcm_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['TCGA-SKCM.cnv_gex_meth_mut']]
  y <- st[match(rownames(M), bcr_patient_barcode)]$sampletype
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  # plot_tsne(M, y)
  res$method <- tool
  return(res)
}))

# with glmnet 
ggplot(skcm_stats[label == 'TM'], aes(x = method, y = auc)) + 
  geom_bar(aes(fill = method), stat = 'identity') + 
  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2) + 
  labs(y = 'AUC',x = '') + theme(legend.title = element_blank())

# plot maui LFs predictive of metastatic/primary tumor
M <- LFs$maui_factors[['TCGA-SKCM.cnv_gex_meth_mut']]
y <- st[match(rownames(M), bcr_patient_barcode)]$sampletype
M <- M[!is.na(y),]
y <- y[!is.na(y)]
dt <- get_factor_specific_variables(M, y)
top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', topN = 5)[['variable']]

plot_tsne(M[,top_vars], y)

df <- data.frame(M[,top_vars])
df$y <- as.factor(y)
fit <- ranger::ranger(seed = 42, formula = y ~ ., data = df, probability = T, num.trees = 1000)

cp <- data.frame(fit$predictions)
rownames(cp) <- rownames(df)
cp$type <- st[match(rownames(cp), bcr_patient_barcode)]$sampletype

# find metastatic samples that look like primary tumor samples
d1 <- cp[cp$type == 'TM' & cp$TM > 0.7,]
d1$confident <- 'confident_metastatic'
d2 <- cp[cp$type == 'TM' & cp$TM < 0.7,]
d2$confident <- 'not_confident_metastatic'
plot_survival(surv, rbind(d1, d2)[,'confident',drop = F], duration_col = 'OS.time',
              event_col = 'OS')

# check cancer state scores to see if unconfident metastatic samples have similar scores as primary ones
dataDir <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore/msigdb_canonical_pathways/'
files <- dir(dataDir, '.scores.csv', full.names = T)
scores <- do.call(rbind, sapply(simplify = F, files, function(x) {
 data.table::fread(x)
}))
colnames(scores)[1] <- 'bcr_patient_barcode'

cp$group <- 'primary'
cp[cp$type == 'TM' & cp$TM >= 0.7,]$group <- 'confident_metastatic'
cp[cp$type == 'TM' & cp$TM < 0.7,]$group <- 'not_confident_metastatic'

# get cancer state scores
dt <- scores[match(rownames(cp), bcr_patient_barcode),]
M <- as.matrix(dt[,-1])
rownames(M) <- dt$bcr_patient_barcode

dt <- get_factor_specific_variables(M, as.factor(cp$group))

B <- get_basis_matrix(M, cp$group)
#B2 <- apply(B, 2, function(x) x/max(x)+max(x)/4)
#pheatmap(B2)
pheatmap(B[,unique(dt[padj < 0.05]$variable)])

# what are the genes that distinguish confident metastatic from 
M <- LFs$maui_factors$`TCGA-SKCM.cnv_gex_meth_mut`
plot_label_specific_vars(M, cp$group, top_vars_per_label = 5)
dt <- get_factor_specific_variables(M, cp$group)
top <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)

```

# Show subtypes depend on the objective

Patients clustering depend on the objective of stratification

```{r}
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')

project <- 'pan-gastrointestinal'
datatype <- 'cnv_gex_meth_mut'
experiment <- paste0(project, '.', datatype)

M <- LFs$maui_factors[[experiment]]
M <- t(subset_features_by_variance(t(remove_redundant_variables(M)), topN = 20))

clin <- process_clinical_covariates(rownames(M))
clin <- subset(clin, select = intersect(colnames(clin), clin_factors))

pfi_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'PFI', duration_col = 'PFI.time')
pfi_clust <- get_kmeans_subtypes(M[,unique(pfi_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
pfi_k <- names(which.min(sapply(colnames(pfi_clust), function(x) plot_survival(surv, pfi_clust[,x,drop = F], 'PFI.time', 'PFI', T))))

os_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'OS', duration_col = 'OS.time')
os_clust <- get_kmeans_subtypes(M[,unique(os_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
os_k <- names(which.min(sapply(colnames(os_clust), function(x) plot_survival(surv, os_clust[,x,drop = F], 
                                                                             'OS.time', 'OS', T))))

# combine with known pan-gi subtype annotation

df <- merge(pfi_clust[,pfi_k,drop = F], os_clust[,os_k,drop = F], by = 'row.names')
df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected

tsne <- plot_tsne(remove_redundant_variables(M), returnData = T)

df$tSNE1 <- tsne[df$Row.names,]$tSNE1
df$tSNE2 <- tsne[df$Row.names,]$tSNE2

mdf <- melt(df, id.vars = c('Row.names', 'tSNE1', 'tSNE2'))

ggplot(mdf, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes(color = as.factor(value))) + 
  facet_wrap(~ variable, nrow = 1)
```

# predict mutational signatures
```{r}
# import mutational signature data
mutsig <- data.table::fread('/data/local/buyar/pancancer_diagnostics/data/mutational_signatures/signature_profile_sample.txt')
mutsig$Tumor_Sample_Barcode <- gsub("\\.", "-", mutsig$Tumor_Sample_Barcode)
# only keep TCGA samples
mutsig <- mutsig[grep('TCGA', Tumor_Sample_Barcode)]
dt <- data.table::dcast(mutsig, Tumor_Sample_Barcode ~ Signature, value.var = 'Contribution')

cl <- parallel::makeCluster(30) 
parallel::clusterExport(cl, varlist = c('dt', 'LFs'))
res <- do.call(rbind, pbapply::pbsapply(simplify = F, cl = cl, unique(mutsig$Signature), function(x) {
  require(data.table)
  source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
  r <- do.call(rbind, lapply(names(LFs), function(m) {
    message(m)
    M <- LFs[[m]]$pancancer.cnv_gex_meth_mut
    s <- predict_numerical_glmnet(M, dt[match(rownames(M), Tumor_Sample_Barcode)][[x]])
    s$tool <- m
    return(s)
  }))
  r$sig <- x
  return(r)
}))
parallel::stopCluster(cl)

ggplot(res, aes(x = reorder(sig, Rsquare), y = Rsquare)) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') + 
  coord_flip() + labs(x = '')

df <- melt(dcast(res, sig ~ tool, value.var = 'Rsquare'), 
                id.vars = c('sig', 'maui_factors'))

ggplot(df, aes(x = maui_factors, y = value)) + geom_point() + 
  facet_wrap(~ variable, nrow = 1) + 
  geom_abline(slope = 1)
```

# predictability of clinical factors

```{r, eval = FALSE}
# predict tumor purity using latent factors 
res <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pancancer.cnv_gex_meth_mut']]
  y <- process_clinical_covariates(rownames(M))[,'tumor_purity_IHC']
  res <- predict_numerical_glmnet(M, y)
  res$method <- tool
  return(res)
}))

# predict tumor stage for pan-gi 
pangi_stats <- predict_labels(LFs, surv, tcga_subtypes, 
                              'pan-gastrointestinal', 'cnv_gex_meth_mut', label.type = 'clinical')
ggplot(pangi_stats[label.type == 'ajcc_pathologic_tumor_stage'][label != 'Undefined'], aes(x = label, y = test_auc)) + 
  geom_bar(aes(fill = method), stat = 'identity', position = 'dodge') +
  coord_flip()

# tumor stage 
label <- 'age_at_initial_pathologic_diagnosis'
experiment <- 'TCGA-BRCA.cnv_gex_meth_mut'
stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][[experiment]]
  y <- process_clinical_covariates(rownames(M))[[label]]
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y)
  res <- predict_numerical_glmnet(M, y)
  res$method <- tool
  return(res)
}))
# with glmnet multi-label pred
ggplot(stats, aes(x = method, y = Accuracy)) + 
  geom_bar(stat = 'identity', aes(fill = method)) + 
  geom_errorbar(aes(ymin = AccuracyLower, ymax = AccuracyUpper), width = 0.2)

```

# predictability of cancer states

```{r, eval = F}
# read cancer sea gene set scores
#dataDir <- '/data/local/buyar/pancancer_diagnostics/data/geneset_scores/cancersea'
#files <- dir(dataDir, '.scores.csv', full.names = T)
#scores <- do.call(rbind, sapply(simplify = F, files[1:3], function(x) {
#  data.table::fread(x)
#}))
#colnames(scores)[1] <- 'bcr_patient_barcode'

scores <- readRDS('/data/local/buyar/arcas/subtyping_paper/data/mogsa_scores/MOGSA_scores.RDS')
scores <- do.call(rbind, lapply(scores$cnv_gex_meth_mut, function(x) data.table(x, keep.rownames = T)))
colnames(scores)[1] <- 'bcr_patient_barcode'

# elastic net regression on pathway scores usign LFs as input
predict_scores <- function(scores, LFs, label, experiment)
  do.call(rbind, lapply(names(LFs), function(tool) {
    message(tool)
    M <- LFs[[tool]][[experiment]]
    y <- scores[match(rownames(M), bcr_patient_barcode)][[label]]
    M <- M[!is.na(y),]
    y <- y[!is.na(y)]
    res <- predict_numerical_glmnet(M, y)
    res$method <- tool
  return(res)
}))

labels <- c('ANGIOGENESIS', 'GLYCOLYSIS', 'COAGULATION', 'DNA_REPAIR', 
            'HYPOXIA', 'ESTROGEN_RESPONSE_EARLY', 'ESTROGEN_RESPONSE_LATE' ,
            'EPITHELIAL_MESENCHYMAL_TRANSITION')
brca_scores <- do.call(rbind, lapply(labels, function(l) {
  message(l)
  df <- predict_scores(scores, LFs, l, 'TCGA-BRCA.cnv_gex_meth_mut')
  df$label <- l
  return(df)
}))

ggplot(brca_scores, aes(x = label, y = Rsquare)) + 
  geom_bar(stat = 'identity', aes(fill = method), position = 'dodge') + 
  coord_flip()
```

# pan-cancer survival factors 
```{r, eval = FALSE}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
                  'PFI', 'PFI.time')
experiment <- 'pancancer.cnv_gex_meth_mut'
res <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][[experiment]]
  clin <- process_clinical_covariates(rownames(M))
  clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
  cl <- parallel::makeCluster(50)
  parallel::clusterExport(cl, varlist = c('clin', 'M'))
  res <- do.call(rbind, pbapply::pblapply(cl = cl, 1:100, function(i) {
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    require(data.table)
    s <- sample(rownames(M), 0.5 * nrow(M))
    dt <- get_prognostic_covariates(clin[s,], M[s,])
    dt$i <- i
    return(dt)
  }))
  parallel::stopCluster(cl)
  res$method <- tool
  return(res)
}))

# get top factor per method
plots <- sapply(simplify = F, names(LFs), function(tool) {
  M <- LFs[[tool]][[experiment]]
  top_fac <- res[method == tool][,length(unique(i)),by = 'variable'][order(V1, decreasing = T)]$variable[1:3]
  cowplot::plot_grid(plotlist = lapply(top_fac, function(x) {
     p <- plot_survival(surv, data.frame('cl' = ifelse(M[,x] > median(M[,x]), 
                                             'high', 'low')))
     p$plot
  }), nrow = 1)
})
dev.off()
cowplot::plot_grid(plotlist = plots, nrow = 3, labels = names(plots))

M <- LFs[[tool]][[experiment]]
plot_tsne(M, M[,'LF32'])
```

<!-- Cluster samples into k = number of cancer types in the project -->

<!-- ```{r} -->
<!-- cl <- parallel::makeCluster(20) -->
<!-- parallel::clusterExport(cl, varlist = c('LFs', 'projectDict')) -->
<!-- clusters <- sapply(simplify = F, names(LFs), function(tool) { -->
<!--   message("Clustering top ",tool) -->
<!--   pbapply::pbsapply(cl = cl, simplify = F, names(LFs[[tool]]), function(experiment) { -->
<!--     message(experiment) -->
<!--     project <- strsplit(experiment, '\\.')[[1]][1] -->
<!--     datatype <- strsplit(experiment, '\\.')[[1]][2] -->
<!--     source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R') -->
<!--     require(data.table) -->
<!--     M <- LFs[[tool]][[experiment]] -->
<!--     if(tool == 'maui_factors') { -->
<!--       M <- t(subset_features_by_variance(t(M), topN = 10)) -->
<!--     } else { -->
<!--       M <- M[,head(colnames(M), 10),drop = F] -->
<!--     } -->
<!--     val <- c('internal')  -->
<!--     #if(ncol(M) > 2) {val <- c('internal', 'stability')} -->
<!--     #k <- length(projectDict[[project]]) -->
<!--     res <- clValid::clValid(obj = M, -->
<!--                           nClust = 3:10, -->
<!--                           clMethods = 'kmeans', -->
<!--                           validation = val, -->
<!--                           maxitems = 10000, -->
<!--                           nstart = 1000) -->
<!--     memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']]))) -->
<!--     colnames(memberships) <- paste0('k', colnames(memberships)) -->
<!--     stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')] -->
<!--     colnames(stats) <- c('metric', 'k', 'clustering_method', 'score') -->
<!--     return(list('clustering_stats' = stats, 'cluster_memberships' = memberships)) -->
<!--   }) -->
<!-- }) -->
<!-- parallel::stopCluster(cl) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- stats <- do.call(rbind, lapply(names(clusters), function(tool) { -->
<!--   do.call(rbind, lapply(names(clusters[[tool]]), function(experiment) { -->
<!--       cancer_type <- strsplit(experiment, "\\.")[[1]][1] -->
<!--       datatypes <- strsplit(experiment, "\\.")[[1]][2] -->
<!--       stats <- clusters[[tool]][[experiment]][['clustering_stats']] -->
<!--       stats$cancer_type <- cancer_type -->
<!--       stats$datatypes <- datatypes -->
<!--       stats$method <- tool -->
<!--       return(stats) -->
<!--   })) -->
<!-- })) -->
<!-- ``` -->

<!-- clustering metrics -->
<!-- ```{r} -->
<!-- ggboxplot(stats[datatypes == 'cnv_gex_mut'], x = 'method', y = 'score',  -->
<!--           add = 'jitter', facet.by = 'metric', color = 'method') -->
<!-- ``` -->

<!-- Find out how much the cluster labels correspond to cancer types  -->
<!-- ```{r} -->
<!-- cl <- parallel::makeCluster(20) -->
<!-- parallel::clusterExport(cl, varlist = c('clusters', 'projectDict', 'surv', )) -->
<!-- # compute mutual information stats (between cluster labels and cancer types) -->
<!-- MI_stats <- do.call(rbind, sapply(simplify = F, names(clusters), function(tool) { -->
<!--   message("computing MI for ",tool) -->
<!--   do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, names(clusters[[tool]]), -->
<!--                                    function(experiment) { -->
<!--     message(experiment) -->
<!--     project <- strsplit(experiment, '\\.')[[1]][1] -->
<!--     datatype <- strsplit(experiment, '\\.')[[1]][2] -->
<!--     source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R') -->
<!--     require(data.table) -->
<!--     k <- stats[method == tool][cancer_type == project][datatypes == datatype][,.SD[which.max(score)]]$k -->
<!--     clust <- clusters[[tool]][[experiment]][['cluster_memberships']][,paste0('k',k),drop=F] -->
<!--     labels <- surv[match(rownames(clust), bcr_patient_barcode)]$project -->

<!--     data.table('method' = tool, 'project' = project, 'datatypes' = datatype,  -->
<!--                'cramersV' = compute_cramers_V(cbind(clust, labels))[['cramersV']],  -->
<!--                'ARI' = aricode::ARI(clust[,1], labels),  -->
<!--                'AMI' = aricode::AMI(clust[,1], labels)) -->
<!--                                    })) -->
<!--   })) -->
<!-- parallel::stopCluster(cl) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ggboxplot(MI_stats[datatypes == 'cnv_gex_mut'],  -->
<!--           x = 'method', y = 'AMI',  -->
<!--           add = 'jitter') +  -->
<!--     stat_compare_means(method = 'wilcox.test', paired = T, method.args = list('alternative' = 'greater'),  -->
<!--                      comparisons = list(c('maui_factors', 'mofa_factors'),  -->
<!--                                         c('maui_factors', 'pca_factors'))) + -->
<!--   labs(title = 'One-vs-all test-AUC by cancer type') -->

<!-- ``` -->

<!-- Unsupervised clustering experiment -->
<!-- ```{r} -->
<!-- get_plot <- function(M, plot.type = 'cancertype') { -->

<!--   df <- plot_tsne(M) -->
<!--   if(plot.type == 'cancertype') { -->
<!--     df$tcga <- surv[match(rownames(df), bcr_patient_barcode)]$project -->
<!--   } else if (plot.type == 'subtype') { -->
<!--     df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected -->
<!--   } -->
<!--   clust <- get_kmeans_subtypes(M[rownames(df),], k_vals = length(unique(df$tcga)), 0) -->
<!--   df$maui <- as.factor(clust[rownames(df),]) -->

<!--   p <- ggplot(df, aes(x = tSNE1, y = tSNE2))   -->
<!--   p1 <- p + geom_point(aes(color = maui)) + -->
<!--     geom_text_repel(data = get_centroid(df[,1:2], df$maui),  -->
<!--               aes(x,y,label = text.label), size = 6) -->
<!--   p2 <- p + geom_point(aes(color = tcga)) +  -->
<!--           geom_text_repel(data = get_centroid(df[,1:2], df$tcga),  -->
<!--               aes(x,y,label = text.label), size = 5) -->

<!--   p <- cowplot::plot_grid(p1, p2) -->
<!--   # also compute mutual information -->
<!--   ami <- aricode::AMI(df$tcga, df$maui) -->
<!--   return(list('p' = p, 'ami' = ami)) -->
<!-- } -->

<!-- pr <- 'pancancer' #'pancancer' # 'pan-gastrointestinal' -->
<!-- e <- paste0(pr, '.cnv_gex_meth_mut') -->
<!-- M <- LFs$maui_factors[[e]] -->
<!-- ct <- surv[match(rownames(M), bcr_patient_barcode)]$project -->
<!-- st <-  tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected -->

<!-- dt.ct <- get_factor_specific_variables(M, ct) -->
<!-- ct_vars <- unique(get_top_by_label(dt.ct[padj < 0.05][order(padj)], label.type = 'ref_cl', 2)[['variable']]) -->
<!-- dt.st <- get_factor_specific_variables(M, st) -->
<!-- st_vars <- unique(get_top_by_label(dt.st[padj < 0.05][order(padj)], 'ref_cl', 2)[['variable']]) -->

<!-- # LFs with most variance -->
<!-- top_vars <- rownames(subset_features_by_variance(t(M), topN = 42)) -->
<!-- # using top vars will cause the clusters dominated by the major component to dominate -->

<!-- p1 <- get_plot(M, 'cancertype') # use all factors  (unsupervised) -->
<!-- p2 <- get_plot(M[,top_vars], 'cancertype') # use top most variable factors (unsupervised) -->
<!-- p3 <- get_plot(M[,ct_vars], 'cancertype') # use cancer type specific factor (supervised ) -->
<!-- p4 <- get_plot(remove_redundant_variables(M, cutoff = 0.6), 'cancertype') # (unsupervised) -->
<!-- p5 <- get_plot(LFs$pca_factors[[e]]) # pca -->
<!-- p5.1 <- get_plot(LFs$pca_factors[[e]][,1:10]) # pca -->
<!-- p5.2 <- get_plot(LFs$pca_factors[[e]][,1:20]) # pca -->
<!-- p5.4 <- get_plot(LFs$pca_factors[[e]][,1:40]) -->

<!-- p6 <- get_plot(LFs$mofa_factors[[e]]) # mofa -->

<!-- ps <- lapply(c(5:8/10), function(r) { -->
<!--   get_plot(remove_redundant_variables(M, cutoff = r), 'cancertype') -->
<!-- }) -->

<!-- dt <- do.call(rbind, lapply(experiments, function(e) { -->
<!--   x <- LFs$maui_factors[[e]] -->
<!--   y <- remove_redundant_variables(x, cutoff = 0.6) -->
<!--   data.table('before' = ncol(x), 'after' = ncol(y), 'n' = nrow(x), 'exp' = e,  -->
<!--              'redundant' = ncol(x) - ncol(y)) -->
<!-- })) -->

<!-- p5 <- ggscatter(dt, x = 'n', y = 'redundant', add = 'reg.line') + -->
<!--    stat_cor(label.x = 400, label.y = 80, size = 6) + -->
<!--   labs(x = 'Number of samples', y = 'Percentage of redundant LFs') + -->
<!--   stat_regline_equation(label.x = 400, label.y = 75, size = 6) +  -->
<!--   geom_text_repel(aes(label = gsub('.cnv_gex_meth_mut', '', exp))) -->

<!-- ``` -->

### 2021-03-10 src/common_functions.R

@@ -1,0 +1,1198 @@
+# common functions used in different analyses
+
+compute_PCA <- function(exp, topN = 50) {
+  top <- head(names(sort(apply(exp, 1, sd), decreasing = T)), 5000)
+  M <- t(exp[top,])
+  pca <- stats::prcomp(M, rank. = topN)
+  return(pca)
+}
+
+# given TCGA identifiers, create a table of batch information
+# see https://docs.gdc.cancer.gov/Encyclopedia/pages/TCGA_Barcode/
+getBatchTable <- function(TCGA_ids) {
+  dt <- data.table(do.call(rbind, lapply(strsplit(TCGA_ids, "-"), function(x) t(data.frame(x)))))
+  colnames(dt) <- c('project', 'TSS', 'participant', 'sample_vial', 'portion_analyte', 'plate', 'center')
+  dt$bcr_patient_barcode <- paste(dt$project, dt$TSS, dt$participant, sep = '-')
+  return(dt)
+}
+
+
+# given a data.table of samples and batch covariates, 
+# simplify the batch table: 
+# 1. to keep only common batches (e.g. combine batches with few samples)
+# 2. remove batch covariates that only contain a single factor 
+process_batch_table <- function(df, min_samples_per_batch = 10) {
+  l <- lapply(colnames(df), function(n) {
+    message(n)
+    x <- df[,get(n),drop = T]
+    if(length(unique(x)) == 1) {
+      return(NULL)
+    }
+    rare_batches <- names(which(table(x) < min_samples_per_batch))
+    x <- ifelse(x %in% rare_batches, 'other', x)
+    if(length(unique(x)) == 1) {
+      return(NULL)
+    }
+    if(sum(table(x) > min_samples_per_batch) < 2) {
+      return(NULL)
+    }
+    x <- data.frame(x, stringsAsFactors = T)
+    colnames(x) <- n
+    return(x)
+  })
+  dt <- data.table(do.call(cbind, l[!unlist(lapply(l, is.null))]))
+  if(ncol(dt) == 0) {
+    return(NULL)
+  }
+  return(dt)
+}
+
+
+# given the survival data and a matrix (patients on rows, features on columns), 
+# return a fit model 
+# colnames in M must match bcr_patient_barcode in surv,
+# using PFI as default outcome
+fit_RSF <- function(surv, M, ntree = 1000, 
+                    duration_col = 'PFI.time', 
+                    event_col = 'PFI') {
+  require(data.table)
+  df <- as.data.frame(M)
[…]

### 2021-03-10 src/run_MCIA.R

@@ -1,0 +1,30 @@
+# https://www.bioconductor.org/packages/release/bioc/html/omicade4.html
+args <- commandArgs(trailingOnly = TRUE)
+library(ggplot2)
+library(omicade4)
+
+Assays <- args[1] # path to RDS object containing a list of assay matrices
+outFile <- args[2] # path to output file
+
+A <- readRDS(Assays)
+
+nf <- ifelse(ncol(A[[1]]) > 1000, 50, 20) # round(max(10, sqrt(ncol(A[[1]]))))
+message(date(), " => Running MCIA looking for ",nf," factors")
+mcoin <- mcia(A, cia.nf=nf)
+
+LF <- mcoin$mcoa$SynVar
+rownames(LF) <- gsub("\\.", "-", rownames(LF))
+
+FeatureWeights <- mcoin$mcoa$axis
+
+# save factors
+write.table(x = LF, 
+            file = outFile,
+            quote = F, row.names = T, sep = ',')
+
+# save feature weights
+write.table(x = FeatureWeights, 
+            file = gsub('.csv$', '.feature_weights.csv', outFile),
+            quote = F, row.names = T, sep = ',')
+
+message(date(), " => Finished MCIA")

### 2021-03-10 src/run_PCA.R

@@ -1,0 +1,14 @@
+args <- commandArgs(trailingOnly = TRUE)
+Assays <- args[1] # path to a CSV file
+outFile <- args[2] # path to output file
+source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
+M <- as.matrix(read.csv(Assays, check.names = FALSE))
+PCA <- compute_PCA(M, topN = 100)
+write.table(x = PCA$x, 
+            file = outFile,
+            quote = F, row.names = T, sep = ',')
+# save feature rotation matrix
+write.table(x = PCA$rotation, 
+            file = gsub('.csv', '.feature_weights.csv', outFile),
+            quote = F, row.names = T, sep = ',')
+

### 2021-03-10 src/run_maui.py

@@ -1,0 +1,29 @@
+import argparse
+import os
+import pandas as pd
+import maui
+from arcaslib import squishy_maui
+
+parser = argparse.ArgumentParser(description='Run MAUI to get latent factors from the given assay matrix.')
+
+parser.add_argument("assay_matrix", help = "Path to CSV format feature x sample matrix")
+parser.add_argument("outFile", help = "Path to output file for latent factors")
+parser.add_argument("Nlatent", help = "Number of latent factors to search for")
+parser.add_argument("Nepochs", help = "Number of epochs to run the training algorithm")
+parser.add_argument("Nthreads", help = "Number of threads used to run maui")
+args = parser.parse_args()
+
+#read assay matrix
+assay = pd.read_csv(args.assay_matrix) 
+print(assay.head())
+#start session
+from keras import backend as K
+K.set_session(K.tf.Session(config=K.tf.ConfigProto(intra_op_parallelism_threads=int(args.Nthreads), inter_op_parallelism_threads=int(args.Nthreads))))
+maui_model = maui.Maui(n_hidden=[1000], n_latent=int(args.Nlatent), epochs=int(args.Nepochs))
+z = maui_model.fit_transform({'assay': assay})
+# save LF to file
+z.to_csv(args.outFile)
+
+# get feature contributions to each LF
+nwp = maui_model.get_neural_weight_product()
+nwp.to_csv(''.join([os.path.splitext(args.outFile)[0], '.feature_weights.csv']))

### 2021-03-10 src/run_maui.sh

@@ -1,0 +1,8 @@
+source activate maui
+export PYTHONPATH=/home/buyar/.conda/envs/maui/lib/python3.6/site-packages
+assays=$1 # assay matrix file path
+outFile=$2 # path to output file
+N_lfs=$3 #number of latent factors
+N_epochs=$4 #nmuber of epochs
+N_threads=$5 # number of threads
+python /data/local/buyar/arcas/subtyping_paper/src/run_maui.py ${assays} ${outFile} ${N_lfs} ${N_epochs} ${N_threads} 

### 2021-03-10 src/run_mofa.R

@@ -1,0 +1,35 @@
+library(reticulate)
+# run this in interactive mode, turn it off for command-line (I know it is stupid)
+# use_python('/home/buyar/.cache/basilisk/1.2.0/MOFA2-1.0.0/mofa_env/bin/python3.7', required = TRUE)
+args <- commandArgs(trailingOnly = TRUE)
+s <- import("sys")
+message("Python version:", s$version)
+library(ggplot2)
+library(MOFA2)
+
+Assays <- args[1] # path to RDS object containing a list of assay matrices
+outFile <- args[2] # path to output file
+topFeatures <- args[3] # number of top features to use in training
+
+A <- readRDS(Assays)
+MOFAobject <- create_mofa(A)
+modelOptions <- MOFA2::get_default_model_options(MOFAobject)
+modelOptions$num_factors <- min(100, floor(median(sapply(A, ncol))/4)) # set to 0.25 of number of samples 
+MOFAobject <- prepare_mofa(MOFAobject, model_options = modelOptions)
+message(date(), " training MOFA")
+MOFAobject.trained <- run_mofa(MOFAobject, use_basilisk = TRUE)
+message(date(), " => Saving factors to file")
+factors <- get_factors(MOFAobject.trained, factors = "all")
+write.table(x = factors[[1]], 
+            file = outFile,
+            quote = F, row.names = T, sep = ',')
+
+# extract weights
+weights <- get_weights(MOFAobject.trained, views = "all", factors = "all")
+write.table(x = do.call(rbind, weights), 
+            file = gsub('.csv$', '.feature_weights.csv', outFile),
+            quote = F, row.names = T, sep = ',')
+
+message(date(), ' => Finished mofa!')
+
+

### 2021-03-10 src/score_genesets_tcga.R

@@ -1,0 +1,141 @@
+# use ssGSEA approach to score each TCGA sample (using gene expression datasets)
+# 
+library(singscore)
+library(data.table)
+source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
+geneSetFolder <- '/data/local/buyar/arcas/subtyping_paper/data/GeneSets'
+ens2hgnc <- readRDS('/data/local/buyar/arcas/subtyping_paper/data/ens2hgnc.RDS') #id mapping 
+GDC_data_dir <- '/data/local/buyar/pancancer_diagnostics/data/GDCdata_prepared' # path to TCGA data 
+outDir <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
+
+# f: path to gmt format file 
+read_msigdb <- function(f) {
+  geneSets <- lapply(readLines(f), function(x) {
+    unlist(strsplit(x, "\t"))
+  })
+  #add names
+  names(geneSets) <- unlist(lapply(geneSets, function(x) x[1]))
+  #remove first two 
+  geneSets <- lapply(geneSets, function(x) x[-c(1:2)])
+  return(geneSets)
+}
+# d: path to folder that contains cancerSEA cell state signature files 
+read_cancersea <- function(d) {
+  files <- dir(d, '.txt$', full.names = T)
+  geneSets <- lapply(files, 
+                               function(f) {
+                                 dt <- data.table::fread(f)
+                                 return(unique(dt$EnsembleID))
+                               })
+  names(geneSets) <- gsub(".txt$", "", basename(files))
+  return(geneSets)
+}
+
+# f: path to file containing immune gene sets used in xCell package 
+read_xCell_geneSets <- function(f, group_by_cell_type = FALSE) {
+  l <- readLines(f)
+  geneSets <- lapply(l[-1], function(x) {
+    res <- unlist(strsplit(gsub(" ", "_", x), "\t"))
+    return(setdiff(res[-c(1:2)], ''))
+  })
+  
+  names(geneSets) <- unlist(lapply(l[-1], function(x) 
+    gsub(" ", "_", unlist(strsplit(x, "\t"))[1])))
+  
+  # there are multiple sources for cell types, which can be grouped together 
+  if(group_by_cell_type == TRUE) {
+    dt <- data.table('name' = names(geneSets), 
[…]

### 2021-03-10 src/setup_experiments.R

@@ -1,0 +1,73 @@
+library(argparser, quietly = TRUE)
+library(data.table)
+
+p <- argparser::arg_parser('Setup assays for the pipeline')
+p <- argparser::add_argument(p, "--settings", help="path to settings file for the experiment")
+argv <- argparser::parse_args(p)
+
+message(date(), " => importing settings from ",file.path(argv$settings))
+settings <- yaml::read_yaml(argv$settings)
+# read utility functions
+source(settings$utility_script)
+
+assayDir <- settings$assay_output$folder
+if(settings$features$use_geneset == TRUE) {
+  ens2hgnc <- readRDS(settings$ens2hgnc)
+  geneSets <- read_msigdb(settings$geneset)
+  geneSets.ensembl <- map_ens2hgnc(geneSets, ens2hgnc, "hgnc2ens")
+  GOI <- unique(unlist(geneSets.ensembl))
+} else {
+  GOI <- NULL
+}
+
+surv <- do.call(rbind, readRDS(settings$surv))
+
+# path to folder with prepared TCGA data
+dataDir <- settings$GDCdataDir
+
+projects <- settings$projects
+projectDict <- sapply(yaml::read_yaml(settings$projectDict), 
+                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
+
+if(!dir.exists(assayDir)) {
+  dir.create(assayDir)
+}
+
+cl <- parallel::makeCluster(length(projects))
+parallel::clusterExport(cl = cl, varlist = c('settings', 'projectDict', 'dataDir', 
+                                             'surv', 'assayDir', 'GOI'))
+pbapply::pblapply(cl = cl, projects, function(pr) {
+  require(data.table)
+  source(settings$utility_script)
+  message(date(), " => Preparing assays for project ", pr, " with cancer types ", paste(projectDict[[pr]], collapse = ',') , 
+          " for data types ",paste(settings$datatypes, collapse = ','))
+  dataList <- prepareData(dataDir = dataDir, clin = surv, correctBatchEffects = settings$correctBatches, 
+                          topN = settings$features$topN, GOI = GOI, TCGA_Disease_IDs = projectDict[[pr]], 
+                          datatypes = settings$datatypes, 
+                          gex_flavor = settings$datatype_flavors$gex) 
+  # save assays 
+  lapply(sapply(settings$datatype_combinations, function(x) strsplit(x, ',')), 
+         function(datatypes) {
+           out <- file.path(assayDir, 
+                            paste0(pr,'.',
[…]

### 2021-03-10 src/snakefile.py

@@ -1,0 +1,68 @@
+
+import os
+ASSAYDIR = '/data/local/buyar/arcas/subtyping_paper/pancancer/assays'
+SRCDIR = '/data/local/buyar/arcas/subtyping_paper/src'
+ASSAYS = set([os.path.splitext(f)[0] for f in os.listdir(ASSAYDIR) if re.match(r'.*\.csv$', f)])
+RSCRIPT = '/opt/R-4.0.2/lib64/R/bin/Rscript'
+OUTDIR = 'output'
+LOG_DIR = os.path.join(OUTDIR, "logs")
+
+rule all:
+    input:
+        expand(os.path.join(OUTDIR, 'maui_factors', "{assay}.maui_factors.csv"), assay = ASSAYS),
+        expand(os.path.join(OUTDIR, 'maui_factors', "{assay}.maui_factors.feature_weights.csv"), assay = ASSAYS),
+        expand(os.path.join(OUTDIR, 'mofa_factors', "{assay}.mofa_factors.csv"), assay = ASSAYS),
+        expand(os.path.join(OUTDIR, 'mofa_factors', "{assay}.mofa_factors.feature_weights.csv"), assay = ASSAYS),
+        expand(os.path.join(OUTDIR, 'pca_factors', "{assay}.pca_factors.csv"), assay = ASSAYS),
+        expand(os.path.join(OUTDIR, 'pca_factors', "{assay}.pca_factors.feature_weights.csv"), assay = ASSAYS),
+	expand(os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.csv"), assay = ASSAYS), 
+	expand(os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.feature_weights.csv"), assay = ASSAYS)
+
+rule run_mcia:
+    input:
+        os.path.join(ASSAYDIR, "{assay}.csv")
+    output:
+        os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.csv"),
+        os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.feature_weights.csv")
+    log: os.path.join(LOG_DIR, "run_mcia.{assay}.log")
+    params:
+        script = os.path.join(SRCDIR, 'run_MCIA.R')
+    shell:
+        "bash {params.script} {input} {output[0]} > {log} 2>&1"
+
+rule run_maui:
+    input:
+        os.path.join(ASSAYDIR, "{assay}.csv")
[…]

### 2021-03-10 src/snakefile.py
fix file paths and path to RSCRIPT

@@ -20,7 +20,7 @@ rule all:
 
 rule run_mcia:
     input:
-        os.path.join(ASSAYDIR, "{assay}.csv")
+        os.path.join(ASSAYDIR, "{assay}.RDS")
     output:
         os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.csv"),
         os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.feature_weights.csv")
@@ -28,7 +28,7 @@ rule run_mcia:
     params:
         script = os.path.join(SRCDIR, 'run_MCIA.R')
     shell:
-        "bash {params.script} {input} {output[0]} > {log} 2>&1"
+        "{RSCRIPT} {params.script} {input} {output[0]} > {log} 2>&1"
 
 rule run_maui:
     input:

### 2021-03-10 src/common_functions.R
add cnv_flavor option

@@ -722,10 +722,12 @@ parseMAF <- function(project = 'TCGA-BRCA',
 # project: Name of TCGA project (e.g TCGA-BRCA)
 # topN: top feature count to pick. For mutations, it is by sum of mutations, for others it is by coefvar.
 # gex_flavor: can be gex_raw or gex.fpkm
+# cnv_flavor: can be "cnv" or "cnv_gistic"
 prepareData <- function(dataDir, clin, correctBatchEffects = TRUE, 
                         topN = 1000, GOI = NULL,
                         TCGA_Disease_IDs, datatypes = c('gex', 'cnv', 'mut', 'meth'),
-                        gex_flavor) {
+                        gex_flavor, 
+                        cnv_flavor) {
   patientIDs <- unique(clin[project %in% TCGA_Disease_IDs]$bcr_patient_barcode)
   dataList <- list('assay' = list(), 
                    'batch' = list())
@@ -821,9 +823,14 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     # get CNV data
     cnv <- lapply(TCGA_Disease_IDs, function(pr) {
       message(date(), " => importing cnv data for ",pr)
-      df.tibble <- readRDS(file.path(dataDir, paste0(pr, '.cnv.RDS')))
-      cnv <- as.matrix(df.tibble[,-c(1:3)])
-      rownames(cnv) <-  gsub("\\.[0-9]+$", "", df.tibble$`Gene Symbol`)
+      dat <- readRDS(file.path(dataDir, paste(pr, cnv_flavor, 'RDS', sep = '.')))
+      if(cnv_flavor == 'cnv') {
+        cnv <- as.matrix(dat[,-c(1:3)])
+        rownames(cnv) <-  gsub("\\.[0-9]+$", "", dat$`Gene Symbol`)
+      } else if (cnv_flavor == 'cnv_gistic') {
+        cnv <- dat
+      }
+      
       return(cnv)
     })
     common_features <- Reduce(intersect, lapply(cnv, rownames))

### 2021-03-10 src/common_functions.R
select cnv features differently based on cnv_flavor used

@@ -830,15 +830,21 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
       } else if (cnv_flavor == 'cnv_gistic') {
         cnv <- dat
       }
-      
       return(cnv)
     })
     common_features <- Reduce(intersect, lapply(cnv, rownames))
     cnv <- do.call(cbind, lapply(cnv, function(x) x[common_features,]))
     
     if(!is.null(GOI)) {cnv <- cnv[intersect(GOI, rownames(cnv)),]}
-    # get top features # pick features which are most often gained/lost
-    selected <- head(names(sort(apply(cnv, 1, function(x) sum(x == 0)))), topN) 
+    
+    selected <- NULL
+    if(cnv_flavor == 'cnv') {
+      # get top features # pick features which are most often gained/lost
+      selected <- head(names(sort(apply(cnv, 1, function(x) sum(x == 0)))), topN) 
+    } else if(cnv_flavor == 'cnv_gistic') {
+      # pick top features by variance in gistic scores
+      selected <- head(names(sort(apply(cnv, 1, var), decreasing = T)), topN) 
+    }
     cnv <- cnv[selected,]
     rownames(cnv) <- paste0('cnv.', rownames(cnv))
     

### 2021-03-10 src/setup_experiments.R
include cnv_flavor as argument

@@ -41,10 +41,13 @@ pbapply::pblapply(cl = cl, projects, function(pr) {
   source(settings$utility_script)
   message(date(), " => Preparing assays for project ", pr, " with cancer types ", paste(projectDict[[pr]], collapse = ',') , 
           " for data types ",paste(settings$datatypes, collapse = ','))
-  dataList <- prepareData(dataDir = dataDir, clin = surv, correctBatchEffects = settings$correctBatches, 
-                          topN = settings$features$topN, GOI = GOI, TCGA_Disease_IDs = projectDict[[pr]], 
+  dataList <- prepareData(dataDir = dataDir, clin = surv, 
+                          correctBatchEffects = settings$correctBatches, 
+                          topN = settings$features$topN, GOI = GOI, 
+                          TCGA_Disease_IDs = projectDict[[pr]], 
                           datatypes = settings$datatypes, 
-                          gex_flavor = settings$datatype_flavors$gex) 
+                          gex_flavor = settings$datatype_flavors$gex, 
+                          cnv_flavor = settings$datatype_flavors$cnv) 
   # save assays 
   lapply(sapply(settings$datatype_combinations, function(x) strsplit(x, ',')), 
          function(datatypes) {

### 2021-03-10 src/snakefile.py
fix path to assay dir

@@ -1,7 +1,7 @@
 
 import os
-ASSAYDIR = '/data/local/buyar/arcas/subtyping_paper/pancancer/assays'
-SRCDIR = '/data/local/buyar/arcas/subtyping_paper/src'
+ASSAYDIR = '/data/local/buyar/arcas/pancancer_multiomics_manuscript/assays' 
+SRCDIR = '/data/local/buyar/arcas/pancancer_multiomics_manuscript/src' 
 ASSAYS = set([os.path.splitext(f)[0] for f in os.listdir(ASSAYDIR) if re.match(r'.*\.csv$', f)])
 RSCRIPT = '/opt/R-4.0.2/lib64/R/bin/Rscript'
 OUTDIR = 'output'

### 2021-03-10 src/run_MCIA.R
look for 20 factors for MCIA

@@ -8,7 +8,7 @@ outFile <- args[2] # path to output file
 
 A <- readRDS(Assays)
 
-nf <- ifelse(ncol(A[[1]]) > 1000, 50, 20) # round(max(10, sqrt(ncol(A[[1]]))))
+nf <- 20
 message(date(), " => Running MCIA looking for ",nf," factors")
 mcoin <- mcia(A, cia.nf=nf)
 

### 2021-03-10 src/manuscript_figures.Rmd
drop analyses that are not part of the manuscript

@@
 knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
-source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
+settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
+source(settings$utility_script)
 library(data.table)
@@
 
-inputDir <- '/data/local/buyar/arcas/subtyping_paper/pancancer/output/'
-assay_dir <- '/data/local/buyar/arcas/subtyping_paper/pancancer/assays/'
-surv <- do.call(rbind, readRDS('/data/local/buyar/pancancer_diagnostics/data/GDCdata_processed2/TCGA.surv.RDS'))
-projectDict <- sapply(yaml::read_yaml('/data/local/buyar/arcas/subtyping_paper/data/projects.dict.yaml'), 
+inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
+assay_dir <- settings$assay_output$folder
+surv <- do.call(rbind, readRDS(settings$surv))
+projectDict <- sapply(yaml::read_yaml(settings$projectDict), 
                       function(x) gsub(' ', '', unlist(strsplit(x, ','))))
@@
 ```
-
-# Classify metastatic samples
-
-Use TCGA-SKCM as case study to see if we can predict which samples are metastatic and which ones are primary
-We have multi-omics data from 360 metastatic tumors and 100 primary tumors (distinct group of patients)
-```{r, eval = FALSE}
-# read batch info to see which patient samples are metastatic
-batches <- do.call(rbind, lapply(dir(assay_dir, "TCGA-SKCM.*info.tsv", full.names = T),
-                  data.table::fread))
-# get sample types from sample barcodes
-st <- get_tcga_sample_types(batches$V1)
-st$bcr_patient_barcode <- sub("^(.{12}).+$", "\\1", st$barcode)
-# primary tumors: TP, metastatic tumors: TM
-st <- unique(st[sampletype %in% c('TP', 'TM')][,2:3])
-
-skcm_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
-  message(tool)
-  M <- LFs[[tool]][['TCGA-SKCM.cnv_gex_meth_mut']]
-  y <- st[match(rownames(M), bcr_patient_barcode)]$sampletype
-  M <- M[!is.na(y),]
-  y <- y[!is.na(y)]
-  #res <- predict_cluster_labels(M, y, num.trees = 2000)
-  res <- predict_cluster_labels_glmnet(M, y)
-  # plot_tsne(M, y)
-  res$method <- tool
-  return(res)
-}))
-
-# with glmnet 
-ggplot(skcm_stats[label == 'TM'], aes(x = method, y = auc)) + 
-  geom_bar(aes(fill = method), stat = 'identity') + 
-  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2) + 
-  labs(y = 'AUC',x = '') + theme(legend.title = element_blank())
-
-# plot maui LFs predictive of metastatic/primary tumor
-M <- LFs$maui_factors[['TCGA-SKCM.cnv_gex_meth_mut']]
-y <- st[match(rownames(M), bcr_patient_barcode)]$sampletype
-M <- M[!is.na(y),]
-y <- y[!is.na(y)]
-dt <- get_factor_specific_variables(M, y)
-top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', topN = 5)[['variable']]
-
-plot_tsne(M[,top_vars], y)
-
-df <- data.frame(M[,top_vars])
-df$y <- as.factor(y)
-fit <- ranger::ranger(seed = 42, formula = y ~ ., data = df, probability = T, num.trees = 1000)
-
-cp <- data.frame(fit$predictions)
-rownames(cp) <- rownames(df)
-cp$type <- st[match(rownames(cp), bcr_patient_barcode)]$sampletype
-
-# find metastatic samples that look like primary tumor samples
-d1 <- cp[cp$type == 'TM' & cp$TM > 0.7,]
-d1$confident <- 'confident_metastatic'
-d2 <- cp[cp$type == 'TM' & cp$TM < 0.7,]
-d2$confident <- 'not_confident_metastatic'
-plot_survival(surv, rbind(d1, d2)[,'confident',drop = F], duration_col = 'OS.time',
-              event_col = 'OS')
-
-# check cancer state scores to see if unconfident metastatic samples have similar scores as primary ones
-dataDir <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore/msigdb_canonical_pathways/'
-files <- dir(dataDir, '.scores.csv', full.names = T)
-scores <- do.call(rbind, sapply(simplify = F, files, function(x) {
- data.table::fread(x)
-}))
-colnames(scores)[1] <- 'bcr_patient_barcode'
-
-cp$group <- 'primary'
-cp[cp$type == 'TM' & cp$TM >= 0.7,]$group <- 'confident_metastatic'
-cp[cp$type == 'TM' & cp$TM < 0.7,]$group <- 'not_confident_metastatic'
-
-# get cancer state scores
-dt <- scores[match(rownames(cp), bcr_patient_barcode),]
-M <- as.matrix(dt[,-1])
-rownames(M) <- dt$bcr_patient_barcode
-
-dt <- get_factor_specific_variables(M, as.factor(cp$group))
-
-B <- get_basis_matrix(M, cp$group)
-#B2 <- apply(B, 2, function(x) x/max(x)+max(x)/4)
-#pheatmap(B2)
-pheatmap(B[,unique(dt[padj < 0.05]$variable)])
-
-# what are the genes that distinguish confident metastatic from 
-M <- LFs$maui_factors$`TCGA-SKCM.cnv_gex_meth_mut`
-plot_label_specific_vars(M, cp$group, top_vars_per_label = 5)
-dt <- get_factor_specific_variables(M, cp$group)
-top <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)
-
-```
-
-# Show subtypes depend on the objective
-
-Patients clustering depend on the objective of stratification
-
-```{r}
-clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-
-project <- 'pan-gastrointestinal'
-datatype <- 'cnv_gex_meth_mut'
-experiment <- paste0(project, '.', datatype)
-
-M <- LFs$maui_factors[[experiment]]
-M <- t(subset_features_by_variance(t(remove_redundant_variables(M)), topN = 20))
-
-clin <- process_clinical_covariates(rownames(M))
-clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
-
-pfi_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'PFI', duration_col = 'PFI.time')
-pfi_clust <- get_kmeans_subtypes(M[,unique(pfi_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
-pfi_k <- names(which.min(sapply(colnames(pfi_clust), function(x) plot_survival(surv, pfi_clust[,x,drop = F], 'PFI.time', 'PFI', T))))
-
-os_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'OS', duration_col = 'OS.time')
-os_clust <- get_kmeans_subtypes(M[,unique(os_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
-os_k <- names(which.min(sapply(colnames(os_clust), function(x) plot_survival(surv, os_clust[,x,drop = F], 
-                                                                             'OS.time', 'OS', T))))
-
-# combine with known pan-gi subtype annotation
-
-df <- merge(pfi_clust[,pfi_k,drop = F], os_clust[,os_k,drop = F], by = 'row.names')
-df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected
-
-tsne <- plot_tsne(remove_redundant_variables(M), returnData = T)
-
-df$tSNE1 <- tsne[df$Row.names,]$tSNE1
-df$tSNE2 <- tsne[df$Row.names,]$tSNE2
-
-mdf <- melt(df, id.vars = c('Row.names', 'tSNE1', 'tSNE2'))
-
-ggplot(mdf, aes(x = tSNE1, y = tSNE2)) + 
-  geom_point(aes(color = as.factor(value))) + 
-  facet_wrap(~ variable, nrow = 1)
-```
-
-# predict mutational signatures
-```{r}
-# import mutational signature data
-mutsig <- data.table::fread('/data/local/buyar/pancancer_diagnostics/data/mutational_signatures/signature_profile_sample.txt')
-mutsig$Tumor_Sample_Barcode <- gsub("\\.", "-", mutsig$Tumor_Sample_Barcode)
-# only keep TCGA samples
-mutsig <- mutsig[grep('TCGA', Tumor_Sample_Barcode)]
-dt <- data.table::dcast(mutsig, Tumor_Sample_Barcode ~ Signature, value.var = 'Contribution')
-
-cl <- parallel::makeCluster(30) 
-parallel::clusterExport(cl, varlist = c('dt', 'LFs'))
-res <- do.call(rbind, pbapply::pbsapply(simplify = F, cl = cl, unique(mutsig$Signature), function(x) {
-  require(data.table)
-  source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
-  r <- do.call(rbind, lapply(names(LFs), function(m) {
-    message(m)
-    M <- LFs[[m]]$pancancer.cnv_gex_meth_mut
-    s <- predict_numerical_glmnet(M, dt[match(rownames(M), Tumor_Sample_Barcode)][[x]])
-    s$tool <- m
-    return(s)
-  }))
-  r$sig <- x
-  return(r)
-}))
-parallel::stopCluster(cl)
-
-ggplot(res, aes(x = reorder(sig, Rsquare), y = Rsquare)) + 
-  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') + 
-  coord_flip() + labs(x = '')
-
-df <- melt(dcast(res, sig ~ tool, value.var = 'Rsquare'), 
-                id.vars = c('sig', 'maui_factors'))
-
-ggplot(df, aes(x = maui_factors, y = value)) + geom_point() + 
-  facet_wrap(~ variable, nrow = 1) + 
-  geom_abline(slope = 1)
-```
-
-# predictability of clinical factors
-
-```{r, eval = FALSE}
-# predict tumor purity using latent factors 
-res <- do.call(rbind, lapply(names(LFs), function(tool) {
-  message(tool)
-  M <- LFs[[tool]][['pancancer.cnv_gex_meth_mut']]
-  y <- process_clinical_covariates(rownames(M))[,'tumor_purity_IHC']
-  res <- predict_numerical_glmnet(M, y)
-  res$method <- tool
-  return(res)
-}))
-
-# predict tumor stage for pan-gi 
-pangi_stats <- predict_labels(LFs, surv, tcga_subtypes, 
-                              'pan-gastrointestinal', 'cnv_gex_meth_mut', label.type = 'clinical')
-ggplot(pangi_stats[label.type == 'ajcc_pathologic_tumor_stage'][label != 'Undefined'], aes(x = label, y = test_auc)) + 
-  geom_bar(aes(fill = method), stat = 'identity', position = 'dodge') +
-  coord_flip()
-
-# tumor stage 
-label <- 'age_at_initial_pathologic_diagnosis'
-experiment <- 'TCGA-BRCA.cnv_gex_meth_mut'
-stats <- do.call(rbind, lapply(names(LFs), function(tool) {
-  message(tool)
-  M <- LFs[[tool]][[experiment]]
-  y <- process_clinical_covariates(rownames(M))[[label]]
-  M <- M[!is.na(y),]
-  y <- y[!is.na(y)]
-  #res <- predict_cluster_labels(M, y)
-  res <- predict_numerical_glmnet(M, y)
-  res$method <- tool
-  return(res)
-}))
-# with glmnet multi-label pred
-ggplot(stats, aes(x = method, y = Accuracy)) + 
-  geom_bar(stat = 'identity', aes(fill = method)) + 
-  geom_errorbar(aes(ymin = AccuracyLower, ymax = AccuracyUpper), width = 0.2)
-
-```
-
-# predictability of cancer states
-
-```{r, eval = F}
-# read cancer sea gene set scores
-#dataDir <- '/data/local/buyar/pancancer_diagnostics/data/geneset_scores/cancersea'
-#files <- dir(dataDir, '.scores.csv', full.names = T)
-#scores <- do.call(rbind, sapply(simplify = F, files[1:3], function(x) {
-#  data.table::fread(x)
-#}))
-#colnames(scores)[1] <- 'bcr_patient_barcode'
-
-scores <- readRDS('/data/local/buyar/arcas/subtyping_paper/data/mogsa_scores/MOGSA_scores.RDS')
-scores <- do.call(rbind, lapply(scores$cnv_gex_meth_mut, function(x) data.table(x, keep.rownames = T)))
-colnames(scores)[1] <- 'bcr_patient_barcode'
-
-# elastic net regression on pathway scores usign LFs as input
-predict_scores <- function(scores, LFs, label, experiment)
-  do.call(rbind, lapply(names(LFs), function(tool) {
-    message(tool)
-    M <- LFs[[tool]][[experiment]]
-    y <- scores[match(rownames(M), bcr_patient_barcode)][[label]]
-    M <- M[!is.na(y),]
-    y <- y[!is.na(y)]
-    res <- predict_numerical_glmnet(M, y)
-    res$method <- tool
-  return(res)
-}))
-
-labels <- c('ANGIOGENESIS', 'GLYCOLYSIS', 'COAGULATION', 'DNA_REPAIR', 
-            'HYPOXIA', 'ESTROGEN_RESPONSE_EARLY', 'ESTROGEN_RESPONSE_LATE' ,
-            'EPITHELIAL_MESENCHYMAL_TRANSITION')
-brca_scores <- do.call(rbind, lapply(labels, function(l) {
-  message(l)
-  df <- predict_scores(scores, LFs, l, 'TCGA-BRCA.cnv_gex_meth_mut')
-  df$label <- l
-  return(df)
-}))
-
-ggplot(brca_scores, aes(x = label, y = Rsquare)) + 
-  geom_bar(stat = 'identity', aes(fill = method), position = 'dodge') + 
-  coord_flip()
-```
-
-# pan-cancer survival factors 
-```{r, eval = FALSE}
-# factors to use in the model
-clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
-                  'PFI', 'PFI.time')
-experiment <- 'pancancer.cnv_gex_meth_mut'
-res <- do.call(rbind, lapply(names(LFs), function(tool) {
-  message(tool)
-  M <- LFs[[tool]][[experiment]]
-  clin <- process_clinical_covariates(rownames(M))
-  clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
-  cl <- parallel::makeCluster(50)
-  parallel::clusterExport(cl, varlist = c('clin', 'M'))
-  res <- do.call(rbind, pbapply::pblapply(cl = cl, 1:100, function(i) {
-    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
-    require(data.table)
-    s <- sample(rownames(M), 0.5 * nrow(M))
-    dt <- get_prognostic_covariates(clin[s,], M[s,])
-    dt$i <- i
-    return(dt)
-  }))
-  parallel::stopCluster(cl)
-  res$method <- tool
-  return(res)
-}))
-
-# get top factor per method
-plots <- sapply(simplify = F, names(LFs), function(tool) {
-  M <- LFs[[tool]][[experiment]]
-  top_fac <- res[method == tool][,length(unique(i)),by = 'variable'][order(V1, decreasing = T)]$variable[1:3]
-  cowplot::plot_grid(plotlist = lapply(top_fac, function(x) {
-     p <- plot_survival(surv, data.frame('cl' = ifelse(M[,x] > median(M[,x]), 
-                                             'high', 'low')))
-     p$plot
-  }), nrow = 1)
-})
-dev.off()
-cowplot::plot_grid(plotlist = plots, nrow = 3, labels = names(plots))
-
-M <- LFs[[tool]][[experiment]]
-plot_tsne(M, M[,'LF32'])
-```
-
-<!-- Cluster samples into k = number of cancer types in the project -->
-
-<!-- ```{r} -->
-<!-- cl <- parallel::makeCluster(20) -->
-<!-- parallel::clusterExport(cl, varlist = c('LFs', 'projectDict')) -->
-<!-- clusters <- sapply(simplify = F, names(LFs), function(tool) { -->
-<!--   message("Clustering top ",tool) -->
-<!--   pbapply::pbsapply(cl = cl, simplify = F, names(LFs[[tool]]), function(experiment) { -->
-<!--     message(experiment) -->
-<!--     project <- strsplit(experiment, '\\.')[[1]][1] -->
-<!--     datatype <- strsplit(experiment, '\\.')[[1]][2] -->
-<!--     source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R') -->
-<!--     require(data.table) -->
-<!--     M <- LFs[[tool]][[experiment]] -->
-<!--     if(tool == 'maui_factors') { -->
-<!--       M <- t(subset_features_by_variance(t(M), topN = 10)) -->
-<!--     } else { -->
-<!--       M <- M[,head(colnames(M), 10),drop = F] -->
-<!--     } -->
-<!--     val <- c('internal')  -->
-<!--     #if(ncol(M) > 2) {val <- c('internal', 'stability')} -->
-<!--     #k <- length(projectDict[[project]]) -->
-<!--     res <- clValid::clValid(obj = M, -->
-<!--                           nClust = 3:10, -->
-<!--                           clMethods = 'kmeans', -->
-<!--                           validation = val, -->
-<!--                           maxitems = 10000, -->
-<!--                           nstart = 1000) -->
-<!--     memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']]))) -->
-<!--     colnames(memberships) <- paste0('k', colnames(memberships)) -->
-<!--     stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')] -->
-<!--     colnames(stats) <- c('metric', 'k', 'clustering_method', 'score') -->
-<!--     return(list('clustering_stats' = stats, 'cluster_memberships' = memberships)) -->
-<!--   }) -->
-<!-- }) -->
-<!-- parallel::stopCluster(cl) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- stats <- do.call(rbind, lapply(names(clusters), function(tool) { -->
-<!--   do.call(rbind, lapply(names(clusters[[tool]]), function(experiment) { -->
-<!--       cancer_type <- strsplit(experiment, "\\.")[[1]][1] -->
-<!--       datatypes <- strsplit(experiment, "\\.")[[1]][2] -->
-<!--       stats <- clusters[[tool]][[experiment]][['clustering_stats']] -->
-<!--       stats$cancer_type <- cancer_type -->
-<!--       stats$datatypes <- datatypes -->
-<!--       stats$method <- tool -->
-<!--       return(stats) -->
-<!--   })) -->
-<!-- })) -->
-<!-- ``` -->
-
-<!-- clustering metrics -->
-<!-- ```{r} -->
-<!-- ggboxplot(stats[datatypes == 'cnv_gex_mut'], x = 'method', y = 'score',  -->
-<!--           add = 'jitter', facet.by = 'metric', color = 'method') -->
-<!-- ``` -->
-
-<!-- Find out how much the cluster labels correspond to cancer types  -->
-<!-- ```{r} -->
-<!-- cl <- parallel::makeCluster(20) -->
-<!-- parallel::clusterExport(cl, varlist = c('clusters', 'projectDict', 'surv', )) -->
-<!-- # compute mutual information stats (between cluster labels and cancer types) -->
-<!-- MI_stats <- do.call(rbind, sapply(simplify = F, names(clusters), function(tool) { -->
-<!--   message("computing MI for ",tool) -->
-<!--   do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, names(clusters[[tool]]), -->
-<!--                                    function(experiment) { -->
-<!--     message(experiment) -->
-<!--     project <- strsplit(experiment, '\\.')[[1]][1] -->
-<!--     datatype <- strsplit(experiment, '\\.')[[1]][2] -->
-<!--     source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R') -->
-<!--     require(data.table) -->
-<!--     k <- stats[method == tool][cancer_type == project][datatypes == datatype][,.SD[which.max(score)]]$k -->
-<!--     clust <- clusters[[tool]][[experiment]][['cluster_memberships']][,paste0('k',k),drop=F] -->
-<!--     labels <- surv[match(rownames(clust), bcr_patient_barcode)]$project -->
-
-<!--     data.table('method' = tool, 'project' = project, 'datatypes' = datatype,  -->
-<!--                'cramersV' = compute_cramers_V(cbind(clust, labels))[['cramersV']],  -->
-<!--                'ARI' = aricode::ARI(clust[,1], labels),  -->
-<!--                'AMI' = aricode::AMI(clust[,1], labels)) -->
-<!--                                    })) -->
-<!--   })) -->
-<!-- parallel::stopCluster(cl) -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- ggboxplot(MI_stats[datatypes == 'cnv_gex_mut'],  -->
-<!--           x = 'method', y = 'AMI',  -->
-<!--           add = 'jitter') +  -->
-<!--     stat_compare_means(method = 'wilcox.test', paired = T, method.args = list('alternative' = 'greater'),  -->
-<!--                      comparisons = list(c('maui_factors', 'mofa_factors'),  -->
-<!--                                         c('maui_factors', 'pca_factors'))) + -->
-<!--   labs(title = 'One-vs-all test-AUC by cancer type') -->
-
-<!-- ``` -->
-
-<!-- Unsupervised clustering experiment -->
-<!-- ```{r} -->
-<!-- get_plot <- function(M, plot.type = 'cancertype') { -->
-
-<!--   df <- plot_tsne(M) -->
-<!--   if(plot.type == 'cancertype') { -->
-<!--     df$tcga <- surv[match(rownames(df), bcr_patient_barcode)]$project -->
-<!--   } else if (plot.type == 'subtype') { -->
-<!--     df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected -->
-<!--   } -->
-<!--   clust <- get_kmeans_subtypes(M[rownames(df),], k_vals = length(unique(df$tcga)), 0) -->
-<!--   df$maui <- as.factor(clust[rownames(df),]) -->
-
-<!--   p <- ggplot(df, aes(x = tSNE1, y = tSNE2))   -->
-<!--   p1 <- p + geom_point(aes(color = maui)) + -->
-<!--     geom_text_repel(data = get_centroid(df[,1:2], df$maui),  -->
-<!--               aes(x,y,label = text.label), size = 6) -->
-<!--   p2 <- p + geom_point(aes(color = tcga)) +  -->
-<!--           geom_text_repel(data = get_centroid(df[,1:2], df$tcga),  -->
-<!--               aes(x,y,label = text.label), size = 5) -->
-
-<!--   p <- cowplot::plot_grid(p1, p2) -->
-<!--   # also compute mutual information -->
-<!--   ami <- aricode::AMI(df$tcga, df$maui) -->
-<!--   return(list('p' = p, 'ami' = ami)) -->
-<!-- } -->
-
-<!-- pr <- 'pancancer' #'pancancer' # 'pan-gastrointestinal' -->
-<!-- e <- paste0(pr, '.cnv_gex_meth_mut') -->
-<!-- M <- LFs$maui_factors[[e]] -->
-<!-- ct <- surv[match(rownames(M), bcr_patient_barcode)]$project -->
-<!-- st <-  tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected -->
-
-<!-- dt.ct <- get_factor_specific_variables(M, ct) -->
-<!-- ct_vars <- unique(get_top_by_label(dt.ct[padj < 0.05][order(padj)], label.type = 'ref_cl', 2)[['variable']]) -->
-<!-- dt.st <- get_factor_specific_variables(M, st) -->
-<!-- st_vars <- unique(get_top_by_label(dt.st[padj < 0.05][order(padj)], 'ref_cl', 2)[['variable']]) -->
-
-<!-- # LFs with most variance -->
-<!-- top_vars <- rownames(subset_features_by_variance(t(M), topN = 42)) -->
-<!-- # using top vars will cause the clusters dominated by the major component to dominate -->
-
-<!-- p1 <- get_plot(M, 'cancertype') # use all factors  (unsupervised) -->
-<!-- p2 <- get_plot(M[,top_vars], 'cancertype') # use top most variable factors (unsupervised) -->
-<!-- p3 <- get_plot(M[,ct_vars], 'cancertype') # use cancer type specific factor (supervised ) -->
-<!-- p4 <- get_plot(remove_redundant_variables(M, cutoff = 0.6), 'cancertype') # (unsupervised) -->
-<!-- p5 <- get_plot(LFs$pca_factors[[e]]) # pca -->
-<!-- p5.1 <- get_plot(LFs$pca_factors[[e]][,1:10]) # pca -->
-<!-- p5.2 <- get_plot(LFs$pca_factors[[e]][,1:20]) # pca -->
-<!-- p5.4 <- get_plot(LFs$pca_factors[[e]][,1:40]) -->
-
-<!-- p6 <- get_plot(LFs$mofa_factors[[e]]) # mofa -->
-
-<!-- ps <- lapply(c(5:8/10), function(r) { -->
-<!--   get_plot(remove_redundant_variables(M, cutoff = r), 'cancertype') -->
-<!-- }) -->
-
-<!-- dt <- do.call(rbind, lapply(experiments, function(e) { -->
-<!--   x <- LFs$maui_factors[[e]] -->
-<!--   y <- remove_redundant_variables(x, cutoff = 0.6) -->
-<!--   data.table('before' = ncol(x), 'after' = ncol(y), 'n' = nrow(x), 'exp' = e,  -->
-<!--              'redundant' = ncol(x) - ncol(y)) -->
-<!-- })) -->
-
-<!-- p5 <- ggscatter(dt, x = 'n', y = 'redundant', add = 'reg.line') + -->
-<!--    stat_cor(label.x = 400, label.y = 80, size = 6) + -->
-<!--   labs(x = 'Number of samples', y = 'Percentage of redundant LFs') + -->
-<!--   stat_regline_equation(label.x = 400, label.y = 75, size = 6) +  -->
-<!--   geom_text_repel(aes(label = gsub('.cnv_gex_meth_mut', '', exp))) -->
-
-<!-- ``` -->

### 2021-03-10 src/manuscript_figures.Rmd
collect figures into a list to print them later

@@
 knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
+# todo: pass this as argument
 settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
@@
 
+# todo: move this to settings file
 inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
@@
 tcga_subtypes <- process_tcga_subtypes()
+
+Figures <- c() # accummulate all figures in this
+SuppFigures <- c() # accumulate all supplementary figures in this
+
 ```
@@
 
-Fig1 <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B')), 
+Fig <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B')), 
                   plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('C', 'D')), 
@@
 
-print(Fig1)
-ggsave('pancancer_cancer_type_prediction.pdf', plot = Fig1, units = 'in', width = 12, height = 9)
+Figures <- c(Figures, Fig)
+ggsave('pancancer_cancer_type_prediction.pdf', plot = Fig, units = 'in', width = 12, height = 9)
 
@@
 
-Fig2 <- plot_grid(plot_grid(p1, p2, nrow = 2), p3, ncol = 2, rel_widths = c(2, 3), 
+Fig <- plot_grid(plot_grid(p1, p2, nrow = 2), p3, ncol = 2, rel_widths = c(2, 3), 
           labels = 'AUTO')
-print(Fig2)
+print(Fig)
 
-ggsave('pancancer_subtype_prediction.pdf', plot = Fig2, units = 'in', width = 12, height = 9)
+ggsave('pancancer_subtype_prediction.pdf', plot = Fig, units = 'in', width = 12, height = 9)
 
-# train/test accuracy correlation 
-#ggscatter(pancancer_subtype_stats, x = 'train_auc', y = 'test_auc', 
- #         color = 'method', facet.by = 'method', add = 'reg.line')
+SuppFigures <- c(SuppFigures, Fig)
 ```
@@
 
-#M <- LFs$pca_factors[[e]]
-#ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
-
 # for maui, we drop redundant LFs 
@@
 Fig <- cowplot::plot_grid(p1$p, p2$p, p3$p, p4, nrow = 4, labels = 'AUTO')
-print(Fig)
 ggsave("pancancer.unsupervised_clustering.pdf", Fig, units = 'in', width = 10, height = 12)
+
+Figures <- c(Figures, Fig)
 ```
@@
 cl <- parallel::makeCluster(20)
-parallel::clusterExport(cl, varlist = c('LFs', 'experiments'))
+parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
 clusters <- sapply(simplify = F, names(LFs), function(tool) {
@@
     message(experiment)
-    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
+    source(settings$utility_script)
     require(data.table)
@@
 
-Fig4 <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
+Fig <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
                   p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))
-print(Fig4)
-ggsave('cluster_stability_metrics.pdf', plot = Fig4, units = 'in', 
+ggsave('cluster_stability_metrics.pdf', plot = Fig, units = 'in', 
        width = 10, height = 8)
 
+Figures <- c(Figures, Fig)
 ```
@@
 cl <- parallel::makeCluster(20)
-parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
+parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 
+                                        'surv', 'experiments', 'settings'))
 surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
     require(data.table)
-    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
+    source(settings$utility_script)
     message(experiment)
@@
     clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
-    res <- get_prognostic_covariates(clin, M)
+    res <- get_prognostic_covariates(surv, clin, M)
     if(!is.null(res)) {
@@
                    label_size = 8)
-print(p1)
-ggsave(filename = 'surv.adjustedcurves.prog_lfs.pdf', plot = p1, width = 12, height = 8, units = 'in')
-
 # plot hazard ratios for top LF per cancer type 
@@
   geom_errorbar(aes(ymin = log(hr_low), ymax = log(hr_high)), width = 0.2) + 
-  #geom_text(aes(label = latent_factor, y = ifelse(hr > 1, log(hr+0.005), log(hr-0.005))), 
-   #         nudge_x = 0.25) + 
   labs(x = '', y = 'log(hazard-ratio) of top latent factor per cancer\n(adjusted for tumor stage + age + gender)')+
@@
 
-p <- cowplot::plot_grid(p3, p2, nrow = 1, labels = 'AUTO')
+Fig <- cowplot::plot_grid(cowplot::plot_grid(p3, p2, nrow = 1, labels = c('A', 'B')),
+                        p1, labels = c('', 'C'), nrow = 2)
 
-print(p)
-ggsave(filename = 'surv.Cindex_improv.hazard_ratio.pdf', plot = p, width = 12, height = 8, units = 'in')
+ggsave(filename = 'surv.Cindex_improv.hazard_ratio.pdf', plot = p, width = 12, height = 14, units = 'in')
 
+Figures <- c(Figures, Fig)
 ```
@@
 cl <- parallel::makeCluster(10)
-parallel::clusterExport(cl, varlist = c('experiments', 'LFs'))
+parallel::clusterExport(cl, varlist = c('experiments', 'LFs', 'settings'))
 clin.stats <- do.call(rbind, 
@@
                require(data.table)
-               source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
+               source(settings$utility_script)
                M <- LFs$maui_factors[[experiment]]
@@
                         factors <- factors[!is.na(factors)]
-                        dt <- get_factor_specific_variables(m, factors) #compute_anova(m, factors)
+                        dt <- get_factor_specific_variables(m, factors) 
                         dt$label <- x
@@
 # top 5 variables per clinical factor 
-
 top <- clin.stats[padj < 0.05,.SD[which.min(padj)], by = label][order(padj),.SD[which.min(padj)],by = project]
@@
   factors <- factors[!is.na(factors)]
-  plot_label_specific_vars(M, factors, string_width = 24)+ #, top_vars = top_vars) + 
+  plot_label_specific_vars(M, factors, string_width = 24)+ 
     labs(title = pr, subtitle = l) + 
@@
 p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
-p <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
-print(p)
+Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
 ggsave(filename = 'clinical_factors.LFs.pdf', p, units = 'in', width = 12, height = 9)
+Figures <- c(Figures, Fig)
 ```
@@
 # read MSI data
-dataDir <- '/data/local/buyar/pancancer_diagnostics/data/GDCdata_prepared'
+dataDir <- settings$GDCdataDir
 MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
@@
 msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))
-# with glmnet and error bars 
-# for supplements (includes other tools)
-# p1 <- ggplot(msi_stats, aes(x = method, y = auc)) + 
-#   geom_bar(aes(fill = method), stat = 'identity') + 
-#   geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2) + 
-#   labs(y = 'AUC',x = '') + theme(legend.title = element_blank()) + 
-#   facet_grid(~ label)
 
@@
   scale_color_gradient(low = 'gray', high = 'red')
-p2.3 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(cancertype))) + 
-  theme(legend.title = element_blank())
-p2.4 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(subtype))) + 
-  theme(legend.title = element_blank())
 
-print(cowplot::plot_grid(p2.1, p2.2, p2.3, p2.4, ncol = 2, labels = 'AUTO'))
 # top features that contribute to top factors relevant to MSI status prediction
@@
 
-p <- cowplot::plot_grid(p1, p3, p2.1, p2.2, labels = 'AUTO')
-print(p)
-ggsave(filename = 'MSI.status.pdf', p, units = 'in', width = 10, height = 8)
+Fig <- cowplot::plot_grid(p1, p3, p2.1, p2.2, labels = 'AUTO')
+ggsave(filename = 'MSI.status.pdf', Fig, units = 'in', width = 10, height = 8)
 
+Figures <- c(Figures, Fig)
 ```
@@
 require(cluster)
-tcga_subtypes <- process_tcga_subtypes()
-# subtype specific LFs
-get_top_by_label <- function(dt, label.type, topN = 5) {
-  do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
-}
-
-pr <- 'pan-gastrointestinal' #'TCGA-BRCA' # 'glioma' 
+pr <- 'pan-gastrointestinal' 
 experiment <- paste0(pr, '.cnv_gex_meth_mut')
@@
 
-# use gap statistic to get clusters
 vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 3)[['variable']])
-# use gap statistic to find optimum clusters 
-#r1 <- cluster::clusGap(M[,vars], FUNcluster = kmeans, K.max = 20, iter.max = 50, B = 100)
-#plot(r1)
-#k <- cluster::maxSE(r1$Tab[,3], r1$Tab[,4], method = 'firstSEmax', SE.factor = 1)
-k <- length(na.omit(unique(st)))
+
+k <- length(na.omit(unique(st))) # set k to number of known subtypes
 # cluster samples using only subtype specific factors 
@@
 # use hallmark gene sets to annotate cluster specific functions
+# TODO: update this to point to /data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore
+# and remove paths from script (put it in settings.yaml)
 scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))
@@
 # altogether
-P <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
+Fig <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
                          labels = 'AUTO', ncol = 2)
-print(P)
 ggsave(filename = paste0('semi_supervised.clustering.', experiment,'.pdf'), 
-       P, units = 'in', 
+       Fig, units = 'in', 
        width = 10, height = 12)
 
+Figures <- c(Figures, Fig)
 ```
@@
 # factors to use in the model
-clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
-                  'PFI', 'PFI.time')
+clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
 
@@
     require(survminer)
-    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
+    source(settings$utility_script)
     message(experiment)
@@
     clin <- process_clinical_covariates(rownames(M))
-    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
-    res <- get_prognostic_covariates(clin, M)
+    res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
     # now cluster samples based on only survival predictive factors
@@
 
-# use hallmark gene sets to annotate cluster specific functions
-#scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))
-
 # use canonical pathways to annotate prognostic groups
+# TODO: update this to point to /data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore
+# and don't use hard-coded paths 
 scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_canonical_pathways/ssgsea.canonical_pathways.RDS'))
 
-# scores geneset scores
-# M survival predictive LF matrix
+# scores: geneset scores
+# M: survival predictive LF matrix
 # maui_clust: best survival prognostic clustering of M 
@@
   colnames(df)[3] <- 'maui'
-  #df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
 
@@
 
-Pc <- cowplot::plot_grid(p1, plot_grid(P$p1, P$p2, nrow = 1, labels = c('B', 'C')), 
+Fig <- cowplot::plot_grid(p1, plot_grid(P$p1, P$p2, nrow = 1, labels = c('B', 'C')), 
                    P$h1$gtable, nrow = 3, labels = c('A', '', 'D'))
 
-print(Pc)
 ggsave(filename = paste0('semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
-       Pc, units = 'in', 
+       Fig, units = 'in', 
        width = 10, height = 12)
 
+Figures <- c(Figures, Fig)
+
 ```

### 2021-03-11 src/common_functions.R
remove patients with sd NA

@@ -850,8 +850,14 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     
     # remove normal samples
     cnv <- subsetAssayData(cnv, patientIDs, replaceIDs = FALSE)
-    # remove patients that are all zeros
-    cnv <- cnv[,names(which(apply(cnv, 2, function(x) sum(x != 0)) > 0))]
+    if(cnv_flavor == 'cnv') {
+      # remove patients that are all zeros
+      cnv <- cnv[,names(which(apply(cnv, 2, function(x) sum(x != 0)) > 0))]
+    } else if(cnv_flavor == 'cnv_gistic') {
+      # remove patients for whom when data is scaled sd equals to NA
+      keep <- names(which(!is.na(apply(scale(cnv), 2, sd))))
+      cnv <- cnv[,keep]
+    }
     
     # get batch info
     batches <- getBatchTable(colnames(cnv))

### 2021-03-11 src/common_functions.R
add new functions

@@ -1215,3 +1215,24 @@ remove_redundant_variables <- function(M, perc = 99, cutoff = NULL) { #, cutoff 
   return(M)
 }
 
+compute_AMI <- function(M, k, target_labels) {
+  df <-  data.frame(get_kmeans_subtypes(M, k_vals = k, 0))
+  colnames(df) <- 'cluster'
+  ami <- aricode::AMI(df$cluster, target_labels)
+  return(ami)
+}
+
+# geneSetScoresFolder: path to folder with gene set scores
+# annotation: subfolder to import
+import_geneset_scores <- function(geneSetScoresFolder, annotation = 'msigdb_hallmarks') {
+  # use hallmark gene sets to annotate cluster specific functions
+  files <- dir(file.path(geneSetScoresFolder, annotation), 
+               '.scores.csv', full.names = T)
+  scores <- do.call(rbind, sapply(simplify = F, files, function(x) {
+    dt <- data.table::fread(x)
+    M <- as.matrix(dt[,-1])
+    rownames(M) <- dt$V1
+    return(M)
+  }))
+  return(scores)
+}

### 2021-03-11 src/manuscript_figures.Rmd
update figures; reorganize, keep figure counters, change figure filenames

@@
 inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
+geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
 assay_dir <- settings$assay_output$folder
@@
 
-Figures <- c() # accummulate all figures in this
-SuppFigures <- c() # accumulate all supplementary figures in this
+Figures <- list() # accummulate all figures in this
+SuppFigures <- list() # accumulate all supplementary figures in this
 
+TOOLS <- c('pca_factors', 'maui_factors', 'mofa_factors', 'mcia_factors')
+
+fig_count <- 0
+supFig_count <- 0
 ```
@@
 ```{r import_lfs}
-LFs <- sapply(simplify = F, c('pca_factors', 'maui_factors', 'mofa_factors'), function(folder) {
+LFs <- sapply(simplify = F, TOOLS, function(folder) {
   message('Importing ',folder)
@@
 ```{r import_feature_weights}
-FWs <- sapply(simplify = F, c('pca_factors', 'maui_factors', 'mofa_factors'), function(folder) {
+FWs <- sapply(simplify = F, TOOLS, function(folder) {
   message('Importing ',folder)
@@
                            project, datatype, label.type, ...) {
-  source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
+  source(settings$utility_script)
   predict_stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
@@
 plot_comp1 <- function(dt) {
+  dt$method <- toupper(gsub("_factors", "", dt$method))
+  comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
   ggboxplot(dt, 
@@
           add = 'jitter', legend = 'none') + 
-    stat_compare_means(method = 'wilcox.test', paired = T, method.args = list('alternative' = 'greater'), 
-                     comparisons = list(c('maui_factors', 'mofa_factors'), 
-                                        c('maui_factors', 'pca_factors'))) +
+    stat_compare_means(method = 'wilcox.test', paired = T, 
+                       method.args = list('alternative' = 'greater'), 
+                     comparisons = comps) +
     labs(y = 'AUC') + theme(axis.text = element_text(size = 10))
@@
 plot_comp2 <- function(dt) {
+  dt$method <- toupper(gsub("_factors", "", dt$method))
   dtc <- melt(dcast(dt, label ~ method, value.var = 'test_auc'), 
-              id.vars = c('label', 'maui_factors'))
-  dtc$difference <- dtc$maui_factors - dtc$value
+              id.vars = c('label', 'MAUI'))
+  dtc$difference <- dtc$MAUI - dtc$value
   ggplot(dtc, 
-         aes(x = maui_factors, y = value)) + 
+         aes(x = MAUI, y = value)) + 
     geom_point() + geom_abline(slope = 1) + 
@@
                                   'cnv_gex_meth_mut', label.type = 'cancer_type', num.trees = 1000)
-p1 <- plot_comp1(pancancer_stats) + theme(axis.text.x = element_blank())
+p1 <- plot_comp1(pancancer_stats) + theme(axis.title.x = element_blank())
 p2 <- plot_comp2(pancancer_stats)
@@
 
-Fig <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B')), 
-                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('C', 'D')), 
+# do unsupervised clustering and compute AMI 
+experiment <- 'pancancer.cnv_gex_meth_mut'
+df <- do.call(rbind, pbapply::pblapply(TOOLS, function(x) {
+  message(x)
+  M <- LFs[[x]][[experiment]]
+  if(x == 'maui_factors') {
+    M <- remove_redundant_variables(M, perc = 99)
+  }
+  # find cancer types per sample
+  ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
+  data.frame('tool' = x, 
+             'AMI' = compute_AMI(M, k = length(unique(ct)), target_labels = ct))
+}))
+df$tool <- toupper(gsub("_factors", "", df$tool))
+p5 <- ggplot(df, aes(x = tool, y = AMI)) +
+  geom_bar(aes(fill = tool), stat = 'identity') + 
+  geom_text(aes(label = round(AMI, 2))) + 
+  theme(legend.title = element_blank(), axis.title.x = element_blank())
+
+Fig <- plot_grid(plot_grid(p1, p2, p5, nrow = 1, labels = c('A', 'B', 'C')), 
+                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('D', 'E')), 
                   nrow = 2,
@@
 
-Figures <- c(Figures, Fig)
-ggsave('pancancer_cancer_type_prediction.pdf', plot = Fig, units = 'in', width = 12, height = 9)
+fig_count <- fig_count + 1
+Figures[[fig_count]] <- Fig
+ggsave(paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf'), 
+              plot = Fig, units = 'in', width = 12, height = 9)
 
@@
 
-p1 <- plot_comp1(pancancer_subtype_stats) 
+p1 <- plot_comp1(pancancer_subtype_stats) + theme(axis.title.x = element_blank())
 p2 <- plot_comp2(pancancer_subtype_stats)
@@
 
-Fig <- plot_grid(plot_grid(p1, p2, nrow = 2), p3, ncol = 2, rel_widths = c(2, 3), 
-          labels = 'AUTO')
-print(Fig)
+# data type contribution to top LF per factor
+p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
+                                           'pancancer', 'cnv_gex_meth_mut', factors = st, 
+                                           top_lfs = 1, top_features = 100) 
+p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
+  theme(legend.position = 'right', legend.background = element_blank())
 
-ggsave('pancancer_subtype_prediction.pdf', plot = Fig, units = 'in', width = 12, height = 9)
+Fig <- plot_grid(plot_grid(plot_grid(p1, p2 + theme(axis.text.x = element_text(angle = 45)), 
+                                     nrow = 2, labels = c('A', 'B')), 
+                           p3 + theme(legend.position = 'none', 
+                                      axis.text.x = element_text(angle = 45)), 
+                           labels = c('', 'C'), ncol = 2, rel_widths = c(1,3)), 
+                 p4 + theme(axis.text.x = element_text(angle = 45)), 
+                 labels = c('', '', 'D'), ncol = 2, rel_widths = c(2.5,1))
 
-SuppFigures <- c(SuppFigures, Fig)
-```
-
-# Unsupervised pan-cancer clustering
-
-We do unsupervised clustering of samples and measure AMI to observe which tool's clusters correspond best to 
-ground truth (cancer types)
-
-```{r unsupervised_clustering, fig.height=12, fig.width=10}
-get_plot <- function(M, plot.type = 'cancertype') {
-
-  df <- plot_tsne(M)
-  if(plot.type == 'cancertype') {
-    df$tcga <- surv[match(rownames(df), bcr_patient_barcode)]$project
-    df <- df[!is.na(df$tcga),]
-  } else if (plot.type == 'subtype') {
-    df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
-    df <- df[!is.na(df$tcga),]
-  }
-  clust <- get_kmeans_subtypes(M[rownames(df),], k_vals = length(unique(df$tcga)), 0)
-  df$cluster <- as.factor(clust[rownames(df),])
-
-  p <- ggplot(df, aes(x = tSNE1, y = tSNE2))  
-  p1 <- p + geom_point(aes(color = cluster)) +
-    geom_text_repel(data = get_centroid(df[,1:2], df$cluster), 
-              aes(x,y,label = text.label), size = 6) + 
-    theme(legend.position = 'none')
-  p2 <- p + geom_point(aes(color = tcga)) + 
-          geom_text_repel(data = get_centroid(df[,1:2], df$tcga), 
-              aes(x,y,label = gsub('TCGA-', '', text.label)), size = 3) +
-    theme(legend.position = 'none')
-
-  p <- cowplot::plot_grid(p1, p2)
-  # also compute mutual information
-  ami <- aricode::AMI(df$tcga, df$cluster)
-  return(list('p' = p, 'ami' = ami))
-}
-
-pr <-  'pancancer' #'pancancer 99%' # 'nsclc 98%' pan-gastrointestinal 98% 'pan-gyn 98%'
-e <- paste0(pr, '.cnv_gex_meth_mut')
-
-# for maui, we drop redundant LFs 
-p1 <- get_plot(remove_redundant_variables(LFs$maui_factors[[e]], perc = 99), 'cancertype') 
-p2 <- get_plot(LFs$pca_factors[[e]], 'cancertype') # use all pcs, 
-p3 <- get_plot(LFs$mofa_factors[[e]], 'cancertype') # mofa
-
-# use AMI values to make another plot
-p4 <- ggplot(melt(data.frame('MAUI' = p1$ami, 'PCA' = p2$ami, 'MOFA' = p3$ami)), 
-       aes(x = variable, y = value)) + 
-  geom_bar(aes(fill = variable), stat = 'identity') + 
-  geom_text(aes(y = value + 0.05, label = round(value, 2))) + 
-  labs(y = 'AMI', x = '') + 
-  theme(legend.title = element_blank()) + coord_equal()
-
-Fig <- cowplot::plot_grid(p1$p, p2$p, p3$p, p4, nrow = 4, labels = 'AUTO')
-ggsave("pancancer.unsupervised_clustering.pdf", Fig, units = 'in', width = 10, height = 12)
-
-Figures <- c(Figures, Fig)
+supFig_count <- supFig_count + 1
+ggsave(paste0('SuppFigure', supFig_count, '.pancancer_subtype_prediction.pdf'), 
+       plot = Fig, units = 'in', width = 18, height = 12)
+SuppFigures[[supFig_count]] <- Fig
 ```
@@
     if(tool == 'maui_factors') {
-      Mr <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
-      M <- t(subset_features_by_variance(t(Mr), topN = 10))
-     } else {
-      M <- M[,head(colnames(M), 10),drop = F]
+      M <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
     } 
+    M <- t(subset_features_by_variance(t(M), topN = 10))
+
     val <- c('internal') 
@@
 dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb]
+dt1$method <- toupper(gsub("_factors", "", dt1$method))
+comps <- lapply(setdiff(unique(dt1$method), 'MAUI'), function(x) c('MAUI', x))
+
 p1 <- ggboxplot(dt1, 
@@
   stat_compare_means(method = 'wilcox.test', paired = TRUE, 
-                     comparisons = list(c('maui_factors', 'pca_factors'), 
-                                        c('maui_factors', 'mofa_factors'))) + 
+                     comparisons = comps) + 
   labs(title = 'Silhouette score', x = '') + 
@@
 dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb]
+dt2$method <- toupper(gsub("_factors", "", dt2$method))
+comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
 p2 <- ggboxplot(dt2, 
@@
   stat_compare_means(method = 'wilcox.test', paired = TRUE, 
-                     comparisons = list(c('maui_factors', 'pca_factors'), 
-                                        c('maui_factors', 'mofa_factors'))) + 
+                     comparisons = comps) + 
   labs(title = 'Cluster Stability Score (APN)', x = '') + 
@@
 dt$kv <- paste0('kmeans(k=',dt$k,')')
+dt$method <- toupper(gsub("_factors", "", dt$method))
 p3 <- ggplot(dt,
@@
                   p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))
-ggsave('cluster_stability_metrics.pdf', plot = Fig, units = 'in', 
+
+fig_count <- fig_count + 1
+Figures[[fig_count]] <- Fig
+
+ggsave(paste0('Figure', fig_count, '.cluster_stability_metrics.pdf'), 
+       plot = Fig, units = 'in', 
        width = 10, height = 8)
-
-Figures <- c(Figures, Fig)
 ```
@@
     clin <- process_clinical_covariates(rownames(M))
-    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
-    res <- get_prognostic_covariates(surv, clin, M)
+    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
+    res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
     if(!is.null(res)) {
@@
       dat$Row.names <- NULL
+      dat <- get_surv_df(surv, dat) # add PFI values
       # compute hazard ratios 
@@
       # compute Cindex with all survival-predictive LFs and without LFs 
-      vars <- head(res[order(pval)]$variable, 10)
-      rsf_stats <- fit_RSF_clin(clin, M[,vars], ntree = 1000)
+      vars <- res[pval < 0.05]$variable
+      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin), M[,vars], ntree = 1000)
 
@@
 
-ggsave(filename = 'surv.Cindex_improv.hazard_ratio.pdf', plot = p, width = 12, height = 14, units = 'in')
+fig_count <- fig_count + 1
+ggsave(filename = paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf'), 
+                         plot = Fig, width = 12, height = 14, units = 'in')
 
-Figures <- c(Figures, Fig)
+Figures[[fig_count]] <- Fig
 ```
@@
 Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
-ggsave(filename = 'clinical_factors.LFs.pdf', p, units = 'in', width = 12, height = 9)
-Figures <- c(Figures, Fig)
+fig_count <- fig_count + 1
+ggsave(filename = paste0('Figure', fig_count, '.clinical_factors.LFs.pdf'), 
+       plot = Fig, units = 'in', width = 12, height = 9)
+Figures[[fig_count]] <- Fig
 ```
@@
 p3 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
-  geom_bar(aes(fill = Var1), stat='identity') + theme(legend.title = element_blank()) + 
+  geom_bar(aes(fill = Var1), stat='identity') + 
+  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45)) + 
   labs(x = paste0("top features contributing to ",top_lf))
 
-Fig <- cowplot::plot_grid(p1, p3, p2.1, p2.2, labels = 'AUTO')
-ggsave(filename = 'MSI.status.pdf', Fig, units = 'in', width = 10, height = 8)
+Fig <- cowplot::plot_grid(p1, 
+                          plot_grid(p2.1, p2.2, p3, nrow = 1, labels = c('B', 'C', 'D')),
+                          nrow = 2, labels = c('A', ''), rel_heights = c(5,3))
+fig_count <- fig_count + 1
+ggsave(filename = paste0("Figure", fig_count, '.MSI.status.pdf'), 
+                         plot = Fig, units = 'in', width = 10, height = 8)
 
-Figures <- c(Figures, Fig)
+Figures[[fig_count]] <- Fig
 ```
@@
 # use hallmark gene sets to annotate cluster specific functions
-# TODO: update this to point to /data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore
-# and remove paths from script (put it in settings.yaml)
-scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))
+scores <- import_geneset_scores(geneSetScoresFolder, 'msigdb_hallmarks')
+
 S <- scores[intersect(rownames(M), rownames(scores)),]
+colnames(S) <- gsub("HALLMARK_", "", colnames(S))
 clust$tcga <- tcga_subtypes[match(rownames(clust), bcr_patient_barcode)]$Subtype_Selected
@@
                          labels = 'AUTO', ncol = 2)
-ggsave(filename = paste0('semi_supervised.clustering.', experiment,'.pdf'), 
+
+fig_count <- fig_count + 1
+ggsave(filename = paste0('Figure', fig_count, '.semi_supervised.clustering.', experiment,'.pdf'), 
        Fig, units = 'in', 
@@
 
-Figures <- c(Figures, Fig)
+Figures[[fig_count]] <- Fig
 ```
@@
 cl <- parallel::makeCluster(20)
-parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
+parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 
+                                        'experiments', 'settings'))
 prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
@@
     clin <- process_clinical_covariates(rownames(M))
+    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
     res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
@@
     }
-    clust <- get_spectral_clusters(M[,prog_lfs], k_vals = 2:10)
+    clust <-  get_kmeans_subtypes(M[,prog_lfs], k_vals = 3:8, min_cluster_size = 0)
     # find value of k which minimizes log-rank p-value
@@
   if(is.null(prog.clust[[experiment]])) {return(NULL)}
-  df <- prog.clust[[experiment]][['prog.clust']]
+  df <- data.frame(prog.clust[[experiment]][['prog.clust']])
   colnames(df)[1] <- 'maui'
@@
 # use canonical pathways to annotate prognostic groups
-# TODO: update this to point to /data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore
-# and don't use hard-coded paths 
-scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_canonical_pathways/ssgsea.canonical_pathways.RDS'))
+scores <- import_geneset_scores(geneSetScoresFolder, 'msigdb_canonical_pathways')
 
@@
 # plot examples for which we improve upon tcga
-dt <- dcast(stats, experiment ~ source, value.var = 'pval')
+dt <- data.table(dcast(stats, experiment ~ source, value.var = 'pval'))
 dt$improvement <- -log10(dt$maui) - -log10(dt$tcga)
@@
 
-pr <- 'TCGA-UCEC'
-e <- paste0(pr, '.cnv_gex_meth_mut')
+e <- dt[1,]$experiment
 M <- prog.clust[[e]]$prog.lf
@@
 
-ggsave(filename = paste0('semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
-       Fig, units = 'in', 
+fig_count <- fig_count + 1
+ggsave(filename = paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
+       plot = Fig, units = 'in', 
        width = 10, height = 12)
 
-Figures <- c(Figures, Fig)
-
+Figures[[fig_count]] <- Fig
 ```

### 2021-03-11 src/manuscript_figures.Rmd
only print figures to a word document - provide urls to high-res images

@@
 ---
-title: "Pan-cancer analysis"
+title: "Manuscript Figures"
 output:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
+knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
 # todo: pass this as argument
@@
 geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
+FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas'
 assay_dir <- settings$assay_output$folder
@@
 Figures[[fig_count]] <- Fig
-ggsave(paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf'), 
+fig.name <- paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
               plot = Fig, units = 'in', width = 12, height = 9)
@@
 
+```{r clustering_metrics, fig.width=10, fig.height=8}
 # clustering metrics of various cancers 
+# Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs
 
-Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
-```{r clustering_metrics, fig.width=10, fig.height=8}
 multi_omics_comb <- 'cnv_gex_meth_mut'
@@
 Figures[[fig_count]] <- Fig
-
-ggsave(paste0('Figure', fig_count, '.cluster_stability_metrics.pdf'), 
+fig.name <- paste0('Figure', fig_count, '.cluster_stability_metrics.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name),        
        plot = Fig, units = 'in', 
@@
 
+```{r survival_factors, fig.width=12, fig.height=8}
 # survival-predictive factors
+#I look for survival-predictive maui LFs, show how much c-index improves on top of 
+#clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.
 
-I look for survival-predictive maui LFs, show how much c-index improves on top of 
-clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.
-
-```{r survival_factors, fig.width=12, fig.height=8}
 # factors to use in the model
@@
 fig_count <- fig_count + 1
-ggsave(filename = paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf'), 
+Figures[[fig_count]] <- Fig
+fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(filename = file.path(FigureFolder, fig.name), 
        plot = Fig, width = 12, height = 14, units = 'in')
-
-Figures[[fig_count]] <- Fig
 ```
 
+```{r clinical_factors, fig.width=12, fig.height=9}
 # clinical factor associated MAUI LFs
+# Find maui LFs associated to clinical factors and plot them
 
-Find maui LFs associated to clinical factors and plot them
-
-```{r clinical_factors, fig.width=12, fig.height=9}
 experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
@@
 p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
+
 Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
+
 fig_count <- fig_count + 1
-ggsave(filename = paste0('Figure', fig_count, '.clinical_factors.LFs.pdf'), 
+Figures[[fig_count]] <- Fig
+fig.name <- paste0('Figure', fig_count, '.clinical_factors.LFs.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(filename = file.path(FigureFolder, fig.name), 
        plot = Fig, units = 'in', width = 12, height = 9)
-Figures[[fig_count]] <- Fig
 ```
 
+```{r msi_stats, fig.width=10, fig.height=8}
 # Predicting MSI status using latent factors
-```{r msi_stats, fig.width=10, fig.height=8}
 # predict MSI status for pan-gastrointestinal tumors
@@
 fig_count <- fig_count + 1
-ggsave(filename = paste0("Figure", fig_count, '.MSI.status.pdf'), 
+Figures[[fig_count]] <- Fig
+fig.name <- paste0('Figure', fig_count, '.MSI.status.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(filename = file.path(FigureFolder, fig.name), 
        plot = Fig, units = 'in', width = 10, height = 8)
-
-Figures[[fig_count]] <- Fig
 ```
 
+```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
 # Semi-supervised clustering of samples 
-
-1. Cluster samples by LFs associated to known TCGA subtypes or those that are associated to survival outcomes. 
-2. Annotate each cluster using cancer state scores or hallmark gene set scores
+# 1. Cluster samples by LFs associated to known TCGA subtypes or those that are associated to survival outcomes. 
+#2. Annotate each cluster using cancer state scores or hallmark gene set scores
 
 ## TCGA subtype informed clusters 
-
-### Use-case: pan-gastrointestinal cancers
-```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
+## Use-case: pan-gastrointestinal cancers
 require(survminer)
@@
 fig_count <- fig_count + 1
-ggsave(filename = paste0('Figure', fig_count, '.semi_supervised.clustering.', experiment,'.pdf'), 
+Figures[[fig_count]] <- Fig
+fig.name <- paste0('Figure', fig_count, '.semi_supervised.clustering.', experiment,'.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(filename = file.path(FigureFolder, fig.name), 
        Fig, units = 'in', 
        width = 10, height = 12)
-
-Figures[[fig_count]] <- Fig
 ```
 
+```{r semi_supervised_survival, fig.width=10, fig.height=12}
+
 ## Clustering informed by survival outcomes
+## Pick survival-predictive LFs and cluster samples based on those
 
-Pick survival-predictive LFs and cluster samples based on those
-```{r semi_supervised_survival, fig.width=10, fig.height=12}
 # factors to use in the model
@@
 fig_count <- fig_count + 1
-ggsave(filename = paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
+Figures[[fig_count]] <- Fig
+fig.name <- paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.', experiment,'.pdf')
+names(Figures[fig.count]) <- fig.name
+ggsave(filename = file.path(FigureFolder, fig.name), 
        plot = Fig, units = 'in', 
@@
 
-Figures[[fig_count]] <- Fig
 ```
+
+```{r print_figures, results='asis', include=TRUE, fig.width=10, fig.height=11}
+base_url <- 'https://bimsbstatic.mdc-berlin.de'
+for(i in 1:length(Figures)) {
+  fig.url <- gsub("/data/bimsbstatic/public", base_url, file.path(FigureFolder, names(Figures)[i]))
+  cat("# Figure",i)
+  print(Figures[[i]])
+  cat("High-res image:",fig.url,"\n\n")
+}
+```

### 2021-03-11 src/manuscript_figures.Rmd
print legends to word document

@@
 
-# todo: move this to settings file
+# todo: move paths to settings file
 inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
@@
 SuppFigures <- list() # accumulate all supplementary figures in this
-
+Legends <- list() # figure legend text
 TOOLS <- c('pca_factors', 'maui_factors', 'mofa_factors', 'mcia_factors')
@@
               plot = Fig, units = 'in', width = 12, height = 9)
+Legends[[fig_count]] <- "Predictive power of latent factors in distinguishing cancer types.
+A-B) comparing maui with PCA/MOFA/MCIA.
+AUC: Area under the ROC curve computed on the test dataset (30% of held-out data) trained on 70% of the dataset using a Random Forest model.
+Each point represents one cancer type (pairwise one-sided Wilcoxon test on AUCs, alternative = maui AUC is not greater than others) C) Adjusted Mutual Information (AMI) value computed by comparing unsupervised clustering labels (k-means algorithm, k set to 21) with the cancer types.
+D) Top maui LFs detected per cancer type.
+Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples.
+E) Percentage of top feature types contributing to top predictive maui latent factor per cancer type."
 
@@
        width = 10, height = 8)
+Legends[[fig_count]] <- "Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in A) and cluster stability (Average Proportion of Non-Overlap (APN) in B).
+Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) based on the top 10 LFs/PCs derived from each tool.
+C) Same as in panels A and B), but split by the value of k used in clustering."
 ```
@@
        plot = Fig, width = 12, height = 14, units = 'in')
+Legends[[fig_count]] <- "Using MAUI LFs for survival prediction.
+A) Harrell’s C-index computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs.
+B) Hazard ratios of top LFs picked for each cancer adjusted for clinical factors (tumor stage + age + gender).
+Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
 ```
@@
        plot = Fig, units = 'in', width = 12, height = 9)
+Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type.
+B) Example case: top LFs associated with each histological subtype of TCGA-SARC.
+C) Example case: top LFs associated with each histological grade of TCGA-HNSC."
 ```
@@
        plot = Fig, units = 'in', width = 10, height = 8)
+Legends[[fig_count]] <- "Predicting MSI status of pan-gastrointestinal tumors using MAUI latent factors.
+A) AUC values for predicting MSI status using MAUI LFs B) Top features that contribute to MAUI latent factor (‘LF95’) that is the most predictive latent factor in distinguishing samples with high microsatellite instability from the rest.
+C) tSNE plots of samples generated using top MSI-predictive LFs (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI latent factor (‘LF95’) which is most predictive of MSI-H samples."
 ```
@@
        width = 10, height = 12)
+Legends[[fig_count]] <- "Clustering of pan-gastrointestinal tumors guided by TCGA subtypes.
+Subtype specific MAUI LFs are used to cluster the samples, which yields better survival curves as shown in C,D).
+Clusters are annotated using pathway activity scores (top cluster associated cancer hallmarks)."
 ```
@@
 fig.name <- paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.', experiment,'.pdf')
-names(Figures[fig.count]) <- fig.name
+names(Figures)[fig_count] <- fig.name
 ggsave(filename = file.path(FigureFolder, fig.name), 
@@
        width = 10, height = 12)
-
+Legends[[fig_count]] <- "Log-rank test of prognostic subtypes by MAUI and existing TCGA subtypes.
+B) tSNE plot of MAUI LFs that are predictive of survival outcomes for TCGA-UCEC.
+C) Kaplan-Meier curve of the best/worst surviving prognostic MAUI clusters D) Biological pathway scores that distinguish best/worst surviving clusters."
 ```
@@
   fig.url <- gsub("/data/bimsbstatic/public", base_url, file.path(FigureFolder, names(Figures)[i]))
-  cat("# Figure",i)
+  cat("# Figure",i,"[high-res image](",fig.url,")\n\n")
   print(Figures[[i]])
-  cat("High-res image:",fig.url,"\n\n")
+  cat("\n Figure",i,":",Legends[[i]],"\n\n")
 }
 ```
+
+`r date()`

### 2021-03-12 src/common_functions.R
allow clin to be null

@@ -1034,16 +1034,19 @@ process_clinical_covariates <- function(patientIDs) {
 
 # build a cox model for each variable along correcting for existing clinical variables
 # return list of covariates that are significant 
-get_prognostic_covariates <- function(surv, clin, explored_covariates, 
+get_prognostic_covariates <- function(surv, clin = NULL, explored_covariates, 
                                       event_col = 'PFI', duration_col = 'PFI.time') {
   require(survival)
   res <- do.call(rbind, pbapply::pblapply(colnames(explored_covariates), function(v) {
     #message(v,"\n")
     x <- explored_covariates[,v,drop = F]
-    dat <- merge(clin, x, by = 'row.names')
-    rownames(dat) <- dat$Row.names
-    dat$Row.names <- NULL
-    
+    if(!is.null(clin)) {
+      dat <- merge(clin, x, by = 'row.names')
+      rownames(dat) <- dat$Row.names
+      dat$Row.names <- NULL
+    } else {
+      dat <- data.frame(x)
+    }
     dat <- cbind(dat, data.frame('status' = surv[match(rownames(dat), bcr_patient_barcode)][,get(event_col)], 
                                'time' = surv[match(rownames(dat), bcr_patient_barcode)][,get(duration_col)]))
     dat <- dat[!is.na(dat$time),]
@@ -1053,12 +1056,7 @@ get_prognostic_covariates <- function(surv, clin, explored_covariates,
     s <- summary(cox1)
     pval <- as.numeric(s$coefficients[v,][5])
     cindex <- s$concordance[['C']]
-    # build a model without the covariate and compute cindex 
-    cox2 <- coxph(Surv(time, status) ~ ., 
-                  data = subset(dat, select = c(colnames(clin), 'status', 'time')))
-    s2 <- summary(cox2)
-    cindex2 <- s2$concordance[['C']]
-    return(data.table('variable' = v, 'pval' = pval, 'cindex' = cindex, 'cindex_change' = cindex - cindex2))
+    return(data.table('variable' = v, 'pval' = pval, 'cindex' = cindex))
   }))
   if(!is.null(res)) {
     res <- res[pval < 0.05]

### 2021-03-15 src/common_functions.R
make filtering faster - add messages

@@ -754,6 +754,7 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     # reorder features and concatenate columns
     mut <- do.call(cbind, lapply(mut, function(x) x[common_features,]))
     if(!is.null(GOI)) {mut <- mut[intersect(rownames(mut), GOI),]}
+    
     # select genes with most non-zero samples
     selected <- head(names(sort(apply(mut, 1, function(x) sum(x > 0)), decreasing = T)), topN)
     
@@ -793,6 +794,7 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     
     # subset by features
     if (!is.null(GOI)) {gex <- gex[intersect(rownames(gex), GOI),]}
+    
     gex <- subset_features_by_variance(gex, topN = topN)
     rownames(gex) <- paste0('gex.', rownames(gex))
     
@@ -1200,14 +1202,27 @@ plot_label_specific_vars <- function(M, labels, top_vars_per_label = 1,
 }
 
 # given a matrix, remove variables that are redundant (Based on correlation cut-off)
-remove_redundant_variables <- function(M, perc = 99, cutoff = NULL) { #, cutoff = 0.7) {
+# samples on the rows, variables on the columns
+# downsize_byTopVar : integer value. If set, the top most variable features are first selected.
+remove_redundant_variables <- function(M, perc = 99, cutoff = NULL, downsize_byTopVar = NULL) { #, cutoff = 0.7) {
+  if(!is.null(downsize_byTopVar)) {
+    message(date()," => selecting top ",downsize_byTopVar, " features by variance first")
+    M <- t(subset_features_by_variance(t(M), downsize_byTopVar))
+  }
+  message(date()," => computing pairwise correlations")
+  x <- Rfast::cora(M)
   if(is.null(cutoff)) {
+    message(date(), " => looking for correlation cutoff...")
     # define cutoff based on correlation distribution, remove those above 99% 
-    x <- cor(M)
-    cutoff <- quantile(x[upper.tri(x)], 1:100/100)[[perc]]
+    # sample features for efficiency
+    s <- sample(1:nrow(x), 1000)
+    xs <- x[s,s]
+    cutoff <- quantile(xs[upper.tri(xs)], 1:100/100)[[perc]]
+    message(date(), " => setting correlation cut-off to ",cutoff, " at ",perc,"th percentile")
   }
   require(caret)
[…]

### 2021-03-16 src/common_functions.R
remove redundant feature from gex/mut and meth

@@ -755,6 +755,9 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     mut <- do.call(cbind, lapply(mut, function(x) x[common_features,]))
     if(!is.null(GOI)) {mut <- mut[intersect(rownames(mut), GOI),]}
     
+    # remove redundant features
+    mut <- t(remove_redundant_variables(t(mut), perc = 99, downsize_byTopVar = 15000))
+    
     # select genes with most non-zero samples
     selected <- head(names(sort(apply(mut, 1, function(x) sum(x > 0)), decreasing = T)), topN)
     
@@ -794,6 +797,10 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     
     # subset by features
     if (!is.null(GOI)) {gex <- gex[intersect(rownames(gex), GOI),]}
+    
+    # remove redundant features
+    gex <- t(remove_redundant_variables(t(gex), perc = 99, downsize_byTopVar = 15000))
+    
     
     gex <- subset_features_by_variance(gex, topN = topN)
     rownames(gex) <- paste0('gex.', rownames(gex))
@@ -841,9 +848,11 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     
     selected <- NULL
     if(cnv_flavor == 'cnv') {
+      # TODO: consider removing highly redundant ones
       # get top features # pick features which are most often gained/lost
       selected <- head(names(sort(apply(cnv, 1, function(x) sum(x == 0)))), topN) 
     } else if(cnv_flavor == 'cnv_gistic') {
+      # TODO: consider removing highly redundant ones
       # pick top features by variance in gistic scores
       selected <- head(names(sort(apply(cnv, 1, var), decreasing = T)), topN) 
     }
@@ -887,6 +896,9 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     })
     common_features <- Reduce(intersect, lapply(meth, rownames))
     meth <- do.call(cbind, lapply(meth, function(x) x[common_features,]))
+    
+    # remove redundant features
+    meth <- t(remove_redundant_variables(t(meth), perc = 99, downsize_byTopVar = 15000))
     
     # get top features by variance
     meth <- subset_features_by_variance(meth, topN = topN)
@@ -1215,8 +1227,11 @@ remove_redundant_variables <- function(M, perc = 99, cutoff = NULL, downsize_byT
[…]

### 2021-03-16 src/snakefile.py
get locations from settings file

@@ -1,10 +1,11 @@
 
 import os
-ASSAYDIR = '/data/local/buyar/arcas/pancancer_multiomics_manuscript/assays' 
-SRCDIR = '/data/local/buyar/arcas/pancancer_multiomics_manuscript/src' 
+
+ASSAYDIR = config['assay_output']['folder'] 
+SRCDIR = config['utility_script'] 
 ASSAYS = set([os.path.splitext(f)[0] for f in os.listdir(ASSAYDIR) if re.match(r'.*\.csv$', f)])
 RSCRIPT = '/opt/R-4.0.2/lib64/R/bin/Rscript'
-OUTDIR = 'output'
+OUTDIR = config['pipeline_output']
 LOG_DIR = os.path.join(OUTDIR, "logs")
 
 rule all:

### 2021-03-17 src/setup_experiments.R
print logs to file

@@ -33,7 +33,7 @@ if(!dir.exists(assayDir)) {
   dir.create(assayDir)
 }
 
-cl <- parallel::makeCluster(length(projects))
+cl <- parallel::makeCluster(length(projects), outfile = 'setup_experiments.log')
 parallel::clusterExport(cl = cl, varlist = c('settings', 'projectDict', 'dataDir', 
                                              'surv', 'assayDir', 'GOI'))
 pbapply::pblapply(cl = cl, projects, function(pr) {

### 2021-03-17 src/common_functions.R
turn off redundancy filtering - doesnt make much difference

@@ -756,7 +756,7 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     if(!is.null(GOI)) {mut <- mut[intersect(rownames(mut), GOI),]}
     
     # remove redundant features
-    mut <- t(remove_redundant_variables(t(mut), perc = 99, downsize_byTopVar = 15000))
+    #mut <- t(remove_redundant_variables(t(mut), perc = 99, downsize_byTopVar = 15000))
     
     # select genes with most non-zero samples
     selected <- head(names(sort(apply(mut, 1, function(x) sum(x > 0)), decreasing = T)), topN)
@@ -799,7 +799,7 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     if (!is.null(GOI)) {gex <- gex[intersect(rownames(gex), GOI),]}
     
     # remove redundant features
-    gex <- t(remove_redundant_variables(t(gex), perc = 99, downsize_byTopVar = 15000))
+    #gex <- t(remove_redundant_variables(t(gex), perc = 99, downsize_byTopVar = 15000))
     
     
     gex <- subset_features_by_variance(gex, topN = topN)
@@ -898,7 +898,7 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     meth <- do.call(cbind, lapply(meth, function(x) x[common_features,]))
     
     # remove redundant features
-    meth <- t(remove_redundant_variables(t(meth), perc = 99, downsize_byTopVar = 15000))
+    #meth <- t(remove_redundant_variables(t(meth), perc = 99, downsize_byTopVar = 15000))
     
     # get top features by variance
     meth <- subset_features_by_variance(meth, topN = topN)

### 2021-03-17 src/snakefile.py
fix bug - path to source code

@@ -2,7 +2,7 @@
 import os
 
 ASSAYDIR = config['assay_output']['folder'] 
-SRCDIR = config['utility_script'] 
+SRCDIR = '/data/local/buyar/arcas/pancancer_multiomics_manuscript/src' 
 ASSAYS = set([os.path.splitext(f)[0] for f in os.listdir(ASSAYDIR) if re.match(r'.*\.csv$', f)])
 RSCRIPT = '/opt/R-4.0.2/lib64/R/bin/Rscript'
 OUTDIR = config['pipeline_output']

### 2021-03-17 src/manuscript_figures.Rmd
add survival benchmark, start luad_vs_lusc analysis

@@
 theme_set(ggpubr::theme_pubclean())
-
-# todo: move paths to settings file
-inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
 geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
 FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas'
+
+inputDir <- settings$pipeline_output
 assay_dir <- settings$assay_output$folder
@@
 
+TOOLS <- c('pca_factors', 'maui_factors', 'mofa_factors', 'mcia_factors')
+
 Figures <- list() # accummulate all figures in this
@@
 Legends <- list() # figure legend text
-TOOLS <- c('pca_factors', 'maui_factors', 'mofa_factors', 'mcia_factors')
 
@@
 
-Fig <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
-                  p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))
+# survival prediction accuracy benchmarking
+# find survival-predictive LFs by univariate cox models and use them in RSF
+cl <- parallel::makeCluster(20)
+parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv'))
+surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
+  message("survival benchmark for ",tool)
+  do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
+    message(experiment)
+    source(settings$utility_script)
+    require(data.table)
+    require(survminer)
+    M <- LFs[[tool]][[experiment]]
+    if(tool == 'maui_factors') {
+      M <- remove_redundant_variables(M, 99)
+    }
+    dt <- data.table('cindex.pfi' = NA, 'cindex.os' = NA, 'experiment' = experiment, 'method' = tool)
+    # compute pfi stats
+    vars <- get_prognostic_covariates(surv, clin = NULL, explored_covariates = M,
+                                      event_col = 'PFI', duration_col = 'PFI.time')
+    if(!is.null(vars)) {
+      v <- head(vars[order(pval)][pval < 0.05][['variable']], 10)
+      fit <- fit_RSF(surv, M[,v], 1000, 'PFI.time', 'PFI')
+      dt$cindex.pfi <- round(1 - mean(fit$err.rate, na.rm = T), 2)
+    }
+    vars <- get_prognostic_covariates(surv, clin = NULL, explored_covariates = M,
+                                      event_col = 'OS', duration_col = 'OS.time')
+    if(!is.null(vars)) {
+      v <- head(vars[order(pval)][pval < 0.05][['variable']], 10)
+      fit <- fit_RSF(surv, M[,v], 1000, 'OS.time', 'OS')
+      dt$cindex.os <- round(1 - mean(fit$err.rate, na.rm = T), 2)
+    }
+    return(dt)
+  }))
+}))
+parallel::stopCluster(cl)
+
+dt <- melt(surv.stats, measure.vars = c('cindex.pfi', 'cindex.os'))
+dt[is.na(value)]$value <- min(dt$value, na.rm = T) # 
+dt$method <- toupper(gsub("_factors", "", dt$method))
+dt$project <- unlist(sapply(strsplit(dt$experiment, "\\."), function(x) x[1]))
+dt$endpoint <- ifelse(dt$variable == 'cindex.pfi', 'Progression Free Survival', 
+                      'Overall Survival')
+# PCA/MOFA win, compare against PCA
+comps <- lapply(setdiff(unique(dt$method), 'PCA'), function(x) c('PCA', x))
+p4 <- ggboxplot(dt, x = 'method', y = 'value', color = 'method',
+                add = 'jitter', facet.by = 'endpoint') +
+  stat_compare_means(method = 'wilcox.test', paired = TRUE,
+                     comparisons = comps) +
+  labs(y = "Harrel's C-index", x = '') +
+  theme(legend.position = 'none')
+
+Fig <- plot_grid(plot_grid(p1, p2, labels = c('A', 'B'), ncol = 1),
+                 plot_grid(p3, p4, labels = c('C', 'D'), ncol = 1), 
+                 ncol = 2, rel_widths = c(2,3))
 
@@
        plot = Fig, units = 'in', 
-       width = 10, height = 8)
+       width = 10, height = 10)
 Legends[[fig_count]] <- "Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in A) and cluster stability (Average Proportion of Non-Overlap (APN) in B).
 Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) based on the top 10 LFs/PCs derived from each tool.
-C) Same as in panels A and B), but split by the value of k used in clustering."
+C) Same as in panels A and B), but split by the value of k used in clustering.
+D) Survival outcome prediction accuracy for 21 cancer types by each tool's top 10 survival-related factors based on Overall Survival or Progression Free Survival endpoints."
 ```
@@
 
+```{r}
+# Analysis of non-small-cell lung cancer types (LUSC vs LUAD)
+# reference study: https://www.nature.com/articles/s41598-020-77284-8
+M <- remove_redundant_variables(LFs$maui_factors$nsclc.cnv_gex_meth_mut, perc = 99)
+ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
+
+# find type-specific LFs
+dt <- get_factor_specific_variables(M, ct)
+top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 10)
+s <- unique(top_vars$variable)
+pheatmap(M[,s], show_rownames = F)
+plot_tsne(M[,s], ct)
+plot_umap(M, ct)
+plot_tsne(M, ct)
+clust <- get_spectral_clusters(M, k_vals = 2:6)
+clin <- process_clinical_covariates(rownames(M))
+prog_vars <- get_prognostic_covariates(surv, clin, M)
+prog.clust <- get_kmeans_subtypes(M[,unique(prog_vars$variable)], k_vals = 2:8, min_cluster_size = 0)
+
+which.min(sapply(colnames(prog.clust), function(x) {
+  plot_survival(surv, prog.clust[,x,drop=F], return_pval = T)
+}))
+
+plot_survival(surv, prog.clust[,'kmeans_3', drop = F])
+plot_survival(surv, data.frame('ct' = ct, row.names = rownames(M)), 
+              'PFI.time', 'PFI')
+
+df <- plot_tsne(M[,unique(prog_vars$variable)], ct, returnData = T)
+df$pfi_subtype <- as.factor(prog.clust[rownames(df),'kmeans_3'])
+
+ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
+  geom_point(aes(color = as.factor(pfi_subtype))) +
+  facet_wrap(~ group, nrow = 1 )
+
+ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
+  geom_point(aes(color = group)) +
+  facet_wrap(~ as.factor(pfi_subtype), nrow = 1 )
+
+scores <- sapply(simplify = F, c('msigdb_hallmarks', 'msigdb_canonical_pathways', 'cancersea'), 
+       function(x) {
+         scores <- import_geneset_scores(geneSetScoresFolder, x)
+         S <- scores[intersect(rownames(M), rownames(scores)),]
+         return(S)
+       })
+
+pathways.prog <- get_factor_specific_variables(scores$cancersea, df$pfi_subtype)[padj < 0.05][order(padj)]
+pathways.ct <- get_factor_specific_variables(scores$cancersea, df$group)[padj < 0.05][order(padj)]
+
+get_top_by_label(pathways.ct[padj < 0.05][order(padj)], 'ref_cl', 10)
+
+```
+
 ```{r semi_supervised_subtypes, fig.width=10, fig.height=12}

### 2021-03-17 src/common_functions.R
update glmnet with return roc option, add functions to read msigdb/cancersea

@@ -236,13 +236,14 @@ predict_cluster_labels <- function(M, labels, partition = 0.7, ...) {
   return(auc.dt)
 }
 
-predict_cluster_labels_glmnet <- function(M, labels) {
+# returnROC: if set to TRUE, will return a ROC object, when FALSE returns AUC stats 
+predict_cluster_labels_glmnet <- function(M, labels, returnROC = FALSE) {
   require(caret)
   require(ranger)
   df <- data.frame(M)
   labels <- as.factor(as.character(labels))
   # for each factor level, do a one-vs-all classification with probabilities
-  auc.dt <- do.call(rbind, lapply(levels(labels), function(x) {
+  res <- sapply(simplify = F, levels(labels), function(x) {
     message(x)
     df$y <- as.factor(ifelse(labels == x, 'one', 'rest'))
     set.seed(3456)
@@ -268,13 +269,22 @@ predict_cluster_labels_glmnet <- function(M, labels) {
       test.pred <- predict(model, x.test, type = 'prob')
       r <- pROC::roc(testing$y, test.pred[,'one'])
       test.auc <- as.numeric(pROC::ci.auc(pROC::roc(testing$y, test.pred[,'one'])))
+      if(returnROC == TRUE) {
+        return(r)
+      }
       return(data.table('auc.lower' = test.auc[1], 'auc' = test.auc[2], 
                         'auc.upper' = test.auc[3], 'label' = x))
     } else {
+      if(returnROC == TRUE) {
+        return(NULL)
+      }
       return(data.table('auc.lower' = NA, 'auc' = NA, 'auc.upper' = NA, 'label' = x))
     }
-  }))
-  return(auc.dt)
+  })
+  if(returnROC == FALSE) {
+    return(do.call(rbind, res))
+  }
+  return(res)
 }
 # train a regression model to predict a target variable (y)
 # M: samples on the row, features on the column
@@ -1263,4 +1273,27 @@ import_geneset_scores <- function(geneSetScoresFolder, annotation = 'msigdb_hall
     return(M)
   }))
   return(scores)
+}
+
+# f: path to gmt format file 
+read_msigdb <- function(f) {
+  geneSets <- lapply(readLines(f), function(x) {
+    unlist(strsplit(x, "\t"))
+  })
+  #add names
+  names(geneSets) <- unlist(lapply(geneSets, function(x) x[1]))
+  #remove first two 
+  geneSets <- lapply(geneSets, function(x) x[-c(1:2)])
+  return(geneSets)
+}
[…]

### 2021-03-17 src/manuscript_figures.Rmd
add figures/legends for luad vs lusc analysis

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
+knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
 # todo: pass this as argument
@@
 library(cowplot)
+library(ggridges)
 theme_set(ggpubr::theme_pubclean())
@@
 tcga_subtypes <- process_tcga_subtypes()
+
+ens2hgnc <- readRDS(settings$ens2hgnc)
 
@@
 
+res <- predict_cluster_labels_glmnet(M, ct)
+
 # find type-specific LFs
 dt <- get_factor_specific_variables(M, ct)
-top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 10)
-s <- unique(top_vars$variable)
-pheatmap(M[,s], show_rownames = F)
-plot_tsne(M[,s], ct)
-plot_umap(M, ct)
-plot_tsne(M, ct)
-clust <- get_spectral_clusters(M, k_vals = 2:6)
-clin <- process_clinical_covariates(rownames(M))
-prog_vars <- get_prognostic_covariates(surv, clin, M)
-prog.clust <- get_kmeans_subtypes(M[,unique(prog_vars$variable)], k_vals = 2:8, min_cluster_size = 0)
-
-which.min(sapply(colnames(prog.clust), function(x) {
-  plot_survival(surv, prog.clust[,x,drop=F], return_pval = T)
-}))
-
-plot_survival(surv, prog.clust[,'kmeans_3', drop = F])
-plot_survival(surv, data.frame('ct' = ct, row.names = rownames(M)), 
-              'PFI.time', 'PFI')
-
-df <- plot_tsne(M[,unique(prog_vars$variable)], ct, returnData = T)
-df$pfi_subtype <- as.factor(prog.clust[rownames(df),'kmeans_3'])
-
-ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
-  geom_point(aes(color = as.factor(pfi_subtype))) +
-  facet_wrap(~ group, nrow = 1 )
-
-ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
-  geom_point(aes(color = group)) +
-  facet_wrap(~ as.factor(pfi_subtype), nrow = 1 )
-
-scores <- sapply(simplify = F, c('msigdb_hallmarks', 'msigdb_canonical_pathways', 'cancersea'), 
-       function(x) {
-         scores <- import_geneset_scores(geneSetScoresFolder, x)
-         S <- scores[intersect(rownames(M), rownames(scores)),]
-         return(S)
-       })
-
-pathways.prog <- get_factor_specific_variables(scores$cancersea, df$pfi_subtype)[padj < 0.05][order(padj)]
-pathways.ct <- get_factor_specific_variables(scores$cancersea, df$group)[padj < 0.05][order(padj)]
-
-get_top_by_label(pathways.ct[padj < 0.05][order(padj)], 'ref_cl', 10)
-
+top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 1)
+
+# roc curves for predicting lung cancer types
+rocs <- predict_cluster_labels_glmnet(M, ct, returnROC = TRUE)
+aucs <- sapply(rocs, pROC::auc)
+p1 <- pROC::ggroc(rocs, size = 1)
+p1$data$name <- paste0(p1$data$name, " (AUC=", round(aucs[p1$data$name], 3),")")
+p1 <- p1 + geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color="grey", linetype="dashed") +
+  theme(legend.title = element_blank(), legend.direction = 'vertical')
+
+# tsne plot 
+df <- plot_tsne(M, ct, returnData = T)
+
+# top lf for LUSC
+top_lf_lusc <- top_vars[ref_cl == 'TCGA-LUSC']$variable
+top_lf_luad <- top_vars[ref_cl == 'TCGA-LUAD']$variable
+df <- cbind(df, 
+            M[,top_lf_lusc,drop = F],
+            M[,top_lf_luad,drop = F])
+p <- ggplot(df, aes(x = tSNE1, y = tSNE2)) 
+p2 <- p + geom_point(aes(color = group)) + 
+  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
+            aes(x,y,label = text.label), size = 5)
+p3 <- p + geom_point(aes_string(color = top_lf_lusc)) + 
+  scale_color_gradient(low = 'gray', high = 'red') +
+  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
+            aes(x,y,label = text.label), size = 5)
+p4 <- p + geom_point(aes_string(color = top_lf_luad)) + 
+  scale_color_gradient(low = 'gray', high = 'red') +
+  geom_text_repel(data = get_centroid(df[,1:2], df$group), 
+            aes(x,y,label = text.label), size = 5)
+p5 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 'nsclc', 'cnv_gex_meth_mut', 
+                                     factors = ct, top_lfs = 1, top_features = 500) + 
+  labs(y = 'Relative Contribution\nto top latent factor')
+
+dt <- data.table(M[,top_vars[['variable']]], keep.rownames = T)
+dt <- dt[rowSums(dt[,2:3]) > 0] 
+dt$pr <- surv[match(dt$rn, bcr_patient_barcode)]$project
+
+# alternatively, split samples by LF scores, plot gene set scores 
+scores <- import_geneset_scores(geneSetScoresFolder, annotation = 'cancersea')
+S <- data.table(scores[intersect(rownames(scores), rownames(M)),], keep.rownames = T)
+
+dt <- merge(dt, S, by = 'rn')
+
+mdt <- melt(dt, measure.vars = setdiff(colnames(S), 'rn'))
+
+p6 <- ggplot(mdt, aes(x = value, y = variable)) + 
+  ggridges::geom_density_ridges(aes(fill = get(top_lf_lusc) > get(top_lf_luad)), alpha = 0.7) +
+  labs(x = 'CancerSEA gene set scores', y = 'Cancer Cell States') + 
+  guides(fill=guide_legend(title=paste0(top_lf_lusc, " > ",top_lf_luad)))
+
+Fig <-  cowplot::plot_grid(p1, p2, p3, p4, p5, p6, 
+                         labels = 'AUTO', ncol = 2)
+
+fig_count <- fig_count + 1
+Figures[[fig_count]] <- Fig
+fig.name <- paste0('Figure', fig_count, '.luad_vs_lusc.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(filename = file.path(FigureFolder, fig.name), 
+       Fig, units = 'in', 
+       width = 10, height = 12)
+Legends[[fig_count]] <- "Classification of Non-Small-Cell Lung Adenocarcinoma samples using MAUI latent factors.
+A) ROC curves with AUC values on testing data (40% of the dataset) trained on Elastic Net with 5-fold cross-validation.
+B) t-SNE plot of MAUI LFs colored by cancer types.
+C) t-SNE plot of MAUI LFs colored by
+the top most-predictive latent factor for TCGA-LUSC samples. D) t-SNE plot of MAUI LFs colored by the top 
+most predictive latent factor for TCGA-LUAD samples.
+E) Percentage of top 500 contributing features each of the top most predictive latent factors broken down into categories by cancer type and data type.
+F) Cancer Cell State score distribution of samples grouped by whether the sample has a higher score for LUSC-predictive latent factor than the LUAD-predictive latent factor."
 ```
@@
   print(Figures[[i]])
-  cat("\n Figure",i,":",Legends[[i]],"\n\n")
+  cat("\n\n Figure",i,":",Legends[[i]],"\n\n")
 }

### 2021-03-18 src/manuscript_figures.Rmd
add supplementary figure about subtypes dependency on outcome

@@
 library(ggridges)
+library(ggcorrplot)
+library(ComplexHeatmap)
+library(pheatmap)
 theme_set(ggpubr::theme_pubclean())
@@
 supFig_count <- supFig_count + 1
-ggsave(paste0('SuppFigure', supFig_count, '.pancancer_subtype_prediction.pdf'), 
+SuppFigures[[supFig_count]] <- Fig
+fig.name <- paste0('SuppFigure', fig_count, '.pancancer_subtype_prediction.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
               plot = Fig, units = 'in', width = 18, height = 12)
-SuppFigures[[supFig_count]] <- Fig
+SuppLegends[[fig_count]] <- "Predictive power of latent factors in distinguishing known TCGA cancer subtypes A-B) comparing MAUI with PCA/MOFA/MCIA.
+AUC: Area under the ROC curve computed on the test dataset (30% of held-out data) trained on 70% of the dataset using a Random Forest model.
+Each point represents one cancer subtype (pairwise one-sided Wilcoxon test on AUCs, alternative = maui AUC is not greater than others) C) Top maui LFs detected per cancer subtype (only displaying subtypes with at least 20 samples).
+Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding subtype samples.
+D) Percentage of top feature types contributing to top predictive maui latent factor per cancer subtype."
 ```
@@
 experiment <- paste0(pr, '.cnv_gex_meth_mut')
-M <- LFs$maui_factors[[experiment]]
+M <- remove_redundant_variables(LFs$maui_factors[[experiment]], perc = 99)
 st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
@@
 
+```{r}
+# Show subtypes depend on the objective (dependent variable) - they are dynamic, not universal or carved in stone 
+# for a given cancer. There is not just one set of subtypes for a given cancer. Subtypes need to be phrased 
+# along with the dependent variable used to come up with the subtypes. 
+# Patients clustering depend on the objective of stratification
+clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
+project <- 'pan-gastrointestinal'
+datatype <- 'cnv_gex_meth_mut'
+experiment <- paste0(project, '.', datatype)
+
+M <- LFs$maui_factors[[experiment]]
+M <- remove_redundant_variables(M, perc = 99)
+
+clin <- process_clinical_covariates(rownames(M))
+clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
+
+pfi_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'PFI', duration_col = 'PFI.time')
+pfi_clust <- get_kmeans_subtypes(M[,unique(pfi_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
+pfi_k <- names(which.min(sapply(colnames(pfi_clust), function(x) plot_survival(surv, pfi_clust[,x,drop = F], 'PFI.time', 'PFI', T))))
+
+os_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'OS', duration_col = 'OS.time')
+os_clust <- get_kmeans_subtypes(M[,unique(os_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
+os_k <- names(which.min(sapply(colnames(os_clust), function(x) plot_survival(surv, os_clust[,x,drop = F], 
+                                                                             'OS.time', 'OS', T))))
+
+# combine with known pan-gi subtype annotation
+df <- merge(pfi_clust[,pfi_k,drop = F], os_clust[,os_k,drop = F], by = 'row.names')
+df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected
+df[is.na(df$subtype),]$subtype <- 'Unavailable'
+df$msi <- MSI[match(df$Row.names, bcr_patient_barcode)]$MSI
+df[is.na(df$msi),]$msi <- 'Unavailable'
+rownames(df) <- df$Row.names
+df$Row.names <- NULL
+colnames(df) <- c('Overall Survival', 'Progression Free Survival', 'Molecular Subtype', 'MSI Status')
+df <- data.frame(apply(df, 2, as.factor), check.names = F)
+df <- df[order(df[,1]),]
+m <- apply(df, 2, function(x) as.numeric(as.factor(x)))
+rownames(m) <- rownames(df)
+
+colors <- structure(1:7, names = c("1", "2", "3", "4", "5", "6", "7"))
+p1 <- Heatmap(t(m), name = "cluster", col = colors, cluster_rows = F, cluster_columns = F,
+              show_column_names = F, row_names_side = 'left',
+              column_title = 'Pan-gastrointestinal Cancer Samples\n(ordered by Overall Survival Clusters)')
+p1 <- grid::grid.grabExpr(ComplexHeatmap::draw(p1))
+# compute pairwise AMI 
+ami <- sapply(colnames(df), function(x) {
+  sapply(colnames(df), function(y) {
+    df <- df[!is.na(df[,x]) & !is.na(df[,y]),]
+    aricode::AMI(df[,x], df[,y])
+  })
+})
+
+p2 <- ggcorrplot::ggcorrplot(ami, legend.title = 'AMI', lab = T, type = 'lower', outline.color = 'white') +
+  scale_fill_gradient(limits = c(0,1), low = 'white', high = 'red') + 
+  guides(fill=guide_legend(title='AMI'))
+
+Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
+
+supFig_count <- supFig_count + 1
+SuppFigures[[supFig_count]] <- Fig
+fig.name <- paste0('SuppFigure', fig_count, '.subtype_dependency.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
+              plot = Fig, units = 'in', width = 18, height = 12)
+SuppLegends[[fig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent.
+A) Heatmap of patient cluster memberships based on different objectives: optimum stratification based on overall survival or progression free survival, molecular homogeneity (TCGA pan-GI subtypes), or micro-satellite instability (MSI) status.
+Columns are ordered first by membership in the optimum Overall Survival cluster membership.
+B) Pairwise Adjusted Mutual Information (AMI) of cluster labels obtained via different patient stratification methods.
+The AMI value ranges from 0 to 1 and the higher the value, the more overlap between patient partitions."
+```
+
 ```{r print_figures, results='asis', include=TRUE, fig.width=10, fig.height=11}
@@
 }
+
+for(i in 1:length(SuppFigures)) {
+  fig.url <- gsub("/data/bimsbstatic/public", base_url, file.path(FigureFolder, names(SuppFigures)[i]))
+  cat("# Supplementary Figure",i,"[high-res image](",fig.url,")\n\n")
+  print(SuppFigures[[i]])
+  cat("\n\n Supplementary Figure",i,":",SuppLegends[[i]],"\n\n")
+}
 ```

### 2021-03-19 src/manuscript_figures.Rmd
fix problem with supplementary legend figure counts

@@
 Legends <- list() # figure legend text
+SuppLegends <- list() # supplementary figure legend text
 
@@
 SuppFigures[[supFig_count]] <- Fig
-fig.name <- paste0('SuppFigure', fig_count, '.pancancer_subtype_prediction.pdf')
-names(Figures)[fig_count] <- fig.name
+fig.name <- paste0('SuppFigure', supFig_count, '.pancancer_subtype_prediction.pdf')
+names(SuppFigures)[supFig_count] <- fig.name
 ggsave(file.path(FigureFolder, fig.name), 
               plot = Fig, units = 'in', width = 18, height = 12)
-SuppLegends[[fig_count]] <- "Predictive power of latent factors in distinguishing known TCGA cancer subtypes A-B) comparing MAUI with PCA/MOFA/MCIA.
+SuppLegends[[supFig_count]] <- "Predictive power of latent factors in distinguishing known TCGA cancer subtypes A-B) comparing MAUI with PCA/MOFA/MCIA.
 AUC: Area under the ROC curve computed on the test dataset (30% of held-out data) trained on 70% of the dataset using a Random Forest model.
@@
 SuppFigures[[supFig_count]] <- Fig
-fig.name <- paste0('SuppFigure', fig_count, '.subtype_dependency.pdf')
-names(Figures)[fig_count] <- fig.name
+fig.name <- paste0('SuppFigure', supFig_count, '.subtype_dependency.pdf')
+names(SuppFigures)[supFig_count] <- fig.name
 ggsave(file.path(FigureFolder, fig.name), 
               plot = Fig, units = 'in', width = 18, height = 12)
-SuppLegends[[fig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent.
+SuppLegends[[supFig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent.
 A) Heatmap of patient cluster memberships based on different objectives: optimum stratification based on overall survival or progression free survival, molecular homogeneity (TCGA pan-GI subtypes), or micro-satellite instability (MSI) status.

### 2021-03-22 src/common_functions.R
add new functions to import lfs/fws, plot roc curves, read xcell genesets

@@ -1,5 +1,4 @@
 # common functions used in different analyses
-
 compute_PCA <- function(exp, topN = 50) {
   top <- head(names(sort(apply(exp, 1, sd), decreasing = T)), 5000)
   M <- t(exp[top,])
@@ -10,7 +9,7 @@ compute_PCA <- function(exp, topN = 50) {
 # given TCGA identifiers, create a table of batch information
 # see https://docs.gdc.cancer.gov/Encyclopedia/pages/TCGA_Barcode/
 getBatchTable <- function(TCGA_ids) {
-  dt <- data.table(do.call(rbind, lapply(strsplit(TCGA_ids, "-"), function(x) t(data.frame(x)))))
+  dt <- data.table::data.table(do.call(rbind, lapply(strsplit(TCGA_ids, "-"), function(x) t(data.frame(x)))))
   colnames(dt) <- c('project', 'TSS', 'participant', 'sample_vial', 'portion_analyte', 'plate', 'center')
   dt$bcr_patient_barcode <- paste(dt$project, dt$TSS, dt$participant, sep = '-')
   return(dt)
@@ -40,7 +39,7 @@ process_batch_table <- function(df, min_samples_per_batch = 10) {
     colnames(x) <- n
     return(x)
   })
-  dt <- data.table(do.call(cbind, l[!unlist(lapply(l, is.null))]))
+  dt <- data.table::data.table(do.call(cbind, l[!unlist(lapply(l, is.null))]))
   if(ncol(dt) == 0) {
     return(NULL)
   }
@@ -135,11 +134,11 @@ compute_cramers_V <- function(df) {
   do.call(rbind, lapply(1:(ncol(df)-1), function(i) {
     do.call(rbind, lapply((i+1):ncol(df), function(j) {
       if(length(unique(df[,i])) < 2 | length(unique(df[,j])) < 2) {
-        data.table('ref' = colnames(df)[i], 'target' = colnames(df)[j], 
+        data.table::data.table('ref' = colnames(df)[i], 'target' = colnames(df)[j], 
                    'cramersV' = NA, 'chisq_pval' = NA)
       } else {
         cv <- cv.test(df[,i], df[,j])
-        data.table('ref' = colnames(df)[i], 'target' = colnames(df)[j], 
+        data.table::data.table('ref' = colnames(df)[i], 'target' = colnames(df)[j], 
                    'cramersV' = cv$cramersV, 'chisq_pval' = cv$chisq_pval)
       }
     }))
[…]

### 2021-03-22 src/manuscript_figures.Rmd
simplify cluster stats, update figures for msi stats

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
+knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
 # todo: pass this as argument
@@
 ```{r import_lfs}
-LFs <- sapply(simplify = F, TOOLS, function(folder) {
-  message('Importing ',folder)
-  files <- dir(file.path(inputDir, folder), pattern = 'factors.csv$', full.names = T)
-  LFs <- pbapply::pbsapply(simplify = F, files, function(f) {
-    dt <- data.table::fread(f)
-    M <- as.matrix(dt[,-1])
-    rownames(M) <- dt$V1
-    return(M)
-  })
-  names(LFs) <- unlist(lapply(strsplit(basename(names(LFs)), split = '\\.'), function(x) paste(x[1:2], collapse = '.')))
-  return(LFs)
-})
+LFs <- import_LFs(TOOLS, inputDir)
 ```
@@
 ```{r import_feature_weights}
-FWs <- sapply(simplify = F, TOOLS, function(folder) {
-  message('Importing ',folder)
-  files <- dir(file.path(inputDir, folder), pattern = 'feature_weights.csv$', full.names = T)
-  W <- pbapply::pbsapply(simplify = F, files, function(f) {
-    dt <- data.table::fread(f)
-    M <- as.matrix(dt[,-1])
-    rownames(M) <- gsub("^assay: ", "", dt$V1) # required only for maui
-    return(M)
-  })
-  names(W) <- unlist(lapply(strsplit(basename(names(W)), split = '\\.'), 
-                            function(x) paste(x[1:2], collapse = '.')))
-  return(W)
-})
+FWs <- import_feature_weights(TOOLS, inputDir)
 ```
@@
 parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
-clusters <- sapply(simplify = F, names(LFs), function(tool) {
+cluster.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
   message("Clustering top ",tool)
-  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
+  stats <- do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, 
+                                            intersect(names(LFs[[tool]]), experiments), 
+                    function(experiment) {
     message(experiment)
     source(settings$utility_script)
-    require(data.table)
+    library(data.table)
     M <- LFs[[tool]][[experiment]]
     if(tool == 'maui_factors') {
-      M <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
-    } 
-    M <- t(subset_features_by_variance(t(M), topN = 10))
-
+      M <- remove_redundant_variables(M, perc = 99) 
+    }
+
+    # try topN from 5 to 25 (each tool may capture different amount of information 
+    # by varying number of factors)
+    stats <- do.call(rbind, lapply(seq(5, min(25, ncol(M)), 5), function(n) {
+      message(n)
+      M.sub <- t(subset_features_by_variance(t(M), topN = n))
       val <- c('internal') 
       if(ncol(M) > 2) {val <- c('internal', 'stability')}
-    res <- clValid::clValid(obj = M,
+      res <- clValid::clValid(obj = M.sub,
                             nClust = 3:8,
@@
                             maxitems = 10000,
-                          nstart = 1000)
-    memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']])))
-    colnames(memberships) <- paste0('k', colnames(memberships))
-    stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
+                            nstart = 1000, iter.max = 30)
+      stats <- data.table::as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
       colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
-    return(list('clustering_stats' = stats, 'cluster_memberships' = memberships))
-  })
-})
-parallel::stopCluster(cl)
-
-cluster.stats <- do.call(rbind, lapply(names(clusters), function(tool) {
-  do.call(rbind, lapply(names(clusters[[tool]]), function(experiment) {
-      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
-      datatypes <- strsplit(experiment, "\\.")[[1]][2]
-      stats <- clusters[[tool]][[experiment]][['clustering_stats']]
-      stats$cancer_type <- cancer_type
-      stats$datatypes <- datatypes
+      stats$top_factors <- n
+      return(stats)
+    }))
+    stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
+    stats$datatypes <- strsplit(experiment, "\\.")[[1]][2]
+    return(stats)
+  }))
   stats$method <- tool
@@
 }))
-}))
+parallel::stopCluster(cl)
 
 # plot silhouette/stability score comparison
-dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb]
+dt1 <- cluster.stats[metric == 'Silhouette'][datatypes == multi_omics_comb,
+                                             .SD[which.max(score)],
+                                             by = c('cancer_type', 'method', 'k')]
 dt1$method <- toupper(gsub("_factors", "", dt1$method))
@@
           x = 'method', y = 'score', add = 'jitter', color = 'method') + 
-  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
+  stat_compare_means(method = 'wilcox.test', paired = FALSE, #todo change to true when mofa is finished
                      comparisons = comps) + 
@@
 
-dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb]
+dt2 <- cluster.stats[metric == 'APN'][datatypes == multi_omics_comb,
+                                             .SD[which.min(score)],
+                                             by = c('cancer_type', 'method', 'k')]
 dt2$method <- toupper(gsub("_factors", "", dt2$method))
@@
           x = 'method', y = 'score', add = 'jitter', color = 'method') + 
-  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
+  stat_compare_means(method = 'wilcox.test', paired = FALSE,  #todo change to true when mofa is finished
                      comparisons = comps) + 
@@
 # combine both scores and plot for different values of k
-dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
-            cancer_type + method + k ~ metric, value.var = 'score')
-dt$kv <- paste0('kmeans(k=',dt$k,')')
-dt$method <- toupper(gsub("_factors", "", dt$method))
-p3 <- ggplot(dt,
-       aes(x = Silhouette, y = APN)) + 
-        geom_point(aes(color = method), size = 3) + 
-        facet_wrap(~ kv, ncol = 2) +
-        labs(title = 'Multi-omics clustering: Silhouette vs APN')
+# dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
+#             cancer_type + method + k ~ metric, value.var = 'score')
+# dt$kv <- paste0('kmeans(k=',dt$k,')')
+# dt$method <- toupper(gsub("_factors", "", dt$method))
+# p3 <- ggplot(dt,
+#        aes(x = Silhouette, y = APN)) + 
+#         geom_point(aes(color = method), size = 3) + 
+#         facet_wrap(~ kv, ncol = 2) +
+#         labs(title = 'Multi-omics clustering: Silhouette vs APN')
 
@@
 # PCA/MOFA win, compare against PCA
-comps <- lapply(setdiff(unique(dt$method), 'PCA'), function(x) c('PCA', x))
+comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
 p4 <- ggboxplot(dt, x = 'method', y = 'value', color = 'method',
@@
 
-Fig <- plot_grid(plot_grid(p1, p2, labels = c('A', 'B'), ncol = 1),
-                 plot_grid(p3, p4, labels = c('C', 'D'), ncol = 1), 
-                 ncol = 2, rel_widths = c(2,3))
+Fig <- plot_grid(plot_grid(p1, p2, labels = c('A', 'B'), nrow = 1),
+                 p4, nrow = 2, labels = c('', 'C'))
 
@@
 MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]
-msi_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
-  message(tool)
-  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
+
+M <- remove_redundant_variables(LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']], 
+                                perc = 99)
 y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
@@
 y <- y[!is.na(y)]
-  #res <- predict_cluster_labels(M, y, partition = 0.6, num.trees = 2000)
+
 res <- predict_cluster_labels_glmnet(M, y)
-  #res <- predict_labels_glmnet(M, y)
-  res$method <- tool
-  return(res)
-}))
-msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))
-
-p1 <- ggplot(msi_stats[method == 'MAUI'], aes(x = label, y = auc)) +
+p1 <- ggplot(res, aes(x = label, y = auc)) +
   geom_bar(stat = 'identity', aes(fill = label)) + 
@@
 
-# make some maui specific plots to explore MSI related LFs
-M <- remove_redundant_variables(LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']], 
-                                perc = 99) # make it consistent with pan-cancer
-y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI 
-M <- M[!is.na(y),]
-y <- y[!is.na(y)]
-
+# also plot a roc curve
+rocs <- predict_cluster_labels_glmnet(M, y, returnROC = TRUE)
+p2 <- plot_roc_list(rocs)
+
+# explore MSI related LFs
 # tsne plot of top LF related to MSI-H and MSI classes
@@
 
-p2.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
+p3.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
   theme(legend.title = element_blank())
-p2.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
+p3.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
   geom_point(aes_string(color = top_lf)) + 
@@
 }))
-p3 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
+p4 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
   geom_bar(aes(fill = Var1), stat='identity') + 
@@
 
-Fig <- cowplot::plot_grid(p1, 
-                          plot_grid(p2.1, p2.2, p3, nrow = 1, labels = c('B', 'C', 'D')),
-                          nrow = 2, labels = c('A', ''), rel_heights = c(5,3))
+Fig <- cowplot::plot_grid(plot_grid(p1, p2, ncol = 1, labels = c('A', 'B')),
+                          plot_grid(p3.1, p3.2, p4, ncol = 1, labels = c('C', 'D', 'E')),
+                          nrow = 1)
 fig_count <- fig_count + 1
@@
 ggsave(filename = file.path(FigureFolder, fig.name), 
-       plot = Fig, units = 'in', width = 10, height = 8)
+       plot = Fig, units = 'in', width = 8, height = 8)
 Legends[[fig_count]] <- "Predicting MSI status of pan-gastrointestinal tumors using MAUI latent factors.
@@
 ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
-
-res <- predict_cluster_labels_glmnet(M, ct)
 
@@
 
+bipartite_plots <- unlist(lapply(1:(ncol(df)-1), function(i) {
+  lapply((i+1):ncol(df), function(j) {
+    p <- plot_cluster_comparison(df[,i,drop = F], df[,j,drop = F]) + 
+      theme(plot.margin = unit(c(0.2, 1.5, 0, 1.5), units = 'in'))
+  })
+}), recursive = F)
+
+Fig1 <- cowplot::plot_grid(plotlist = bipartite_plots, nrow = 3, labels = 'AUTO')
+ggsave('test1.pdf', plot = Fig1, units = 'in', width = 12, height = 14)
+
 Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
+#ggsave('test2.pdf', plot = Fig2, units = 'in', width = 14, height = 12)
 

### 2021-03-23 src/run_MCIA.R
use tool configurations in snakefile, update tool scripts

@@ -5,12 +5,12 @@ library(omicade4)
 
 Assays <- args[1] # path to RDS object containing a list of assay matrices
 outFile <- args[2] # path to output file
+nFactors <- args[3] # number of factors to compute
 
 A <- readRDS(Assays)
-
-nf <- 20
-message(date(), " => Running MCIA looking for ",nf," factors")
-mcoin <- mcia(A, cia.nf=nf)
+nFactors <- as.numeric(nFactors)
+message(date(), " => Running MCIA looking for ",nFactors," factors")
+mcoin <- mcia(A, cia.nf=nFactors)
 
 LF <- mcoin$mcoa$SynVar
 rownames(LF) <- gsub("\\.", "-", rownames(LF))

### 2021-03-23 src/run_PCA.R

@@ -1,9 +1,10 @@
 args <- commandArgs(trailingOnly = TRUE)
 Assays <- args[1] # path to a CSV file
 outFile <- args[2] # path to output file
+nFactors <- args[3] # number of components to compute
 source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
 M <- as.matrix(read.csv(Assays, check.names = FALSE))
-PCA <- compute_PCA(M, topN = 100)
+PCA <- compute_PCA(M, topN = as.numeric(nFactors))
 write.table(x = PCA$x, 
             file = outFile,
             quote = F, row.names = T, sep = ',')

### 2021-03-23 src/run_mofa.R

@@ -9,12 +9,13 @@ library(MOFA2)
 
 Assays <- args[1] # path to RDS object containing a list of assay matrices
 outFile <- args[2] # path to output file
-topFeatures <- args[3] # number of top features to use in training
+nFactors <- args[3] # number of factors to compute
 
 A <- readRDS(Assays)
 MOFAobject <- create_mofa(A)
 modelOptions <- MOFA2::get_default_model_options(MOFAobject)
-modelOptions$num_factors <- min(100, floor(median(sapply(A, ncol))/4)) # set to 0.25 of number of samples 
+# number of factors is maximally set to one forth of the total number of samples
+modelOptions$num_factors <- min(as.numeric(nFactors), floor(median(sapply(A, ncol))/4)) 
 MOFAobject <- prepare_mofa(MOFAobject, model_options = modelOptions)
 message(date(), " training MOFA")
 MOFAobject.trained <- run_mofa(MOFAobject, use_basilisk = TRUE)

### 2021-03-23 src/snakefile.py

@@ -16,8 +16,8 @@ rule all:
         expand(os.path.join(OUTDIR, 'mofa_factors', "{assay}.mofa_factors.feature_weights.csv"), assay = ASSAYS),
         expand(os.path.join(OUTDIR, 'pca_factors', "{assay}.pca_factors.csv"), assay = ASSAYS),
         expand(os.path.join(OUTDIR, 'pca_factors', "{assay}.pca_factors.feature_weights.csv"), assay = ASSAYS),
-	expand(os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.csv"), assay = ASSAYS), 
-	expand(os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.feature_weights.csv"), assay = ASSAYS)
+        expand(os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.csv"), assay = ASSAYS), 
+        expand(os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.feature_weights.csv"), assay = ASSAYS)
 
 rule run_mcia:
     input:
@@ -27,9 +27,10 @@ rule run_mcia:
         os.path.join(OUTDIR, 'mcia_factors', "{assay}.mcia_factors.feature_weights.csv")
     log: os.path.join(LOG_DIR, "run_mcia.{assay}.log")
     params:
-        script = os.path.join(SRCDIR, 'run_MCIA.R')
+        script = os.path.join(SRCDIR, 'run_MCIA.R'), 
+        nfactors = config['mcia']['nfactors']
     shell:
-        "{RSCRIPT} {params.script} {input} {output[0]} > {log} 2>&1"
+        "{RSCRIPT} {params.script} {input} {output[0]} {params.nfactors} > {log} 2>&1"
 
 rule run_maui:
     input:
@@ -39,9 +40,12 @@ rule run_maui:
         os.path.join(OUTDIR, 'maui_factors', "{assay}.maui_factors.feature_weights.csv")
     log: os.path.join(LOG_DIR, "run_maui.{assay}.log")
     params:
-        script = os.path.join(SRCDIR, 'run_maui.sh')
+        script = os.path.join(SRCDIR, 'run_maui.sh'),
+        nfactors = config['maui']['nfactors'],
+        epochs = config['maui']['epochs'],
+        threads = config['maui']['threads']
     shell:
[…]

### 2021-03-23 src/common_functions.R
update prepareData function to filter for genes of interest with necessary id mapping

@@ -736,7 +736,7 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
                         topN = 1000, GOI = NULL,
                         TCGA_Disease_IDs, datatypes = c('gex', 'cnv', 'mut', 'meth'),
                         gex_flavor, 
-                        cnv_flavor) {
+                        cnv_flavor, ens2hgnc, cpg2hgnc) {
   patientIDs <- unique(clin[project %in% TCGA_Disease_IDs]$bcr_patient_barcode)
   dataList <- list('assay' = list(), 
                    'batch' = list())
@@ -762,10 +762,11 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     common_features <- Reduce(intersect, lapply(mut, rownames))
     # reorder features and concatenate columns
     mut <- do.call(cbind, lapply(mut, function(x) x[common_features,]))
-    if(!is.null(GOI)) {mut <- mut[intersect(rownames(mut), GOI),]}
-    
-    # remove redundant features
-    #mut <- t(remove_redundant_variables(t(mut), perc = 99, downsize_byTopVar = 15000))
+    if(!is.null(GOI)) {
+      # we assume genes have ensemble ids and GOI are gene symbols
+      s <- intersect(rownames(mut), ens2hgnc[hgnc_symbol %in% GOI]$ref_gene_id) # keep only GOI
+      mut <- mut[s,]
+    }
     
     # select genes with most non-zero samples
     selected <- head(names(sort(apply(mut, 1, function(x) sum(x > 0)), decreasing = T)), topN)
@@ -804,12 +805,11 @@ prepareData <- function(dataDir, clin, correctBatchEffects = TRUE,
     common_features <- Reduce(intersect, lapply(gex, rownames))
     gex <- do.call(cbind, lapply(gex, function(x) x[common_features,]))
     
-    # subset by features
-    if (!is.null(GOI)) {gex <- gex[intersect(rownames(gex), GOI),]}
-    
-    # remove redundant features
-    #gex <- t(remove_redundant_variables(t(gex), perc = 99, downsize_byTopVar = 15000))
-    
+    if (!is.null(GOI)) {
+      # we assume genes have ensemble ids and GOI are gene symbols
+      s <- intersect(rownames(gex), ens2hgnc[hgnc_symbol %in% GOI]$ref_gene_id) # keep only GOI
+      gex <- gex[s,]
+    }
     
[…]

### 2021-03-23 src/setup_experiments.R
pass id mapping tables as argument

@@ -13,9 +13,8 @@ source(settings$utility_script)
 assayDir <- settings$assay_output$folder
 if(settings$features$use_geneset == TRUE) {
   ens2hgnc <- readRDS(settings$ens2hgnc)
-  geneSets <- read_msigdb(settings$geneset)
-  geneSets.ensembl <- map_ens2hgnc(geneSets, ens2hgnc, "hgnc2ens")
-  GOI <- unique(unlist(geneSets.ensembl))
+  cpg2hgnc <- readRDS(settings$cpg2hgnc)
+  GOI <- readLines(settings$geneset)
 } else {
   GOI <- NULL
 }
@@ -35,7 +34,7 @@ if(!dir.exists(assayDir)) {
 
 cl <- parallel::makeCluster(length(projects), outfile = 'setup_experiments.log')
 parallel::clusterExport(cl = cl, varlist = c('settings', 'projectDict', 'dataDir', 
-                                             'surv', 'assayDir', 'GOI'))
+                                             'surv', 'assayDir', 'GOI', 'ens2hgnc', 'cpg2hgnc'))
 pbapply::pblapply(cl = cl, projects, function(pr) {
   require(data.table)
   source(settings$utility_script)
@@ -47,7 +46,9 @@ pbapply::pblapply(cl = cl, projects, function(pr) {
                           TCGA_Disease_IDs = projectDict[[pr]], 
                           datatypes = settings$datatypes, 
                           gex_flavor = settings$datatype_flavors$gex, 
-                          cnv_flavor = settings$datatype_flavors$cnv) 
+                          cnv_flavor = settings$datatype_flavors$cnv, 
+                          ens2hgnc = ens2hgnc, #id mapping 
+                          cpg2hgnc = cpg2hgnc) #id mapping 
   # save assays 
   lapply(sapply(settings$datatype_combinations, function(x) strsplit(x, ',')), 
          function(datatypes) {

### 2021-03-25 src/common_functions.R
add function: compute_geneset_enrichment

@@ -1426,3 +1426,29 @@ read_xCell_geneSets <- function(f, group_by_cell_type = FALSE) {
   
   return(geneSets)
 }
+
+# given a vector of genes and a list of gene sets, compute which gene sets are over-represented in the given gene pool
+# genes: query set of genes
+# geneSets: list of gene sets
+# background: vector of gene universe (optional. if not provided, union of geneSets is used as background)
+compute_geneset_enrichment <- function(genes, geneSets, background = NULL) {
+  if(is.null(background)) {
+    background <- unique(unlist(geneSets))
+  }
+  res <- do.call(rbind, lapply(names(geneSets), function(gs) {
+    x <- geneSets[[gs]]
+    o <- length(intersect(genes, x))
+    p <- length(intersect(x, background))/length(background) # probability of overlap
+    data.table('term' = gs, 'overlap' = o, 'query_size' = length(genes), 'target_size' = length(x), 
+               'expected_overlap' = round(p * length(genes), 2), 
+               'prob' = p, 
+               'pval' = binom.test(x = o, n = length(genes), p = p, alternative = 'greater')[['p.value']])
+  }))
+  res$padj <- p.adjust(res$pval, method = 'BH')
+  return(res)
+}
+
+
+
+
+

### 2021-03-25 src/common_functions.R
add event/time arguments to survival functions

@@ -67,7 +67,7 @@ fit_RSF <- function(surv, M, ntree = 1000,
 }
 
 # given a matrix of numerical variables and a data.frame of clinical variables 
-# including PFI.time (OS.time) or PFI (OS)
+# including time and status
 # fit a random survival forest and compute c index
 fit_RSF_clin <- function(clin, M, ntree = 1000) {
   require(data.table)
@@ -82,10 +82,10 @@ fit_RSF_clin <- function(clin, M, ntree = 1000) {
     return(dat[,x,drop = F])
   }))
   # fit with/without latent factors
-  fit1 <- randomForestSRC::rfsrc(formula = Surv(PFI.time, PFI) ~ .,
+  fit1 <- randomForestSRC::rfsrc(formula = Surv(time, status) ~ .,
                                 data = dat, ntree = ntree)
   Cindex1 <- round(1 - mean(fit1$err.rate, na.rm = T), 2)
-  fit2 <- randomForestSRC::rfsrc(formula = Surv(PFI.time, PFI) ~ .,
+  fit2 <- randomForestSRC::rfsrc(formula = Surv(time, status) ~ .,
                                  data = subset(dat, select = colnames(clin)), 
                                  ntree = ntree)
   Cindex2 <- round(1 - mean(fit2$err.rate, na.rm = T), 2)
@@ -1096,11 +1096,11 @@ get_prognostic_covariates <- function(surv, clin = NULL, explored_covariates,
   }
 }
 
-get_surv_df <- function(surv, M) {
+get_surv_df <- function(surv, M, event_col = 'PFI', duration_col = 'PFI.time') {
   df <- as.data.frame(M)
-  df$PFI <- surv[match(rownames(df), bcr_patient_barcode)]$PFI
-  df$PFI.time <- surv[match(rownames(df), bcr_patient_barcode)]$PFI.time
-  df <- df[!is.na(df$PFI.time),]
+  df <- cbind(df, data.frame('status' = surv[match(rownames(df), bcr_patient_barcode)][,get(event_col)], 
+                             'time' = surv[match(rownames(df), bcr_patient_barcode)][,get(duration_col)]))
+  df <- df[!is.na(df$time),]
   return(df)
 }
 
@@ -1162,13 +1162,13 @@ process_tcga_subtypes <- function() {
 
 # df: data.frame with a single numerical column with sample ids in rows 
 # function finds the cut point in the first column of df that maximizes survival separation into two
[…]

### 2021-03-25 src/manuscript_figures.Rmd
update survival analysis - use RSF to filter variables

@@
 # todo: pass this as argument
-settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
+settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/hallmark_genes/settings.yaml')
 source(settings$utility_script)
@@
 geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
-FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas'
+FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas/test'
 
@@
   if(x == 'maui_factors') {
-    M <- remove_redundant_variables(M, perc = 99)
+    #M <- remove_redundant_variables(M, perc = 99)
+    M <- scale(M)
   }
@@
     # by varying number of factors)
-    stats <- do.call(rbind, lapply(seq(5, min(25, ncol(M)), 5), function(n) {
+    stats <- do.call(rbind, lapply(seq(5, min(20, ncol(M)), 5), function(n) {
       message(n)
@@
       val <- c('internal') 
-      if(ncol(M) > 2) {val <- c('internal', 'stability')}
+      if(ncol(M.sub) > 2) {val <- c('internal', 'stability')}
       res <- clValid::clValid(obj = M.sub,
@@
 
-# combine both scores and plot for different values of k
-# dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
-#             cancer_type + method + k ~ metric, value.var = 'score')
-# dt$kv <- paste0('kmeans(k=',dt$k,')')
-# dt$method <- toupper(gsub("_factors", "", dt$method))
-# p3 <- ggplot(dt,
-#        aes(x = Silhouette, y = APN)) + 
-#         geom_point(aes(color = method), size = 3) + 
-#         facet_wrap(~ kv, ncol = 2) +
-#         labs(title = 'Multi-omics clustering: Silhouette vs APN')
-
 # survival prediction accuracy benchmarking
 # find survival-predictive LFs by univariate cox models and use them in RSF
+clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
 cl <- parallel::makeCluster(20)
-parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv'))
+parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
 surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
@@
     M <- LFs[[tool]][[experiment]]
-    if(tool == 'maui_factors') {
-      M <- remove_redundant_variables(M, 99)
-    }
-    dt <- data.table('cindex.pfi' = NA, 'cindex.os' = NA, 'experiment' = experiment, 'method' = tool)
+    clin <- process_clinical_covariates(rownames(M))
+    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
     # compute pfi stats
-    vars <- get_prognostic_covariates(surv, clin = NULL, explored_covariates = M,
-                                      event_col = 'PFI', duration_col = 'PFI.time')
-    if(!is.null(vars)) {
-      v <- head(vars[order(pval)][pval < 0.05][['variable']], 10)
-      fit <- fit_RSF(surv, M[,v], 1000, 'PFI.time', 'PFI')
-      dt$cindex.pfi <- round(1 - mean(fit$err.rate, na.rm = T), 2)
-    }
-    vars <- get_prognostic_covariates(surv, clin = NULL, explored_covariates = M,
-                                      event_col = 'OS', duration_col = 'OS.time')
-    if(!is.null(vars)) {
-      v <- head(vars[order(pval)][pval < 0.05][['variable']], 10)
-      fit <- fit_RSF(surv, M[,v], 1000, 'OS.time', 'OS')
-      dt$cindex.os <- round(1 - mean(fit$err.rate, na.rm = T), 2)
-    }
-    return(dt)
+    # fit a first model to filter variables 
+    fit <- fit_RSF(surv, M, ntree = 500, 'PFI.time', 'PFI')
+    # using a range of top variables, compute c index
+    dt1 <- do.call(rbind, lapply(2:10, function(i) {
+      vars <- names(head(sort(fit$importance, decreasing = T), i))
+      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'PFI', 'PFI.time'), 
+                                M[,vars], ntree = 500)
+      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
+                       'experiment' = experiment, 'method' = tool, 'topN' = i)
+    }))
+    dt1$endpoint <- 'PFI'
+    # do the same for overall survival
+    fit <- fit_RSF(surv, M, ntree = 500, 'OS.time', 'OS')
+    dt2 <- do.call(rbind, lapply(2:10, function(i) {
+      vars <- names(head(sort(fit$importance, decreasing = T), i))
+      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'OS', 'OS.time'), 
+                                M[,vars], ntree = 500)
+      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
+                       'experiment' = experiment, 'method' = tool, 'topN' = i)
+    }))
+    dt2$endpoint <- 'OS'
+    return(rbind(dt1, dt2))
   }))
@@
 
-dt <- melt(surv.stats, measure.vars = c('cindex.pfi', 'cindex.os'))
-dt[is.na(value)]$value <- min(dt$value, na.rm = T) # 
+dt <- surv.stats
 dt$method <- toupper(gsub("_factors", "", dt$method))
 dt$project <- unlist(sapply(strsplit(dt$experiment, "\\."), function(x) x[1]))
-dt$endpoint <- ifelse(dt$variable == 'cindex.pfi', 'Progression Free Survival', 
+dt$endpoint <- ifelse(dt$endpoint == 'PFI', 'Progression Free Survival', 
                       'Overall Survival')
+# get best performance per tool per project
+dt <- dt[,.SD[which.max(cindex)], by = c('project', 'method', 'endpoint')]
 # PCA/MOFA win, compare against PCA
 comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
-p4 <- ggboxplot(dt, x = 'method', y = 'value', color = 'method',
+p4 <- ggboxplot(dt, x = 'method', y = 'cindex', color = 'method',
                 add = 'jitter', facet.by = 'endpoint') +
@@
 clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
+event <- 'PFI'
+time <- 'PFI.time'
 
@@
 cl <- parallel::makeCluster(20)
-parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 
+parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'event', 'time',
                                         'surv', 'experiments', 'settings'))
@@
   require(data.table)
+  require(survminer)
+  require(survival)
   source(settings$utility_script)
@@
   clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
-    res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
-    if(!is.null(res)) {
-      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
-      res$project <- cancer_type
-      v <- res[order(pval)][1]$variable
+  # first filter variables by importance 
+  fit <- fit_RSF(surv, M, 500, time, event)
+  vars <- names(head(sort(fit$importance, decreasing = T), 5))
+
+  # compute hazard ratios for the top factor
+  v <- vars[1] 
   dat <- merge(clin, M[,v,drop = F], by = 'row.names')
@@
   dat$Row.names <- NULL
-      dat <- get_surv_df(surv, dat) # add PFI values
+  dat <- get_surv_df(surv, dat, event, time)
   # compute hazard ratios 
-      cox1 <- coxph(Surv(PFI.time,PFI) ~ ., 
+  cox1 <- coxph(Surv(time,status) ~ ., 
           data=dat, x = TRUE)
@@
   # split variable into high/low to compute survival adjusted curves
-      cp <- find_survival_cutpoint(dat[,v,drop=F], surv)
+  cp <- find_survival_cutpoint(dat[,v,drop=F], surv, event_col = event, duration_col = time)
   dat[,v] <- ifelse(dat[,v] > cp, paste(v, 'high'), paste(v, 'low'))
-      cox2 <- coxph(Surv(PFI.time,PFI) ~ ., 
+  cox2 <- coxph(Surv(time,status) ~ ., 
                 data=dat, x = TRUE)
@@
   # compute Cindex with all survival-predictive LFs and without LFs 
-      vars <- res[pval < 0.05]$variable
-      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin), M[,vars], ntree = 1000)
-
-      return(list('stats' = res, 'p1' = p1, 'p2' = p2, 'hr' = hr, 
+  rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, event, time), 
+                            M[,vars], ntree = 1000)
+
+  return(list('p1' = p1, 'p2' = p2, 'hr' = hr, 
               'Cindex_base' = rsf_stats$C_wo, 'Cindex_with_LFs' = rsf_stats$C_with))
-    }
+
   return(NULL)
@@
     clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
-    res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
+
+    fit <- fit_RSF(surv, M, duration_col = 'PFI.time', event_col = 'PFI')
+    prog_lfs <- names(head(sort(fit$importance, decreasing = T), 5))
     # now cluster samples based on only survival predictive factors
-    prog_lfs <- unique(res$variable)
-    if(length(prog_lfs) < 2) {
-      return(NULL)
-    }
     clust <-  get_kmeans_subtypes(M[,prog_lfs], k_vals = 3:8, min_cluster_size = 0)
@@
 
-Fig1 <- cowplot::plot_grid(plotlist = bipartite_plots, nrow = 3, labels = 'AUTO')
-ggsave('test1.pdf', plot = Fig1, units = 'in', width = 12, height = 14)
-
 Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
-#ggsave('test2.pdf', plot = Fig2, units = 'in', width = 14, height = 12)
-
 supFig_count <- supFig_count + 1
@@
 The AMI value ranges from 0 to 1 and the higher the value, the more overlap between patient partitions."
+
+Fig <- cowplot::plot_grid(plotlist = bipartite_plots, nrow = 3, labels = 'AUTO')
+supFig_count <- supFig_count + 1
+SuppFigures[[supFig_count]] <- Fig
+fig.name <- paste0('SuppFigure', supFig_count, '.subtype_dependency.bipartite_plots.pdf')
+names(SuppFigures)[supFig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
+              plot = Fig, units = 'in', width = 14, height = 14)
+SuppLegends[[supFig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent"
 ```

### 2021-03-25 src/manuscript_figures.Rmd
disable correlation filtering - apply scaling instead

@@
     if(tool == 'maui_factors') {
-      M <- remove_redundant_variables(M, perc = 99) 
+      #M <- remove_redundant_variables(M, perc = 99) 
+      M <- scale(M)
     }
@@
 
-M <- remove_redundant_variables(LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']], 
-                                perc = 99)
+M <- LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']]
+M <- scale(M) #remove_redundant_variables(M, perc = 99)
 y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
@@
 # reference study: https://www.nature.com/articles/s41598-020-77284-8
-M <- remove_redundant_variables(LFs$maui_factors$nsclc.cnv_gex_meth_mut, perc = 99)
+M <- LFs$maui_factors$nsclc.cnv_gex_meth_mut
+M <- scale(M) #remove_redundant_variables(M, perc = 99)
 ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
@@
 experiment <- paste0(pr, '.cnv_gex_meth_mut')
-M <- remove_redundant_variables(LFs$maui_factors[[experiment]], perc = 99)
+M <- LFs$maui_factors[[experiment]]
+M <- scale(M) #remove_redundant_variables(, perc = 99)
 st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
@@
 M <- LFs$maui_factors[[experiment]]
-M <- remove_redundant_variables(M, perc = 99)
-
 clin <- process_clinical_covariates(rownames(M))

### 2021-03-25 src/manuscript_figures.Rmd
replace AMI computation, reduce parallel threads to 10

@@
   html_document: default
+params:
+  settings_file: ''
+  figures_folder: '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas/test'
+  workdir: '.'
 ---
@@
 knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
-# todo: pass this as argument
-settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/hallmark_genes/settings.yaml')
+knitr::opts_knit$set(root.dir = params$workdir)
+settings <- yaml::read_yaml(params$settings_file)
+FigureFolder <- params$figures_folder
 source(settings$utility_script)
@@
 geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
-FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas/test'
 
@@
 
-TOOLS <- c('pca_factors', 'maui_factors', 'mofa_factors', 'mcia_factors')
+TOOLS <- c('maui_factors', 'mofa_factors', 'mcia_factors', 'pca_factors')
 
@@
 
-# do unsupervised clustering and compute AMI 
-experiment <- 'pancancer.cnv_gex_meth_mut'
-df <- do.call(rbind, pbapply::pblapply(TOOLS, function(x) {
-  message(x)
-  M <- LFs[[x]][[experiment]]
-  if(x == 'maui_factors') {
-    #M <- remove_redundant_variables(M, perc = 99)
+# do unsupervised clustering experiment and compute AMI 
+# downsample pan-cancer dataset 
+df <- data.frame('id' = rownames(LFs$pca_factors$pancancer.cnv_gex_meth_mut))
+df$pr <- surv[match(df$id, bcr_patient_barcode)]$project
+ids <- lapply(split(df, df$pr), function(x) {
+  if(nrow(x) > 200) {
+    sample(x$id, 200)
+  }
+})
+ids <- ids[lengths(ids) == 200]
+k <- 5  
+# sample cancer types and compute ami for different tools 
+sampled <- sapply(simplify = F, 1:50, function(i) {
+  unique(unlist(ids[sample(1:length(ids), k)]))
+})
+cl <- parallel::makeCluster(10)
+parallel::clusterExport(cl, varlist = c('ids', 'df', 'LFs', 'settings', 'sampled', 'k'))
+res <- do.call(rbind, pbapply::pblapply(cl = cl, 1:length(sampled), function(i) {
+  require(data.table)
+  source(settings$utility_script)
+  selected <- sampled[[i]]
+  ct <- df[match(selected, df$id),]$pr
+  message("selected cancers:",paste0(unique(ct), collapse = ','))
+  do.call(rbind, lapply(names(LFs), function(tool) {
+    M <- scale(LFs[[tool]]$pancancer.cnv_gex_meth_mut[selected,])
+    if(tool == 'maui_factors') {
       M <- scale(M)
     }
-  # find cancer types per sample
-  ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
-  data.frame('tool' = x, 
-             'AMI' = compute_AMI(M, k = length(unique(ct)), target_labels = ct))
-}))
-df$tool <- toupper(gsub("_factors", "", df$tool))
-p5 <- ggplot(df, aes(x = tool, y = AMI)) +
-  geom_bar(aes(fill = tool), stat = 'identity') + 
-  geom_text(aes(label = round(AMI, 2))) + 
-  theme(legend.title = element_blank(), axis.title.x = element_blank())
+    ami <- compute_AMI(M, k, ct)
+    return(data.table('i' = i, 'ami' = ami, 'method' = tool, 
+                      'pr' = paste0(unique(ct), collapse = ',')))
+  }))
+}))
+parallel::stopCluster(cl)
+res$method <- toupper(gsub("_factors", "", res$method))
+
+comps <- lapply(setdiff(unique(res$method), 'MAUI'), function(x) c('MAUI', x))
+p5 <- ggboxplot(res, x = 'method', y = 'ami', add = 'jitter', color = 'method') + 
+  stat_compare_means(method = 'wilcox.test', paired = TRUE,
+                     comparisons = comps) +
+  labs(y = "Adjusted Mutual Information", x = '') +
+  theme(legend.position = 'none')
 
@@
 experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
-cl <- parallel::makeCluster(20)
+cl <- parallel::makeCluster(10)
 parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
@@
 clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-cl <- parallel::makeCluster(20)
+cl <- parallel::makeCluster(10)
 parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
@@
 experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
-cl <- parallel::makeCluster(20)
+cl <- parallel::makeCluster(10)
 parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 

### 2021-03-25 src/setup_experiments.R
read id mapping in any case

@@ -11,9 +11,9 @@ settings <- yaml::read_yaml(argv$settings)
 source(settings$utility_script)
 
 assayDir <- settings$assay_output$folder
+ens2hgnc <- readRDS(settings$ens2hgnc)
+cpg2hgnc <- readRDS(settings$cpg2hgnc)
 if(settings$features$use_geneset == TRUE) {
-  ens2hgnc <- readRDS(settings$ens2hgnc)
-  cpg2hgnc <- readRDS(settings$cpg2hgnc)
   GOI <- readLines(settings$geneset)
 } else {
   GOI <- NULL

### 2021-03-25 src/manuscript_figures.Rmd
plot overall survival

@@
 clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-event <- 'PFI'
-time <- 'PFI.time'
+event <- 'OS'
+time <- 'OS.time'
 

### 2021-05-31 src/manuscript_figures.Rmd
add section about immunotherapy response prediction

@@
 library(pheatmap)
+library(survminer)
+library(survival)
 theme_set(ggpubr::theme_pubclean())
@@
   do.call(rbind, lapply(names(LFs), function(tool) {
-    M <- scale(LFs[[tool]]$pancancer.cnv_gex_meth_mut[selected,])
+    M <- LFs[[tool]]$pancancer.cnv_gex_meth_mut[selected,]
     if(tool == 'maui_factors') {
@@
                    y = base, yend = with_lf)) + 
-  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 6) +
+  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 2) +
   coord_flip() + 
@@
 # use top predictive lfs to guide tsne 
-top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 5)[['variable']])
+top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 10)[['variable']])
 
@@
 # maui_clust: best survival prognostic clustering of M 
-plot_clusts <- function(M, maui_clust, scores) {
+plot_clusts <- function(surv, M, maui_clust, scores) {
   df <- plot_tsne(M, as.factor(maui_clust[,1]), returnData = T)
@@
 
+  message("running plot clusts")
+  message("p1")
   p1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
@@
 
+  message("p2")
   p2 <- plot_survival(surv, 
@@
   dt[padj < 0.05][order(padj)]
+  message("h1")
+
   h1 <- plot_label_specific_vars(S[selected,], 
@@
 S <- scores[intersect(rownames(M), rownames(scores)),]
-P <- plot_clusts(M, maui_clust, S)
+P <- plot_clusts(surv, M, maui_clust, S)
 
@@
 
+message("Got the figure")
+
 fig_count <- fig_count + 1
 Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.', experiment,'.pdf')
+fig.name <- paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.pdf')
 names(Figures)[fig_count] <- fig.name
@@
 
+```{r immunotherapy_response}
+# immunotherapy response prediction analysis: 
+# TODO: pass paths via args
+settings_imtrp <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/immunotherapy_response/Mariathasan_et_al/settings.yaml')
+pathways <- readRDS('/data/local/buyar/arcas/subtyping_paper/data/GeneSets/msigdb.pathways.RDS')
+
+LFs_imtrp <- import_LFs(c('maui_factors', 'mofa_factors', 'pca_factors'), settings_imtrp$pipeline_output)
+FWs_imtrp <- import_feature_weights(c('maui_factors'), settings_imtrp$pipeline_output)
+colData <- data.table::fread(settings_imtrp$clin)
+colnames(colData)[1] <- 'bcr_patient_barcode'
+
+dt <- do.call(rbind, lapply(names(LFs_imtrp), function(tool) {
+  M <- LFs_imtrp[[tool]]$mariathasan.gex
+  # find LFs predictive of response to therapy 
+  dt <- get_factor_specific_variables(M, colData$binaryResponse)[order(padj)]
+  #dt <- compute_anova(M, colData$binaryResponse)
+  dt$tool <- tool
+  return(dt)
+}))
+
+dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')]
+
+top_vars <- unique(dt$variable)
+
+# quantify prediction accuracy for finding complete/partial responders
+m <- M[,top_vars]
+f <- colData$binaryResponse
+m <- m[!is.na(f),]
+f <- f[!is.na(f)]
+
+# heatmap of top vars 
+p1 <- pheatmap::pheatmap(get_basis_matrix(m[,top_vars], f), silent = T)
+
+# survival stratification using the top LF 
+require(survminer)
+require(survival)
+top_lf <- dt[order(padj)][ref_cl == 'CR/PR'][1,]$variable
+cp <- find_survival_cutpoint(m[,top_lf,drop = F], colData, event_col = 'censOS', duration_col = 'os')
+
+p2 <- plot_survival(surv = colData, partition.df = data.frame('group' = ifelse(m[,top_lf] > cp, 'high', 'low'), 
+                                                        row.names = rownames(m)), 
+              duration_col = 'os', event_col = 'censOS')
+
+# top go terms for top features of top LF 
+feats <- data.table(W[,top_lf,drop=F], keep.rownames = T)[order(abs(get(top_lf)), decreasing = T)][1:100]
+# top pathways for top lf
+top_terms <- compute_geneset_enrichment(feats$rn, pathways)[order(pval)][padj < 0.05][1:10]
+top_terms$term_wr <- stringr::str_wrap(gsub("_", " ", top_terms$term), width = 32)
+p3 <- ggplot(top_terms, aes(x = reorder(term_wr, -pval), y = -log10(padj))) + 
+  geom_bar(stat = 'identity') + 
+  coord_flip() + labs(x = '') 
+colnames(feats)[2] <- 'top_lf'
+p4 <- ggplot(feats[1:10,], aes(x = reorder(rn, abs(top_lf)), y = abs(top_lf))) + 
+  geom_bar(stat = 'identity') + coord_flip() + labs(x = '', y = 'Feature Importance (absolute)')
+
+# correlation of top LF with tumor mutation burden
+# get samples with both response and tmb annotation
+cd <- colData #colData[!is.na(`Neoantigen burden per MB`)][!is.na(binaryResponse)]
+m2 <- M[cd$bcr_patient_barcode,]
+require(ggpubr)
+df <- data.frame('response' = cd$binaryResponse, 
+                 'TMB' = cd$`Neoantigen burden per MB`, 
+                 'LF' = m2[,top_lf])
+p5 <- ggpubr::ggscatter(df, x = 'LF', y = 'TMB', 
+                        add = 'reg.line', cor.coef = T, cor.coef.coord = c(30, 10), 
+                  cor.coef.size = 4) + labs(x = top_lf, y = 'Tumor Mutation Burden')
+
+# immune phenotypes 
+ip <- colData$`Immune phenotype`
+#ip[ip == 'excluded'] <- NA
+m3 <- M[!is.na(ip),]
+ip <- ip[!is.na(ip)]
+
+dt3 <- get_factor_specific_variables(m3, ip)[order(pval)][padj < 0.01]
+ip_vars <- unique(dt3$variable)
+ip_vars <- get_top_by_label(dt3, 'ref_cl', topN = 5)
+res_ip <- predict_cluster_labels_glmnet(m3[,unique(dt3$variable)], ip, returnROC = T)
+p6 <- plot_roc_list(res_ip)
+
+pheatmap::pheatmap(get_basis_matrix(m3[,unique(ip_vars$variable)], ip), silent = T)
+
+Fig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2$gtable, p4, p3$plot, 
+                                             labels = 'AUTO'), 
+                          cowplot::plot_grid(p5, p6, p7, nrow = 1, 
+                                             labels = c('E', 'F', 'G')), 
+                          nrow = 2, rel_heights = c(2, 1)) 
+fig_count <- fig_count + 1
+Figures[[fig_count]] <- Fig
+fig.name <- paste0('Figure', fig_count, '.immunotherapy_response.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(filename = file.path(FigureFolder, fig.name), 
+       plot = Fig, units = 'in', 
+       width = 10, height = 14)
+Legends[[fig_count]] <- "Predicting responders to anti-PD-L1 treatment in a cohort of Metastatic Urothelial
+Cancers (N=348) using MAUI. A) AUC curve for predicting Complete/Partial Responders (CR/PR) to the anti-PD-L1
+treatment using an Elastic Net Model on a subset of MAUI latent factors individually predictive of treatment response groups.
+B) Average activation score of top latent factors predictive of treatment response groups.
+C) Top pathways enriched among top 100 genes that contribute to the top latent factor (LF44).
+D) Survival stratification of the
+cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4). E) Top 10 genes contributing to LF44
+F) Correlation between LF44 activation score and the tumor (neo-antigen) mutation burden"
+```
+
 ```{r print_figures, results='asis', include=TRUE, fig.width=10, fig.height=11}

### 2021-05-31 src/common_functions.R
add function for multiclass label prediction

@@ -285,6 +285,43 @@ predict_cluster_labels_glmnet <- function(M, labels, returnROC = FALSE) {
   }
   return(res)
 }
+
+
+predict_labels_glmnet_multiclass <- function(M, labels, partition = 0.6, nodes = 5) {
+  df <- data.frame(M)
+  df$y <- as.factor(labels)
+  set.seed(3456)
+  intrain <- caret::createDataPartition(y = df$y, p= partition, list = FALSE)
+  training <- df[intrain,]
+  testing <- df[-intrain,]
+  # Build the model using the training set
+  fit.control <- trainControl(method = "cv", number = 5, 
+                              summaryFunction = multiClassSummary, classProbs = TRUE)
+  set.seed(123)
+  require(doParallel)
+  cl <- makePSOCKcluster(nodes)
+  registerDoParallel(cl)
+  model <- train(
+    y ~., data = training, method = "glmnet",
+    trControl = fit.control, 
+    metric = 'logLoss', #for multi-class prediction
+    tuneLength = 10
+  )
+  stopCluster(cl)
+  
+  # Best tuning parameter
+  message("Best model parameters: alpha=", model$bestTune$alpha, 
+          " lambda=", model$bestTune$lambda)
+  x.test <- model.matrix(y ~., testing)[,-1]
+  test.pred <- predict(model, x.test, type = 'prob')
+  y_pred <- apply(test.pred, 1, function(x) colnames(test.pred)[which.max(x)]) 
+  res <- cbind(data.frame(test.pred), 
+               data.frame('obs' = as.factor(testing$y), 
+                          'pred' = as.factor(y_pred)))
+  return(res)
+}
+
+
 # train a regression model to predict a target variable (y)
 # M: samples on the row, features on the column
 predict_numerical <- function(M, y, ...) {
@@ -1211,7 +1248,6 @@ plot_label_specific_vars <- function(M, labels, top_vars_per_label = 1,
                                 function(x) {head(x, top_vars_per_label)}))
     top_vars <- unique(dt$variable)
   }
-  #plot_tsne(M, M[,'LF36'])
   B <- get_basis_matrix(M[,top_vars,drop=F], labels)
   if(as_heatmap == TRUE) {
     p <- pheatmap::pheatmap(t(B), scale = 'row', silent = T, ...)
@@ -1219,7 +1255,6 @@ plot_label_specific_vars <- function(M, labels, top_vars_per_label = 1,
[…]

### 2021-05-31 src/manuscript_figures.Rmd
replace one-vs-all pancancer prediction with multi-class prediction - drop predict_labels function

@@
 ```{r}
-# Define a function that takes as input LFs and some patient labels (clinical/cancer type/subtype)
-# and evaluates predictability of the labels using the LFs
-predict_labels <- function(LFs, surv, tcga_subtypes, 
-                           project, datatype, label.type, ...) {
-  source(settings$utility_script)
-  predict_stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
-    require(data.table)
-    message("predicting ",label.type," using ",tool)
-    experiment <- paste(project, datatype, sep = '.')
-    M <- LFs[[tool]][[experiment]]
-    if(label.type == 'cancer_type') {
-      labels <- data.frame('cancer_type' = surv[match(rownames(M), bcr_patient_barcode)]$project, 
-                           row.names = rownames(M))
-    } else if (label.type == 'subtype') {
-      labels <- data.frame('subtype' = tcga_subtypes[match(rownames(M),
-                                                           bcr_patient_barcode)]$Subtype_Selected, 
-                           row.names = rownames(M))
-    } else if (label.type == 'clinical') {
-      labels <- process_clinical_covariates(rownames(M))
-    }
-    dt <- do.call(rbind, lapply(colnames(labels), function(x) {
-      l <- labels[,x]
-      if(is.numeric(l)){
-        # categorize samples by intervals
-        l <- as.character(cut_interval(l, 10))
-      }
-      l[is.na(l)] <- 'Undefined'
-      message("Predicting ",x, " for experiment ",experiment)
-      dt <- suppressMessages(predict_cluster_labels(M, l, ...))
-      dt$label.type <- x
-      dt$project <- project
-      dt$datatypes <- datatype
-      dt$method <- tool
-      return(dt)
-    }))
-    return(dt)
-  }))
-  return(predict_stats[!is.na(train_auc)])
-}
-
 plot_comp1 <- function(dt) {
@@
 # Find out if cancer types are predictable using latent factors
-pancancer_stats <- predict_labels(LFs, surv, tcga_subtypes, 'pancancer', 
-                                  'cnv_gex_meth_mut', label.type = 'cancer_type', num.trees = 1000)
-p1 <- plot_comp1(pancancer_stats) + theme(axis.title.x = element_blank())
-p2 <- plot_comp2(pancancer_stats)
+
+res <- sapply(simplify = F, names(LFs), function(tool) {
+  message(date(), " => ", tool)
+  M <- LFs[[tool]]$pancancer.cnv_gex_meth_mut
+  ct <- gsub("-", "_", surv[match(rownames(M), bcr_patient_barcode)]$project)
+  res <- predict_labels_glmnet_multiclass(M, ct, partition = 0.6, nodes = 20)
+})
+
+pancancer_stats <- sapply(res, function(x) {
+  caret::multiClassSummary(x, lev = levels(x$obs))
+})
+
+dt <- data.table(melt(pancancer_stats))[Var1 %in% c('Mean_Balanced_Accuracy')]
+dt$Var2 <- toupper(gsub("_factors", "", dt$Var2))
+p1 <- ggplot(dt, aes(x = Var2, y = value)) + 
+  geom_bar(stat = 'identity', aes(fill = Var2)) + 
+  labs(y = 'Mean Balanced Accuracy') + 
+  theme(legend.title = element_blank(), axis.title.x = element_blank())
+
+# performance visualisation
+m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
+p2 <- pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
+         display_numbers = T, number_format = "%.1f", silent = T)
+
 # plot cancer specific factors

### 2021-05-31 src/manuscript_figures.Rmd
drop pan-cancer unsupervised clustering experiment

@@
   theme(legend.position = 'right', legend.background = element_blank())
-
-# do unsupervised clustering experiment and compute AMI 
-# downsample pan-cancer dataset 
-df <- data.frame('id' = rownames(LFs$pca_factors$pancancer.cnv_gex_meth_mut))
-df$pr <- surv[match(df$id, bcr_patient_barcode)]$project
-ids <- lapply(split(df, df$pr), function(x) {
-  if(nrow(x) > 200) {
-    sample(x$id, 200)
-  }
-})
-ids <- ids[lengths(ids) == 200]
-k <- 5
-# sample cancer types and compute ami for different tools 
-sampled <- sapply(simplify = F, 1:50, function(i) {
-  unique(unlist(ids[sample(1:length(ids), k)]))
-})
-cl <- parallel::makeCluster(10)
-parallel::clusterExport(cl, varlist = c('ids', 'df', 'LFs', 'settings', 'sampled', 'k'))
-res <- do.call(rbind, pbapply::pblapply(cl = cl, 1:length(sampled), function(i) {
-  require(data.table)
-  source(settings$utility_script)
-  selected <- sampled[[i]]
-  ct <- df[match(selected, df$id),]$pr
-  message("selected cancers:",paste0(unique(ct), collapse = ','))
-  do.call(rbind, lapply(names(LFs), function(tool) {
-    M <- LFs[[tool]]$pancancer.cnv_gex_meth_mut[selected,]
-    if(tool == 'maui_factors') {
-      M <- scale(M)
-    }
-    ami <- compute_AMI(M, k, ct)
-    return(data.table('i' = i, 'ami' = ami, 'method' = tool, 
-                      'pr' = paste0(unique(ct), collapse = ',')))
-  }))
-}))
-parallel::stopCluster(cl)
-res$method <- toupper(gsub("_factors", "", res$method))
-
-comps <- lapply(setdiff(unique(res$method), 'MAUI'), function(x) c('MAUI', x))
-p5 <- ggboxplot(res, x = 'method', y = 'ami', add = 'jitter', color = 'method') + 
-  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
-                     comparisons = comps) +
-  labs(y = "Adjusted Mutual Information", x = '') +
-  theme(legend.position = 'none')
 

### 2021-06-01 src/manuscript_figures.Rmd
update pancancer benchmark figure panels

@@
   labs(y = 'Mean Balanced Accuracy') + 
-  theme(legend.title = element_blank(), axis.title.x = element_blank())
+  theme(legend.title = element_blank(), axis.title.x = element_blank(), 
+        axis.text = element_text(size = 14), axis.title = element_text(size = 16))
 
@@
 ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
+
 p3 <- plot_label_specific_vars(M, ct)
@@
 
-Fig <- plot_grid(plot_grid(p1, p2, p5, nrow = 1, labels = c('A', 'B', 'C')), 
-                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('D', 'E')), 
+Fig <- plot_grid(plot_grid(p1, p2$gtable, nrow = 1, labels = c('A', 'B')), 
+                  plot_grid(p3, p4, nrow = 1, labels = c('C', 'D')), 
                   nrow = 2,

### 2021-06-01 src/common_functions.R
decrease label size in tsne plot

@@ -591,7 +591,7 @@ plot_tsne <- function(M, factors = NULL, show.labels = F,
     centroids$group <- as.factor(rownames(centroids))
     p <- p + geom_text(data = centroids, aes(x = x, y = y, 
                                         label = group), 
-                       size = 8)
+                       size = 4)
   }
   return(p)
 }

### 2021-06-01 src/manuscript_figures.Rmd
drop one-vs-all pancancer subtype prediction benchmark

@@
 
-# predictability of subtypes using pancancer LFs
-pancancer_subtype_stats <- predict_labels(LFs, surv, tcga_subtypes, 'pancancer', 
-                                  'cnv_gex_meth_mut', label.type = 'subtype', num.trees = 1000)
-
-p1 <- plot_comp1(pancancer_subtype_stats) + theme(axis.title.x = element_blank())
-p2 <- plot_comp2(pancancer_subtype_stats)
-# plot subtype specific factors
-M <- LFs$maui_factors$pancancer.cnv_gex_meth_mut
-st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
-# remove NA values and pick subtypes with at least 20 samples 
-selected <- names(which(table(st) > 20))
-M <- M[st %in% selected,]
-labels <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
-
-p3 <- plot_label_specific_vars(M, labels)
-
-# data type contribution to top LF per factor
-p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
-                                           'pancancer', 'cnv_gex_meth_mut', factors = st, 
-                                           top_lfs = 1, top_features = 100) 
-p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
-  theme(legend.position = 'right', legend.background = element_blank())
-
-Fig <- plot_grid(plot_grid(plot_grid(p1, p2 + theme(axis.text.x = element_text(angle = 45)), 
-                                     nrow = 2, labels = c('A', 'B')), 
-                           p3 + theme(legend.position = 'none', 
-                                      axis.text.x = element_text(angle = 45)), 
-                           labels = c('', 'C'), ncol = 2, rel_widths = c(1,3)), 
-                 p4 + theme(axis.text.x = element_text(angle = 45)), 
-                 labels = c('', '', 'D'), ncol = 2, rel_widths = c(2.5,1))
-
-supFig_count <- supFig_count + 1
-SuppFigures[[supFig_count]] <- Fig
-fig.name <- paste0('SuppFigure', supFig_count, '.pancancer_subtype_prediction.pdf')
-names(SuppFigures)[supFig_count] <- fig.name
-ggsave(file.path(FigureFolder, fig.name), 
-              plot = Fig, units = 'in', width = 18, height = 12)
-SuppLegends[[supFig_count]] <- "Predictive power of latent factors in distinguishing known TCGA cancer subtypes A-B) comparing MAUI with PCA/MOFA/MCIA.
-AUC: Area under the ROC curve computed on the test dataset (30% of held-out data) trained on 70% of the dataset using a Random Forest model.
-Each point represents one cancer subtype (pairwise one-sided Wilcoxon test on AUCs, alternative = maui AUC is not greater than others) C) Top maui LFs detected per cancer subtype (only displaying subtypes with at least 20 samples).
-Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding subtype samples.
-D) Percentage of top feature types contributing to top predictive maui latent factor per cancer subtype."
 ```

### 2021-06-10 src/common_functions.R
add function to compute gene set scores using singscore

@@ -1483,6 +1483,11 @@ compute_geneset_enrichment <- function(genes, geneSets, background = NULL) {
   return(res)
 }
 
+# function to score a gene set
+score_gene_set <- function(rankData, genes_up, gene_down = NULL) {
+  scores <- singscore::simpleScore(rankData, upSet = genes_up)
+  return(scores$TotalScore)
+}
 
 
 

### 2021-06-10 src/manuscript_figures.Rmd
update pancancer type prediction figure panels

@@
 # Find out if cancer types are predictable using latent factors 
-
 res <- sapply(simplify = F, names(LFs), function(tool) {
@@
 # performance visualisation
-m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
-p2 <- pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
-         display_numbers = T, number_format = "%.1f", silent = T)
+#m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
+#p2 <- pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
+#         display_numbers = T, number_format = "%.1f", silent = T)
 
@@
 ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
+
+p2 <- plot_tsne(M, ct, show.labels = T) + 
+  theme(legend.position = 'none')
 
@@
 p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
-  theme(legend.position = 'right', legend.background = element_blank())
-
-Fig <- plot_grid(plot_grid(p1, p2$gtable, nrow = 1, labels = c('A', 'B')), 
-                  plot_grid(p3, p4, nrow = 1, labels = c('C', 'D')), 
+  theme(legend.position = 'top', legend.background = element_blank(), 
+        legend.title = element_blank())
+
+Fig <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B'),
+                           rel_widths = c(1,2)), 
+                  plot_grid(p4, p3, nrow = 1, labels = c('C', 'D'), 
+                            rel_widths = c(1,2)), 
                   nrow = 2,
@@
 ggsave(file.path(FigureFolder, fig.name),
-              plot = Fig, units = 'in', width = 12, height = 9)
+              plot = Fig, units = 'in', width = 12, height = 12)
 Legends[[fig_count]] <- "Predictive power of latent factors in distinguishing cancer types.
@@
 experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
-cl <- parallel::makeCluster(10)
+cl <- parallel::makeCluster(20)
 parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
@@
     if(tool == 'maui_factors') {
-      #M <- remove_redundant_variables(M, perc = 99) 
-      M <- scale(M)
+      # to avoid highly correlated maui LFs to dominate clusters
+      M <- scale(t(remove_redundant_variables(t(M))))
     } 
-
     # try topN from 5 to 25 (each tool may capture different amount of information 
@@
           x = 'method', y = 'score', add = 'jitter', color = 'method') + 
-  stat_compare_means(method = 'wilcox.test', paired = FALSE, #todo change to true when mofa is finished
+  stat_compare_means(method = 'wilcox.test', paired = TRUE, 
                      comparisons = comps) + 
@@
           x = 'method', y = 'score', add = 'jitter', color = 'method') + 
-  stat_compare_means(method = 'wilcox.test', paired = FALSE,  #todo change to true when mofa is finished
+  stat_compare_means(method = 'wilcox.test', paired = TRUE,  
                      comparisons = comps) + 

### 2021-06-10 src/manuscript_figures.Rmd
update MSI panel with benchmarks

@@
 # survival prediction accuracy benchmarking
-# find survival-predictive LFs by univariate cox models and use them in RSF
+# find survival-predictive LFs by fitting an initial RSF and use 
+# top factors for a final model to compute Harrel's c-index
 clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-cl <- parallel::makeCluster(10)
+cl <- parallel::makeCluster(20)
 parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
@@
 Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.cluster_stability_metrics.pdf')
+fig.name <- paste0('Figure', fig_count, '.cohort_based_metrics.pdf')
 names(Figures)[fig_count] <- fig.name
@@
 clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-event <- 'OS'
-time <- 'OS.time'
+event <- 'PFI'
+time <- 'PFI.time'
 
@@
 
-M <- LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']]
-M <- scale(M) #remove_redundant_variables(M, perc = 99)
+dt <- do.call(rbind, lapply(names(LFs), function(tool) {
+  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
   y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
@@
 
-res <- predict_cluster_labels_glmnet(M, y)
-p1 <- ggplot(res, aes(x = label, y = auc)) +
-  geom_bar(stat = 'identity', aes(fill = label)) + 
-  labs(y = 'AUC', x = '') +
-  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2)+
-  theme(legend.title = element_blank())
-
-# also plot a roc curve
-rocs <- predict_cluster_labels_glmnet(M, y, returnROC = TRUE)
-p2 <- plot_roc_list(rocs)
-
-# explore MSI related LFs
-# tsne plot of top LF related to MSI-H and MSI classes
+  # find LFs predictive of MSI 
   dt <- get_factor_specific_variables(M, y)
-# use top predictive lfs to guide tsne 
-top_vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 10)[['variable']])
-
+  dt$tool <- tool
+  return(dt)
+}))
+
+# compare top predictive factors across tools
+dt$tool <- toupper(gsub("_factors", "", dt$tool))
+p1 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
+       aes(x = ref_cl, y = -log10(padj))) + 
+  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
+  geom_hline(yintercept = -log10(0.05)) + 
+  coord_flip() + 
+  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
+        plot.title = element_text(hjust = 0.5))
+
+# focus on maui LFs to visualize MSI status-related LFs
+M <- LFs$maui_factors[['pan-gastrointestinal.cnv_gex_meth_mut']]
+y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
+M <- M[!is.na(y),]
+y <- y[!is.na(y)]
+top_vars <- unique(get_top_by_label(dt[tool=='MAUI'][padj < 0.05][order(padj)], 'ref_cl', 10)[['variable']])
 df <- plot_tsne(M[,top_vars], y, returnData = T)
-# color tsne by top lf informative for MSI-H
 top_lf <- dt[order(padj)][ref_cl == 'MSI-H']$variable[1]
 df <- merge(df, data.frame(M[,top_lf,drop =F]), by = 'row.names')
-df$cancertype <- surv[match(df$Row.names, bcr_patient_barcode)]$project
-df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected
-
-p3.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
+p2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
+    geom_point(aes(color = as.factor(group))) + 
     theme(legend.title = element_blank())
-p3.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
+p3 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
     geom_point(aes_string(color = top_lf)) + 
     scale_color_gradient(low = 'gray', high = 'red')
+
+p4 <- ggboxplot(df, x = 'group', y = top_lf, color = 'group', 
+          add = 'jitter') + theme(legend.position = 'none', 
+                                  axis.title.x = element_blank())
+
+# plot a roc curve
+rocs <- predict_cluster_labels_glmnet(M, y, returnROC = TRUE)
+p5 <- plot_roc_list(rocs)
 
@@
 }))
-p4 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
+p6 <- ggplot(freq[freq$ind < 5,], aes(x = reorder(rank, ind), y = Freq)) + 
   geom_bar(aes(fill = Var1), stat='identity') + 
@@
 
-Fig <- cowplot::plot_grid(plot_grid(p1, p2, ncol = 1, labels = c('A', 'B')),
-                          plot_grid(p3.1, p3.2, p4, ncol = 1, labels = c('C', 'D', 'E')),
-                          nrow = 1)
+Fig <- cowplot::plot_grid(p1, p5, p2, p3, p4, p6,
+                          nrow = 3)
 fig_count <- fig_count + 1

### 2021-06-10 src/manuscript_figures.Rmd
change pdf size - add labels

@@
 Fig <- cowplot::plot_grid(p1, p5, p2, p3, p4, p6,
-                          nrow = 3)
+                          nrow = 3, labels = 'AUTO')
 fig_count <- fig_count + 1
@@
 ggsave(filename = file.path(FigureFolder, fig.name), 
-       plot = Fig, units = 'in', width = 8, height = 8)
+       plot = Fig, units = 'in', width = 12, height = 12)
 Legends[[fig_count]] <- "Predicting MSI status of pan-gastrointestinal tumors using MAUI latent factors.

### 2021-06-10 src/manuscript_figures.Rmd
update immunothearpy response figure with benchmarks included

@@
 dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')]
-
-top_vars <- unique(dt$variable)
-
-# quantify prediction accuracy for finding complete/partial responders
+dt$tool <- toupper(gsub("_factors", "", dt$tool))
+dt$ref_cl <- ifelse(dt$ref_cl == 'CR/PR', 'Responders', 'Non-responders')
+
+p1 <- ggplot(dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')], 
+       aes(x = ref_cl, y = -log10(padj))) + 
+  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
+  geom_hline(yintercept = -log10(0.05)) + 
+  coord_flip() + 
+  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
+        plot.title = element_text(hjust = 0.5))
+
+# heatmap of top vars 
+# plot maui factors for responders/non-responders
+M <- LFs_imtrp$maui_factors$mariathasan.gex
+top_vars <- unique(dt[tool == 'MAUI'][padj < 0.01]$variable)
 m <- M[,top_vars]
-f <- colData$binaryResponse
+f <- ifelse(colData$binaryResponse == 'CR/PR', 'Responders', 'Non-responders')
 m <- m[!is.na(f),]
 f <- f[!is.na(f)]
-
-# heatmap of top vars 
-p1 <- pheatmap::pheatmap(get_basis_matrix(m[,top_vars], f), silent = T)
+p2 <- pheatmap::pheatmap(get_basis_matrix(m[,top_vars], f), silent = T)
 
@@
 require(survival)
-top_lf <- dt[order(padj)][ref_cl == 'CR/PR'][1,]$variable
+top_lf <- dt[order(padj)][ref_cl == 'Responders'][1,]$variable
 cp <- find_survival_cutpoint(m[,top_lf,drop = F], colData, event_col = 'censOS', duration_col = 'os')
-
-p2 <- plot_survival(surv = colData, partition.df = data.frame('group' = ifelse(m[,top_lf] > cp, 'high', 'low'), 
+p3 <- plot_survival(surv = colData, 
+                    partition.df = data.frame('group' = ifelse(m[,top_lf] > cp, 
+                                              paste0(top_lf, ">", round(cp, 1)),
+                                              paste0(top_lf, "<=", round(cp, 1))), 
                                               row.names = rownames(m)), 
@@
 # top go terms for top features of top LF 
+W <- FWs_imtrp$maui_factors$mariathasan.gex
 feats <- data.table(W[,top_lf,drop=F], keep.rownames = T)[order(abs(get(top_lf)), decreasing = T)][1:100]
 # top pathways for top lf
-top_terms <- compute_geneset_enrichment(feats$rn, pathways)[order(pval)][padj < 0.05][1:10]
-top_terms$term_wr <- stringr::str_wrap(gsub("_", " ", top_terms$term), width = 32)
-p3 <- ggplot(top_terms, aes(x = reorder(term_wr, -pval), y = -log10(padj))) + 
+top_terms <- compute_geneset_enrichment(feats$rn, pathways)[order(pval)][padj < 0.05][1:5]
+top_terms$term_wr <- stringr::str_wrap(gsub("_", " ", top_terms$term), width = 24)
+p4 <- ggplot(top_terms, aes(x = reorder(term_wr, -pval), y = -log10(padj))) + 
   geom_bar(stat = 'identity') + 
@@
 colnames(feats)[2] <- 'top_lf'
-p4 <- ggplot(feats[1:10,], aes(x = reorder(rn, abs(top_lf)), y = abs(top_lf))) + 
+p5 <- ggplot(feats[1:10,], aes(x = reorder(rn, abs(top_lf)), y = abs(top_lf))) + 
   geom_bar(stat = 'identity') + coord_flip() + labs(x = '', y = 'Feature Importance (absolute)')
@@
 # get samples with both response and tmb annotation
-cd <- colData #colData[!is.na(`Neoantigen burden per MB`)][!is.na(binaryResponse)]
-m2 <- M[cd$bcr_patient_barcode,]
-require(ggpubr)
-df <- data.frame('response' = cd$binaryResponse, 
-                 'TMB' = cd$`Neoantigen burden per MB`, 
-                 'LF' = m2[,top_lf])
-p5 <- ggpubr::ggscatter(df, x = 'LF', y = 'TMB', 
-                        add = 'reg.line', cor.coef = T, cor.coef.coord = c(30, 10), 
-                  cor.coef.size = 4) + labs(x = top_lf, y = 'Tumor Mutation Burden')
-
+dt <- do.call(rbind, lapply(names(LFs_imtrp), function(tool) {
+  M <- LFs_imtrp[[tool]]$mariathasan.gex
+  dt <- data.table(cor(M, colData$`Neoantigen burden per MB`, use = 'complete.obs'),
+                   keep.rownames = T)
+  dt$tool <- tool
+  return(dt)
+}))
+dt <- dt[,.SD[which.max(abs(V1))], by = c('tool')]
+dt$tool <- toupper(gsub("_factors", "", dt$tool))
+p6 <- ggplot(dt, aes(x = tool, y = abs(V1))) + 
+  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
+  coord_flip() + 
+  labs(y = 'Correlation with Tumor Mutation Burden') + 
+  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
+        plot.title = element_text(hjust = 0.5)) 
 # immune phenotypes 
-ip <- colData$`Immune phenotype`
-#ip[ip == 'excluded'] <- NA
-m3 <- M[!is.na(ip),]
-ip <- ip[!is.na(ip)]
-
-dt3 <- get_factor_specific_variables(m3, ip)[order(pval)][padj < 0.01]
-ip_vars <- unique(dt3$variable)
-ip_vars <- get_top_by_label(dt3, 'ref_cl', topN = 5)
-res_ip <- predict_cluster_labels_glmnet(m3[,unique(dt3$variable)], ip, returnROC = T)
-p6 <- plot_roc_list(res_ip)
-
-pheatmap::pheatmap(get_basis_matrix(m3[,unique(ip_vars$variable)], ip), silent = T)
-
-Fig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2$gtable, p4, p3$plot, 
-                                             labels = 'AUTO'), 
-                          cowplot::plot_grid(p5, p6, p7, nrow = 1, 
-                                             labels = c('E', 'F', 'G')), 
-                          nrow = 2, rel_heights = c(2, 1)) 
+dt <- do.call(rbind, lapply(names(LFs_imtrp), function(tool) {
+  M <- LFs_imtrp[[tool]]$mariathasan.gex
+  # find LFs predictive of immune phenotypes 
+  dt <- get_factor_specific_variables(M, colData$`Immune phenotype`)[order(padj)]
+  dt$tool <- tool
+  return(dt)
+}))
+
+dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')]
+dt$tool <- toupper(gsub("_factors", "", dt$tool))
+p7 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
+             aes(x = ref_cl, y = -log10(padj))) + 
+  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
+  geom_hline(yintercept = -log10(0.05)) + 
+  labs(x = 'Tumor Immune Phenotypes') + 
+  coord_flip() + 
+  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
+        plot.title = element_text(hjust = 0.5))
+
+Fig <- cowplot::plot_grid(
+  cowplot::plot_grid(p1, p6, p7, nrow = 1, labels = 'AUTO'),
+  cowplot::plot_grid(p2$gtable, p3$plot, p4, p5, nrow = 2, labels = c('D', 'E', 'F', 'G')),
+  nrow = 2, 
+  rel_heights = c(1,2)
+  ) 
 fig_count <- fig_count + 1
@@
        plot = Fig, units = 'in', 
-       width = 10, height = 14)
+       width = 12, height = 14)
 Legends[[fig_count]] <- "Predicting responders to anti-PD-L1 treatment in a cohort of Metastatic Urothelial

### 2021-08-24 src/manuscript_figures.Rmd
update figure legends

@@
 settings <- yaml::read_yaml(params$settings_file)
+message(settings$GDCdataDir)
 FigureFolder <- params$figures_folder
@@
 library(survival)
+library(caret)
 theme_set(ggpubr::theme_pubclean())
@@
 Legends[[fig_count]] <- "Predictive power of latent factors in distinguishing cancer types.
-A-B) comparing maui with PCA/MOFA/MCIA.
-AUC: Area under the ROC curve computed on the test dataset (30% of held-out data) trained on 70% of the dataset using a Random Forest model.
-Each point represents one cancer type (pairwise one-sided Wilcoxon test on AUCs, alternative = maui AUC is not greater than others) C) Adjusted Mutual Information (AMI) value computed by comparing unsupervised clustering labels (k-means algorithm, k set to 21) with the cancer types.
+A) comparing MAUI with PCA/MOFA/MCIA.
+Mean Balanced Accuracy computed on the test dataset (30% of held-out data) trained on 70% of the dataset using an Elastic Net multi-class model.
+B) tSNE plot of MAUI latent factors learned from pan-cancer dataset  C) Percentage of top feature types contributing to top predictive maui latent factor per cancer type.
 D) Top maui LFs detected per cancer type.
-Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples.
-E) Percentage of top feature types contributing to top predictive maui latent factor per cancer type."
+Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples."
 
@@
 Legends[[fig_count]] <- "Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in A) and cluster stability (Average Proportion of Non-Overlap (APN) in B).
-Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) based on the top 10 LFs/PCs derived from each tool.
-C) Same as in panels A and B), but split by the value of k used in clustering.
-D) Survival outcome prediction accuracy for 21 cancer types by each tool's top 10 survival-related factors based on Overall Survival or Progression Free Survival endpoints."
+Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) for each tool.
+C-D) Survival outcome prediction accuracy for 21 cancer types by each tool's top 10 survival-related factors based on Overall Survival (in C) or Progression Free Survival endpoints.
+(in D)."
 ```
@@
        plot = Fig, width = 12, height = 14, units = 'in')
-Legends[[fig_count]] <- "Using MAUI LFs for survival prediction.
+Legends[[fig_count]] <- "Using MAUI LFs for progression-free survival prediction.
 A) Harrell’s C-index computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs.
 B) Hazard ratios of top LFs picked for each cancer adjusted for clinical factors (tumor stage + age + gender).
-Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
+C) Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
 ```
@@
 Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type.
-B) Example case: top LFs associated with each histological subtype of TCGA-SARC.
-C) Example case: top LFs associated with each histological grade of TCGA-HNSC."
+B) Example case: top LFs associated with each gender groups of TCGA-THCA cohort.
+C) Example case: top LFs associated with each histological types of TCGA-UCEC."
 ```
@@
        plot = Fig, units = 'in', width = 12, height = 12)
-Legends[[fig_count]] <- "Predicting MSI status of pan-gastrointestinal tumors using MAUI latent factors.
-A) AUC values for predicting MSI status using MAUI LFs B) Top features that contribute to MAUI latent factor (‘LF95’) that is the most predictive latent factor in distinguishing samples with high microsatellite instability from the rest.
-C) tSNE plots of samples generated using top MSI-predictive LFs (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI latent factor (‘LF95’) which is most predictive of MSI-H samples."
+Legends[[fig_count]] <- "Latent factors predictive of MSI status (high: MSI-H, stable: MSI-S, low: MSI-L) in pan-gastrointestinal cancers A) Comparison of top latent factor per MSI status for each tool in terms of adjusted p-values reflecting the differences between MSI subgroups.
+B) ROC curves for prediction accuracy of MSI status using MAUI latent factors C) tSNE plots of samples generated using top MSI-predictive LFs (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI latent factor LF25, which is most predictive of MSI-H samples.
+E) Distribution of MAUI latent factor LF25 across MSI subgroups in the pan-gastrointestinal cancer samples.
+F) Relative abundance of omics feature types among the top 500 features contributing to MAUI latent factor LF25."
 ```
@@
 Legends[[fig_count]] <- "Classification of Non-Small-Cell Lung Adenocarcinoma samples using MAUI latent factors.
-A) ROC curves with AUC values on testing data (40% of the dataset) trained on Elastic Net with 5-fold cross-validation.
+A) ROC curves with AUC values on testing data (40% of the dataset) using an Elastic Net model with 5-fold cross-validation.
 B) t-SNE plot of MAUI LFs colored by cancer types.
@@
        width = 10, height = 12)
-Legends[[fig_count]] <- "Clustering of pan-gastrointestinal tumors guided by TCGA subtypes.
-Subtype specific MAUI LFs are used to cluster the samples, which yields better survival curves as shown in C,D).
-Clusters are annotated using pathway activity scores (top cluster associated cancer hallmarks)."
+Legends[[fig_count]] <- "Clustering of pan-gastrointestinal tumors guided by TCGA subtypes using subtype-specific MAUI latent factors.
+A) tSNE plot of using MAUI latent factors for pan-gastrointestinal cancers colored by MAUI clusters, B) tSNE plot of using MAUI latent factors for pan-gastrointestinal cancers colored by known TCGA subtypes C) progression-free survival curves for MAUI-based clusters and D) known TCGA subtypes.
+E-F) Gene-set activity scores of annotated cancer hallmarks gene sets enriched for each MAUI-based cluster in E) and known TCGA subtypes in F)."
 ```
@@
        width = 10, height = 12)
-Legends[[fig_count]] <- "Log-rank test of prognostic subtypes by MAUI and existing TCGA subtypes.
-B) tSNE plot of MAUI LFs that are predictive of survival outcomes for TCGA-UCEC.
+Legends[[fig_count]] <- "A) Log-rank test of prognostic subtypes by MAUI and annotated TCGA subtypes.
+B) tSNE plot of MAUI LFs that are predictive of survival outcomes for TCGA-PRAD.
 C) Kaplan-Meier curve of the best/worst surviving prognostic MAUI clusters D) Biological pathway scores that distinguish best/worst surviving clusters."
@@
 Legends[[fig_count]] <- "Predicting responders to anti-PD-L1 treatment in a cohort of Metastatic Urothelial
-Cancers (N=348) using MAUI. A) AUC curve for predicting Complete/Partial Responders (CR/PR) to the anti-PD-L1
-treatment using an Elastic Net Model on a subset of MAUI latent factors individually predictive of treatment response groups.
-B) Average activation score of top latent factors predictive of treatment response groups.
-C) Top pathways enriched among top 100 genes that contribute to the top latent factor (LF44).
-D) Survival stratification of the
-cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4). E) Top 10 genes contributing to LF44
-F) Correlation between LF44 activation score and the tumor (neo-antigen) mutation burden"
+Cancers (N=348) using MAUI.
+A) Comparison of top latent factors predictive of complete/partial response (CR/PR) to the anti-PD-L1.
+B) Comparison of top latent factors with respect to their correlation with tumor mutation burden status of the samples in the cohort C) Comparison of top latent factors predictive of immune phenotypes (inflamed/excluded/desert).
+D) Top MAUI LFs predictive of response to anti-PD-L1 response groups.
+E) Survival stratification of the cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4).
+F) Top pathways enriched among top 100 genes that contribute to the top latent factor (LF44).
+G) Top 10 genes contributing to LF44"
 ```

### 2021-08-25 src/manuscript_figures.Rmd
drop Fig7, semi-supervised pan-gI analysis

@@
 
-```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
-# Semi-supervised clustering of samples 
-# 1. Cluster samples by LFs associated to known TCGA subtypes or those that are associated to survival outcomes. 
-#2. Annotate each cluster using cancer state scores or hallmark gene set scores
-
-## TCGA subtype informed clusters 
-## Use-case: pan-gastrointestinal cancers
-require(survminer)
-require(cluster)
-pr <- 'pan-gastrointestinal' 
-experiment <- paste0(pr, '.cnv_gex_meth_mut')
-M <- LFs$maui_factors[[experiment]]
-M <- scale(M) #remove_redundant_variables(, perc = 99)
-st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
-# find subtype specific LFs 
-dt <- get_factor_specific_variables(M, st)
-
-vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 3)[['variable']])
-
-k <- length(na.omit(unique(st))) # set k to number of known subtypes
-# cluster samples using only subtype specific factors 
-clust <- data.frame(get_kmeans_subtypes(M[,vars], k_vals = k, min_cluster_size = 0))
-
-# plots
-df <- plot_tsne(M[,vars], as.factor(clust[,1]), returnData = T)
-colnames(df)[3] <- 'maui'
-df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
-df$cancertype <- surv[match(rownames(df), bcr_patient_barcode)]$project
-
-p <- ggplot(df, aes(x = tSNE1, y = tSNE2))  
-p1 <- p + geom_point(aes(color = maui)) +
-  geom_text_repel(data = get_centroid(df[,1:2], df$maui), 
-            aes(x,y,label = text.label), size = 6)
-p2 <- p + geom_point(aes(color = tcga)) + 
-        geom_text_repel(data = get_centroid(df[,1:2], df$tcga), 
-            aes(x,y,label = text.label), size = 5)
-p3 <- plot_survival(surv, df[,'maui',drop =F])[['plot']] + theme(legend.position = 'none') + labs(y = 'Progression Free Survival')
-p4 <- plot_survival(surv, df[,'tcga',drop =F])[['plot']] + theme(legend.position = 'none') + labs(y = 'Progression Free Survival')
-
-# use hallmark gene sets to annotate cluster specific functions
-scores <- import_geneset_scores(geneSetScoresFolder, 'msigdb_hallmarks')
-
-S <- scores[intersect(rownames(M), rownames(scores)),]
-colnames(S) <- gsub("HALLMARK_", "", colnames(S))
-clust$tcga <- tcga_subtypes[match(rownames(clust), bcr_patient_barcode)]$Subtype_Selected
-
-# plot top gene sets
-h1 <- plot_label_specific_vars(S, clust[rownames(S), 1], 
-                              as_heatmap = T, cellwidth = 10, 
-                              cellheight = 12,
-                              top_vars_per_label = 2)
-h2 <- plot_label_specific_vars(S, clust[rownames(S), 2], 
-                              as_heatmap = T, cellwidth = 10, 
-                              cellheight = 12, 
-                              top_vars_per_label = 2)
-
-# altogether
-Fig <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
-                         labels = 'AUTO', ncol = 2)
-
-fig_count <- fig_count + 1
-Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.semi_supervised.clustering.', experiment,'.pdf')
-names(Figures)[fig_count] <- fig.name
-ggsave(filename = file.path(FigureFolder, fig.name), 
-       Fig, units = 'in', 
-       width = 10, height = 12)
-Legends[[fig_count]] <- "Clustering of pan-gastrointestinal tumors guided by TCGA subtypes using subtype-specific MAUI latent factors.
-A) tSNE plot of using MAUI latent factors for pan-gastrointestinal cancers colored by MAUI clusters, B) tSNE plot of using MAUI latent factors for pan-gastrointestinal cancers colored by known TCGA subtypes C) progression-free survival curves for MAUI-based clusters and D) known TCGA subtypes.
-E-F) Gene-set activity scores of annotated cancer hallmarks gene sets enriched for each MAUI-based cluster in E) and known TCGA subtypes in F)."
-```
-
 ```{r semi_supervised_survival, fig.width=10, fig.height=12}

### 2021-08-25 src/manuscript_figures.Rmd
Drop Fig 8, semi-supervised stratification of patients based on survival

@@
 
-```{r semi_supervised_survival, fig.width=10, fig.height=12}
-
-## Clustering informed by survival outcomes
-## Pick survival-predictive LFs and cluster samples based on those
-
-# factors to use in the model
-clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-
-experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
-cl <- parallel::makeCluster(10)
-parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 
-                                        'experiments', 'settings'))
-prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
-                                function(experiment){
-    require(data.table)
-    require(survminer)
-    source(settings$utility_script)
-    message(experiment)
-    M <- LFs$maui_factors[[experiment]]
-    # get clinical factors for the patients in matrix
-    clin <- process_clinical_covariates(rownames(M))
-    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
-
-    fit <- fit_RSF(surv, M, duration_col = 'PFI.time', event_col = 'PFI')
-    prog_lfs <- names(head(sort(fit$importance, decreasing = T), 5))
-    # now cluster samples based on only survival predictive factors
-    clust <-  get_kmeans_subtypes(M[,prog_lfs], k_vals = 3:8, min_cluster_size = 0)
-    # find value of k which minimizes log-rank p-value
-    dt <- do.call(rbind, lapply(colnames(clust), function(x) {
-      message(x)
-      pval <- plot_survival(surv, clust[,x,drop=F], return_pval = T)
-      data.table('k' = x, 'pval' = pval)
-    }))
-
-    # return clustering with best result
-    top <- dt[order(pval)][1]$k
-    return(list('prog.lf' = M[,prog_lfs], 'prog.clust' = clust[,top,drop = F]))
-})
-parallel::stopCluster(cl)
-
-# compare maui prognostic clusters with TCGA labels 
-stats <- do.call(rbind, lapply(names(prog.clust), function(experiment) {
-  message(experiment)
-  if(is.null(prog.clust[[experiment]])) {return(NULL)}
-  df <- data.frame(prog.clust[[experiment]][['prog.clust']])
-  colnames(df)[1] <- 'maui'
-  df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected
-  do.call(rbind, lapply(colnames(df), function(x) {
-    if(length(unique(df[,x])) < 2) {
-      pval <- NA
-    } else {
-      pval <- plot_survival(surv, df[,x,drop = F], return_pval = T)
-    }
-    return(data.table('source' = x, 'pval' = pval, 'experiment' = experiment))
-  }))
-}))
-
-stats$cancertype <- gsub('.cnv_gex_meth_mut', '', stats$experiment)
-# which cancers can we improve upon tcga?
-p1 <- ggplot(stats, aes(x = cancertype, y = -log10(pval))) +  
-  geom_bar(stat = 'identity', aes(fill = source), position = 'dodge') +
-  geom_hline(yintercept = -log10(0.05)) + 
-  coord_flip() + 
-  scale_fill_brewer(type = 'qual', palette = 6)
-
-# use canonical pathways to annotate prognostic groups
-scores <- import_geneset_scores(geneSetScoresFolder, 'msigdb_canonical_pathways')
-
-# scores: geneset scores
-# M: survival predictive LF matrix
-# maui_clust: best survival prognostic clustering of M 
-plot_clusts <- function(surv, M, maui_clust, scores) {
-  df <- plot_tsne(M, as.factor(maui_clust[,1]), returnData = T)
-  colnames(df)[3] <- 'maui'
-
-  message("running plot clusts")
-  message("p1")
-  p1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
-    geom_point(aes(color = maui)) +
-    geom_text_repel(data = get_centroid(df[,1:2], df$maui), 
-            aes(x,y,label = text.label), size = 6)
-  # plot survival for the worst prognostic group, and the rest
-  ms <- get_median_survival_by_group(surv, df[,'maui',drop=F])
-
-  # find clusters with worst/best prognosis (by median survival)
-  worst <- names(sort(ms)[1])
-  best <- names(sort(ms, T)[1])
-
-  message("p2")
-  p2 <- plot_survival(surv, 
-                      df[df$maui %in% c(best, worst), 
-                         'maui', drop = F])[['plot']] +
-    labs(y = 'PFI') +
-    theme(legend.position = 'left', legend.title = element_blank()) + 
-    scale_color_brewer(type = 'qual', palette = 6)
-
-  # heatmap of hallmark scores for best/worst survivors
-  selected <- intersect(rownames(df[df$maui %in% c(worst, best),]), rownames(S))
-  dt <- get_factor_specific_variables(S[selected,], df[selected, 'maui'])
-  # select factors 
-  dt[padj < 0.05][order(padj)]
-  message("h1")
-
-  h1 <- plot_label_specific_vars(S[selected,], 
-                                 df[selected, 'maui'], 
-                                 padj_threshold = 0.1,
-                                as_heatmap = T, cellwidth = 15, 
-                                top_vars_per_label = 5, cellheight = 15)
-  return(list('p1' = p1, 'p2' = p2, 'h1' = h1))
-}
-
-# plot examples for which we improve upon tcga
-dt <- data.table(dcast(stats, experiment ~ source, value.var = 'pval'))
-dt$improvement <- -log10(dt$maui) - -log10(dt$tcga)
-dt <- dt[order(improvement, decreasing = T)]
-
-e <- dt[1,]$experiment
-M <- prog.clust[[e]]$prog.lf
-maui_clust <- prog.clust[[e]]$prog.clust
-S <- scores[intersect(rownames(M), rownames(scores)),]
-P <- plot_clusts(surv, M, maui_clust, S)
-
-Fig <- cowplot::plot_grid(p1, plot_grid(P$p1, P$p2, nrow = 1, labels = c('B', 'C')), 
-                   P$h1$gtable, nrow = 3, labels = c('A', '', 'D'))
-
-message("Got the figure")
-
-fig_count <- fig_count + 1
-Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.pdf')
-names(Figures)[fig_count] <- fig.name
-ggsave(filename = file.path(FigureFolder, fig.name), 
-       plot = Fig, units = 'in', 
-       width = 10, height = 12)
-Legends[[fig_count]] <- "A) Log-rank test of prognostic subtypes by MAUI and annotated TCGA subtypes.
-B) tSNE plot of MAUI LFs that are predictive of survival outcomes for TCGA-PRAD.
-C) Kaplan-Meier curve of the best/worst surviving prognostic MAUI clusters D) Biological pathway scores that distinguish best/worst surviving clusters."
-```
-
 ```{r}

### 2021-08-25 src/manuscript_figures.Rmd
drop survival benchmark from Fig 2 - c-index comparison of tools

@@
   labs(title = 'Cluster Stability Score (APN)', x = '') + 
-  theme(legend.position = 'none')
-
-# survival prediction accuracy benchmarking
-# find survival-predictive LFs by fitting an initial RSF and use 
-# top factors for a final model to compute Harrel's c-index
-clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-cl <- parallel::makeCluster(20)
-parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
-surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
-  message("survival benchmark for ",tool)
-  do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
-    message(experiment)
-    source(settings$utility_script)
-    require(data.table)
-    require(survminer)
-    M <- LFs[[tool]][[experiment]]
-    clin <- process_clinical_covariates(rownames(M))
-    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
-    # compute pfi stats
-    # fit a first model to filter variables 
-    fit <- fit_RSF(surv, M, ntree = 500, 'PFI.time', 'PFI')
-    # using a range of top variables, compute c index
-    dt1 <- do.call(rbind, lapply(2:10, function(i) {
-      vars <- names(head(sort(fit$importance, decreasing = T), i))
-      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'PFI', 'PFI.time'), 
-                                M[,vars], ntree = 500)
-      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
-                       'experiment' = experiment, 'method' = tool, 'topN' = i)
-    }))
-    dt1$endpoint <- 'PFI'
-    # do the same for overall survival
-    fit <- fit_RSF(surv, M, ntree = 500, 'OS.time', 'OS')
-    dt2 <- do.call(rbind, lapply(2:10, function(i) {
-      vars <- names(head(sort(fit$importance, decreasing = T), i))
-      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'OS', 'OS.time'), 
-                                M[,vars], ntree = 500)
-      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
-                       'experiment' = experiment, 'method' = tool, 'topN' = i)
-    }))
-    dt2$endpoint <- 'OS'
-    return(rbind(dt1, dt2))
-  }))
-}))
-parallel::stopCluster(cl)
-
-dt <- surv.stats
-dt$method <- toupper(gsub("_factors", "", dt$method))
-dt$project <- unlist(sapply(strsplit(dt$experiment, "\\."), function(x) x[1]))
-dt$endpoint <- ifelse(dt$endpoint == 'PFI', 'Progression Free Survival', 
-                      'Overall Survival')
-# get best performance per tool per project
-dt <- dt[,.SD[which.max(cindex)], by = c('project', 'method', 'endpoint')]
-# PCA/MOFA win, compare against PCA
-comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
-p4 <- ggboxplot(dt, x = 'method', y = 'cindex', color = 'method',
-                add = 'jitter', facet.by = 'endpoint') +
-  stat_compare_means(method = 'wilcox.test', paired = TRUE,
-                     comparisons = comps) +
-  labs(y = "Harrel's C-index", x = '') +
   theme(legend.position = 'none')

### 2021-08-25 src/manuscript_figures.Rmd
drop Fig 3B - hazard ratio barplots with confidence intervals

@@
 
-  # compute hazard ratios for the top factor
+  # compute adjusted survival curve for thb top factor
   v <- vars[1] 
@@
   dat <- get_surv_df(surv, dat, event, time)
-  # compute hazard ratios 
-  cox1 <- coxph(Surv(time,status) ~ ., 
-          data=dat, x = TRUE)
-  hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
-  hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
-  colnames(hr) <- c('hr', 'hr_low', 'hr_high')
-  rownames(hr) <- v
-  p1 <- survminer::ggforest(cox1, data = dat)
 
@@
                 data=dat, x = TRUE)
-  p2 <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')
+  p <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')
 
@@
 
-  return(list('p1' = p1, 'p2' = p2, 'hr' = hr, 
-              'Cindex_base' = rsf_stats$C_wo, 'Cindex_with_LFs' = rsf_stats$C_with))
-
+  return(list('p' = p, 
+              'Cindex_base' = rsf_stats$C_wo, 
+              'Cindex_with_LFs' = rsf_stats$C_with))
   return(NULL)
@@
 
+# combine adjusted survival curve plots into a single plot
 plots <- sapply(simplify = F, names(surv.stats), function(x) {   
-  p <- surv.stats[[x]][['p2']] 
+  p <- surv.stats[[x]][['p']] 
   if(!is.null(p)) { 
@@
                    label_size = 8)
-# plot hazard ratios for top LF per cancer type 
-hr <- do.call(rbind, lapply(names(surv.stats), function(x) {
-  if(is.null(surv.stats[[x]])) {return(NULL)}
-  message(x)
-  df <- surv.stats[[x]][['hr']]
-  df$project <- unlist(strsplit(x, '\\.'))[[1]]
-  return(df)
-}))
-hr$latent_factor <- rownames(hr)
-# remove hr with Inf values
-hr <- hr[!is.infinite(hr$hr_high),]
-
-p2 <- ggplot(hr, aes(x = reorder(paste0(project, ' (', latent_factor, ')'), hr), y = log(hr))) +
-  geom_bar(stat = 'identity') + 
-  geom_errorbar(aes(ymin = log(hr_low), ymax = log(hr_high)), width = 0.2) + 
-  labs(x = '', y = 'log(hazard-ratio) of top latent factor per cancer\n(adjusted for tumor stage + age + gender)')+
-  coord_flip()
 

### 2021-08-25 src/manuscript_figures.Rmd
Merge fig3 with fig4, survival-related LF analysis is now part of clinical factor analysis Drop Fig4B/C (example LFs associated to clinical factors), move 4A (clinical factor-associated LFs) to merge with Fig3A (cindex improvement using LFs), drop Fig3B (hazard ratio plot), move Fig3C (survival adjusted curves) to supplements

@@
 
-```{r survival_factors, fig.width=12, fig.height=8}
-# survival-predictive factors
-#I look for survival-predictive maui LFs, show how much c-index improves on top of 
-#clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.
+```{r clinical_survival_factors, fig.width=12, fig.height=8}
+# here we compute LFs associated to clinical factors (e.g. tumor stage, gender etc) and 
+# survival outcomes (PFI)
 
-# factors to use in the model
-clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-event <- 'PFI'
-time <- 'PFI.time'
-
-experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
-cl <- parallel::makeCluster(20)
-parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'event', 'time',
-                                        'surv', 'experiments', 'settings'))
-surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
-  require(data.table)
-  require(survminer)
-  require(survival)
-  source(settings$utility_script)
-  message(experiment)
-  M <- LFs$maui_factors[[experiment]]
-  # get clinical factors for the patients in matrix
-  clin <- process_clinical_covariates(rownames(M))
-  clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
-  # first filter variables by importance 
-  fit <- fit_RSF(surv, M, 500, time, event)
-  vars <- names(head(sort(fit$importance, decreasing = T), 5))
-
-  # compute adjusted survival curve for thb top factor
-  v <- vars[1] 
-  dat <- merge(clin, M[,v,drop = F], by = 'row.names')
-  rownames(dat) <- dat$Row.names
-  dat$Row.names <- NULL
-  dat <- get_surv_df(surv, dat, event, time)
-
-  # split variable into high/low to compute survival adjusted curves
-  cp <- find_survival_cutpoint(dat[,v,drop=F], surv, event_col = event, duration_col = time)
-  dat[,v] <- ifelse(dat[,v] > cp, paste(v, 'high'), paste(v, 'low'))
-  cox2 <- coxph(Surv(time,status) ~ ., 
-                data=dat, x = TRUE)
-  p <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')
-
-  # compute Cindex with all survival-predictive LFs and without LFs 
-  rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, event, time), 
-                            M[,vars], ntree = 1000)
-
-  return(list('p' = p, 
-              'Cindex_base' = rsf_stats$C_wo, 
-              'Cindex_with_LFs' = rsf_stats$C_with))
-  return(NULL)
-})
-parallel::stopCluster(cl)
-
-# combine adjusted survival curve plots into a single plot
-plots <- sapply(simplify = F, names(surv.stats), function(x) {   
-  p <- surv.stats[[x]][['p']] 
-  if(!is.null(p)) { 
-    p + theme(axis.text.x = element_text(size = 8), 
-              axis.text.y = element_text(size = 8), 
-              axis.title.x = element_blank(), 
-              axis.title.y = element_blank(), 
-              legend.title = element_blank(),
-              legend.position = c(0.5, 0.9),
-              legend.text = element_text(size = 8), 
-              legend.background = element_blank(), 
-              legend.direction = 'vertical') 
-    }
-})
-plots <- plots[!sapply(plots, is.null)]
-p1 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
-                   labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
-                   label_size = 8)
-
-# make a plot to show improvement in C index with LFs on top of clinical factors
-ci <- do.call(rbind, lapply(names(surv.stats), function(x) {
-  if(is.null(surv.stats[[x]])) {return(NULL)}
-  df <- data.frame('base' = surv.stats[[x]][['Cindex_base']], 
-                   'with_lf' = surv.stats[[x]][['Cindex_with_LFs']], 
-                   'project' = unlist(strsplit(x, '\\.'))[[1]])
-  return(df)
-}))
-
-dt <- melt(ci, id.vars = 'project')
-dt$variable <- ifelse(dt$variable == 'base', 'Only clinical factors', 'Clinical factors + MAUI LFs')
-
-p3 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
-  geom_segment(aes(x = reorder(project, with_lf), xend = project, 
-                   y = base, yend = with_lf)) + 
-  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 2) +
-  coord_flip() + 
-  labs(x = '', y = '') + 
-  theme(legend.title = element_blank(), legend.position = 'bottom', 
-        legend.background = element_blank(), 
-        legend.margin = margin(), 
-        legend.direction = 'vertical', legend.box.background = element_blank()) +
-  scale_color_brewer(type = 'qual', palette = 6)
-
-Fig <- cowplot::plot_grid(cowplot::plot_grid(p3, p2, nrow = 1, labels = c('A', 'B')),
-                        p1, labels = c('', 'C'), nrow = 2)
-
-fig_count <- fig_count + 1
-Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
-names(Figures)[fig_count] <- fig.name
-ggsave(filename = file.path(FigureFolder, fig.name), 
-       plot = Fig, width = 12, height = 14, units = 'in')
-Legends[[fig_count]] <- "Using MAUI LFs for progression-free survival prediction.
-A) Harrell’s C-index computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs.
-B) Hazard ratios of top LFs picked for each cancer adjusted for clinical factors (tumor stage + age + gender).
-C) Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
-```
-
-```{r clinical_factors, fig.width=12, fig.height=9}
-# clinical factor associated MAUI LFs
+## 1. clinical factor associated MAUI LFs
 # Find maui LFs associated to clinical factors and plot them
@@
 
-# pick example cases and show more detailed LF activation 
-# top examples by clinical factor type
-# top 5 variables per clinical factor 
-top <- clin.stats[padj < 0.05,.SD[which.min(padj)], by = label][order(padj),.SD[which.min(padj)],by = project]
-plots <- sapply(simplify = FALSE, 1:4, function(i) {
-  pr <- top$project[i]
-  l <- top$label[i]
-  M <- LFs$maui_factors[[paste0(pr, '.cnv_gex_meth_mut')]]
-  factors <- process_clinical_covariates(rownames(M))[[l]]
-  M <- M[!is.na(factors),]
-  factors <- factors[!is.na(factors)]
-  plot_label_specific_vars(M, factors, string_width = 24)+ 
-    labs(title = pr, subtitle = l) + 
-    theme(legend.position = 'none')
+# 2. survival-predictive factors
+#I look for survival-predictive maui LFs, show how much c-index improves on top of 
+#clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.
+
+# factors to use in the model
+clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
+event <- 'PFI'
+time <- 'PFI.time'
+
+experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
+cl <- parallel::makeCluster(10)
+parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'event', 'time',
+                                        'surv', 'experiments', 'settings'))
+surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) { 
+  require(data.table)
+  require(survminer)
+  require(survival)
+  source(settings$utility_script)
+  message(experiment)
+  M <- LFs$maui_factors[[experiment]]
+  # get clinical factors for the patients in matrix
+  clin <- process_clinical_covariates(rownames(M))
+  clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
+  # first filter variables by importance 
+  fit <- fit_RSF(surv, M, 500, time, event)
+  vars <- names(head(sort(fit$importance, decreasing = T), 5))
+
+  # compute adjusted survival curve for thb top factor
+  v <- vars[1] 
+  dat <- merge(clin, M[,v,drop = F], by = 'row.names')
+  rownames(dat) <- dat$Row.names
+  dat$Row.names <- NULL
+  dat <- get_surv_df(surv, dat, event, time)
+
+  # split variable into high/low to compute survival adjusted curves
+  cp <- find_survival_cutpoint(dat[,v,drop=F], surv, event_col = event, duration_col = time)
+  dat[,v] <- ifelse(dat[,v] > cp, paste(v, '>', round(cp, 1)), paste(v, '<=', round(cp, 1)))
+  cox2 <- coxph(Surv(time,status) ~ ., 
+                data=dat, x = TRUE)
+  p <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')
+
+  # compute Cindex with all survival-predictive LFs and without LFs 
+  rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, event, time), 
+                            M[,vars], ntree = 1000)
+
+  return(list('p' = p, 
+              'Cindex_base' = rsf_stats$C_wo, 
+              'Cindex_with_LFs' = rsf_stats$C_with))
 })
+parallel::stopCluster(cl)
 
-p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
+# make a plot to show improvement in C index with LFs on top of clinical factors
+ci <- do.call(rbind, lapply(names(surv.stats), function(x) {
+  if(is.null(surv.stats[[x]])) {return(NULL)}
+  df <- data.frame('base' = surv.stats[[x]][['Cindex_base']], 
+                   'with_lf' = surv.stats[[x]][['Cindex_with_LFs']], 
+                   'project' = unlist(strsplit(x, '\\.'))[[1]])
+  return(df)
+}))
 
-Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
+dt <- melt(ci, id.vars = 'project')
+dt$variable <- ifelse(dt$variable == 'base', 'Only clinical factors', 'Clinical factors + MAUI LFs')
+
+p2 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
+  geom_segment(aes(x = reorder(project, with_lf), xend = project, 
+                   y = base, yend = with_lf)) + 
+  geom_point(data = dt, aes(x = project, y = value, color = variable), size = 2) +
+  coord_flip() + 
+  labs(x = '', y = '') + 
+  theme(legend.title = element_blank(), legend.position = 'bottom', 
+        legend.background = element_blank(), 
+        legend.margin = margin(), 
+        legend.direction = 'vertical', legend.box.background = element_blank()) +
+  scale_color_brewer(type = 'qual', palette = 6)
+
+Fig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2, nrow = 2, labels = c('A', 'B')))
 
@@
 Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.clinical_factors.LFs.pdf')
+fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
 names(Figures)[fig_count] <- fig.name
 ggsave(filename = file.path(FigureFolder, fig.name), 
-       plot = Fig, units = 'in', width = 12, height = 9)
-Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type.
-B) Example case: top LFs associated with each gender groups of TCGA-THCA cohort.
-C) Example case: top LFs associated with each histological types of TCGA-UCEC."
+       plot = Fig, width = 8.27, height = 11.69, units = 'in')
+Legends[[fig_count]] <- "Using MAUI LFs for progression-free survival prediction.
+A) Harrell’s C-index computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs.
+B) Hazard ratios of top LFs picked for each cancer adjusted for clinical factors (tumor stage + age + gender).
+C) Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
+
+# combine adjusted survival curve plots into a single plot 
+plots <- sapply(simplify = F, names(surv.stats), function(x) {   
+  p <- surv.stats[[x]][['p']] 
+  if(!is.null(p)) { 
+    p + theme(axis.text.x = element_text(size = 8), 
+              axis.text.y = element_text(size = 8), 
+              axis.title.x = element_blank(), 
+              axis.title.y = element_blank(), 
+              legend.title = element_blank(),
+              legend.position = c(0.5, 0.9),
+              legend.text = element_text(size = 8), 
+              legend.background = element_blank(), 
+              legend.direction = 'vertical') 
+    }
+})
+plots <- plots[!sapply(plots, is.null)]
+p3 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
+                   labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
+                   label_size = 8)
+
+# print p3 to supplements 
+supFig_count <- supFig_count + 1
+SuppFigures[[supFig_count]] <- p3
+fig.name <- paste0('SuppFigure', supFig_count, '.adjusted_survival_curves.pdf')
+names(SuppFigures)[supFig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
+              plot = Fig, units = 'in', width = 8.27, height = 11.69)
+SuppLegends[[supFig_count]] <- "Survival curves adjusted for tumor stage + age + gender based on a cutpoint of 
+top survival-associated LF per cancer cohort."
+
 ```

### 2021-08-26 src/manuscript_figures.Rmd
drop unused functions

@@
 ```{r}
-plot_comp1 <- function(dt) {
-  dt$method <- toupper(gsub("_factors", "", dt$method))
-  comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
-  ggboxplot(dt, 
-          x = 'method', y = 'test_auc', color = 'method', 
-          add = 'jitter', legend = 'none') + 
-    stat_compare_means(method = 'wilcox.test', paired = T, 
-                       method.args = list('alternative' = 'greater'), 
-                     comparisons = comps) +
-    labs(y = 'AUC') + theme(axis.text = element_text(size = 10))
-}
-
-plot_comp2 <- function(dt) {
-  dt$method <- toupper(gsub("_factors", "", dt$method))
-  dtc <- melt(dcast(dt, label ~ method, value.var = 'test_auc'), 
-              id.vars = c('label', 'MAUI'))
-  dtc$difference <- dtc$MAUI - dtc$value
-  ggplot(dtc, 
-         aes(x = MAUI, y = value)) + 
-    geom_point() + geom_abline(slope = 1) + 
-    facet_wrap(~ variable, nrow = 1) + 
-    labs(y = 'AUC')
-}
-
 get_top_by_label <- function(dt, label.type, topN = 5) {
@@
 plots <- plots[!sapply(plots, is.null)]
-p3 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
+Fig <- cowplot::plot_grid(plotlist = plots, nrow = 6, 
                    labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
@@
 supFig_count <- supFig_count + 1
-SuppFigures[[supFig_count]] <- p3
+SuppFigures[[supFig_count]] <- Fig
 fig.name <- paste0('SuppFigure', supFig_count, '.adjusted_survival_curves.pdf')
@@
 top survival-associated LF per cancer cohort."
-
 ```

### 2021-08-26 src/manuscript_figures.Rmd
move function from Rmd to utility script with common functions

@@
   do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
-}
-
-# find percent contribution of different datatypes to the top factor specific variables 
-# LFs: Latent factors; nested list
-# FWs: Feature weights for each LF; nested list 
-# tool: pca/maui/mofa
-# project: e.g. pancancer
-# datatype: e.g. cnv_gex_mut
-# factors: factor vector (e.g. cancer types for each sample or subtype info for each sample)
-# top_lfs: top LF to pick per factor
-# top_featuers: top features to pick per latent factor
-plot_datatype_contribution_by_factor <- function(LFs, FWs, tool, project, datatype, 
-                                                 factors, top_lfs = 1, top_features = 100) {
-  experiment <- paste0(project, '.', datatype)
-  L <- LFs[[tool]][[experiment]]
-  W <- FWs[[tool]][[experiment]]
-  dt <- get_factor_specific_variables(L, factors)
-  top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', topN = top_lfs)
-  # for each factor, pick top lfs and pick top features per lf and plot composition
-  freq <- do.call(rbind, lapply(split(top_vars, top_vars$ref_cl), function(x) {
-    l <- unique(x$variable) # one or more lfs 
-    # get top features per lf and concatenate
-    f <- do.call(c, lapply(l, function(y) {
-      head(sort(abs(W[,y]), decreasing = T), top_features)
-    }))
-    freq <- data.table(table(sub("^(.+?)\\..+$", "\\1", unique(names(f)))))
-    freq$factor <- unique(x$ref_cl)
-    colnames(freq)[1] <- 'omics'
-    return(freq)
-  }))
-
-  ggplot(freq, aes(x = factor, y = N)) + 
-    geom_bar(stat = 'identity', aes(fill = omics), 
-             position = 'fill') + coord_flip() + 
-    labs(x = '', y = 'Percentage of top features contributing to top latent factors')
 }

### 2021-08-26 src/common_functions.R

@@ -1489,6 +1489,40 @@ score_gene_set <- function(rankData, genes_up, gene_down = NULL) {
   return(scores$TotalScore)
 }
 
+# find percent contribution of different datatypes to the top factor specific variables 
+# LFs: Latent factors; nested list
+# FWs: Feature weights for each LF; nested list 
+# tool: pca/maui/mofa
+# project: e.g. pancancer
+# datatype: e.g. cnv_gex_mut
+# factors: factor vector (e.g. cancer types for each sample or subtype info for each sample)
+# top_lfs: top LF to pick per factor
+# top_featuers: top features to pick per latent factor
+plot_datatype_contribution_by_factor <- function(LFs, FWs, tool, project, datatype, 
+                                                 factors, top_lfs = 1, top_features = 100) {
+  experiment <- paste0(project, '.', datatype)
+  L <- LFs[[tool]][[experiment]]
+  W <- FWs[[tool]][[experiment]]
+  dt <- get_factor_specific_variables(L, factors)
+  top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', topN = top_lfs)
+  # for each factor, pick top lfs and pick top features per lf and plot composition
+  freq <- do.call(rbind, lapply(split(top_vars, top_vars$ref_cl), function(x) {
+    l <- unique(x$variable) # one or more lfs 
+    # get top features per lf and concatenate
+    f <- do.call(c, lapply(l, function(y) {
+      head(sort(abs(W[,y]), decreasing = T), top_features)
+    }))
+    freq <- data.table(table(sub("^(.+?)\\..+$", "\\1", unique(names(f)))))
+    freq$factor <- unique(x$ref_cl)
+    colnames(freq)[1] <- 'omics'
+    return(freq)
+  }))
+  
+  ggplot(freq, aes(x = factor, y = N)) + 
+    geom_bar(stat = 'identity', aes(fill = omics), 
+             position = 'fill') + coord_flip() + 
+    labs(x = '', y = 'Percentage of top features contributing to top latent factors')
+}
 
 
 

### 2021-08-26 src/manuscript_figures.Rmd
merge figure 1 and 2 (cancer type prediction + unsupervised clustering benchmarks)
- move cancer type associated LFs plot to supplements
- move prediction accuracy bar plot to supplements
- make a new confusion matrix heatmap in supplements

@@
 ```{r pancancer_prediction, fig.height=9, fig.width=12}
+
+# In this section, we make figures about pan-cancer classification 
+# and some benchmarks of cohort-based unsupervised clustering across cancers
+
+# 1. pancancer classification
 # Find out if cancer types are predictable using latent factors 
@@
   labs(y = 'Mean Balanced Accuracy') + 
-  theme(legend.title = element_blank(), axis.title.x = element_blank(), 
-        axis.text = element_text(size = 14), axis.title = element_text(size = 16))
+  theme(legend.position = 'none', axis.title.x = element_blank(), 
+        axis.text = element_text(size = 14), axis.title = element_text(size = 16)) + 
+  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1))
 
-# performance visualisation
-#m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
-#p2 <- pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
-#         display_numbers = T, number_format = "%.1f", silent = T)
+# performance visualisation - misclassified examples 
+m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
+p2 <- pheatmap::pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
+          silent = T, cellwidth = 10, cellheight = 10)
 
@@
 
-p2 <- plot_tsne(M, ct, show.labels = T) + 
+p3 <- plot_tsne(M, ct, show.labels = T) + 
   theme(legend.position = 'none')
 
-p3 <- plot_label_specific_vars(M, ct)
+p4 <- plot_label_specific_vars(M, ct)
 
 # data type contribution to top LF per factor
-p4 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
+p5 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 
                                            'pancancer', 'cnv_gex_meth_mut', factors = ct, 
                                            top_lfs = 1, top_features = 100) 
-p4 <- p4 + labs(y = 'Relative Contribution\nto top latent factor') + 
+p5 <- p5 + labs(y = 'Relative Contribution\nto top latent factor') + 
   theme(legend.position = 'top', legend.background = element_blank(), 
@@
 
-Fig <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B'),
-                           rel_widths = c(1,2)), 
-                  plot_grid(p4, p3, nrow = 1, labels = c('C', 'D'), 
-                            rel_widths = c(1,2)), 
-                  nrow = 2,
-                  rel_heights = c(2,3))
-
-fig_count <- fig_count + 1
-Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf')
-names(Figures)[fig_count] <- fig.name
-ggsave(file.path(FigureFolder, fig.name),
-              plot = Fig, units = 'in', width = 12, height = 12)
-Legends[[fig_count]] <- "Predictive power of latent factors in distinguishing cancer types.
-A) comparing MAUI with PCA/MOFA/MCIA.
-Mean Balanced Accuracy computed on the test dataset (30% of held-out data) trained on 70% of the dataset using an Elastic Net multi-class model.
-B) tSNE plot of MAUI latent factors learned from pan-cancer dataset  C) Percentage of top feature types contributing to top predictive maui latent factor per cancer type.
-D) Top maui LFs detected per cancer type.
-Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples."
-
-```
-
-```{r clustering_metrics, fig.width=10, fig.height=8}
+# 2. Cohort-based unsupservised clustering benchmark
 # clustering metrics of various cancers 
-# Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs
-
+# Use top most-variable factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
 multi_omics_comb <- 'cnv_gex_meth_mut'
@@
     } 
-    # try topN from 5 to 25 (each tool may capture different amount of information 
-    # by varying number of factors)
-    stats <- do.call(rbind, lapply(seq(5, min(20, ncol(M)), 5), function(n) {
-      message(n)
-      M.sub <- t(subset_features_by_variance(t(M), topN = n))
+    # use top 5 most-variable factors for clustering
+    M.sub <- t(subset_features_by_variance(t(M), topN = 5))
     val <- c('internal') 
     if(ncol(M.sub) > 2) {val <- c('internal', 'stability')}
-      res <- clValid::clValid(obj = M.sub,
+    cl_res <- clValid::clValid(obj = M.sub,
                           nClust = 3:8,
@@
                           nstart = 1000, iter.max = 30)
-      stats <- data.table::as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
+    stats <- data.table::as.data.table(cl_res@measures)[V1 %in% c('Silhouette', 'APN')]
     colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
-      stats$top_factors <- n
-      return(stats)
-    }))
     stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
@@
 
-p1 <- ggboxplot(dt1, 
+p6 <- ggboxplot(dt1, 
           x = 'method', y = 'score', add = 'jitter', color = 'method') + 
@@
 comps <- lapply(setdiff(unique(dt2$method), 'MAUI'), function(x) c('MAUI', x))
-p2 <- ggboxplot(dt2, 
+p7 <- ggboxplot(dt2, 
           x = 'method', y = 'score', add = 'jitter', color = 'method') + 
@@
 
-Fig <- plot_grid(plot_grid(p1, p2, labels = c('A', 'B'), nrow = 1),
-                 p4, nrow = 2, labels = c('', 'C'))
+# combine plots for main figure 
+Fig <- plot_grid(plot_grid(p3, p5, nrow = 1, labels = c('A', 'B'),
+                           rel_widths = c(2,1)), 
+                  plot_grid(p6, p7, nrow = 1, labels = c('C', 'D')), 
+                  nrow = 2,
+                  rel_heights = c(2,1))
 
@@
 Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.cohort_based_metrics.pdf')
+fig.name <- paste0('Figure', fig_count, '.benchmarks.pdf')
 names(Figures)[fig_count] <- fig.name
 ggsave(file.path(FigureFolder, fig.name),
-       plot = Fig, units = 'in', 
-       width = 10, height = 10)
-Legends[[fig_count]] <- "Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in A) and cluster stability (Average Proportion of Non-Overlap (APN) in B).
-Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) for each tool.
-C-D) Survival outcome prediction accuracy for 21 cancer types by each tool's top 10 survival-related factors based on Overall Survival (in C) or Progression Free Survival endpoints.
-(in D)."
+              plot = Fig, units = 'in', width = 12, height = 12)
+Legends[[fig_count]] <- " A) tSNE plot of MAUI latent factors learned from pan-cancer dataset  B) Percentage of top feature types contributing to top predictive maui latent factor per cancer type.
+C, D) Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in C and cluster stability (Average Proportion of Non-Overlap (APN) in D.
+Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) for each tool."
+
+# supplementary figure
+
+supFig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2$gtable, nrow = 1, 
+                                                labels = c('A', 'B'), rel_widths = c(1.2, 2)), 
+                          p4, nrow = 2, labels = c('', 'C'), rel_heights = c(1.2, 2))
+# print p3 to supplements 
+supFig_count <- supFig_count + 1
+SuppFigures[[supFig_count]] <- supFig
+fig.name <- paste0('SuppFigure', supFig_count, '.benchmarks.supp.pdf')
+names(SuppFigures)[supFig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
+              plot = supFig, units = 'in', width = 8.27, height = 11.69)
+SuppLegends[[supFig_count]] <- "A) Predictive power of latent factors in distinguishing cancer types comparing MAUI with PCA/MOFA/MCIA.
+Mean Balanced Accuracy computed on the test dataset (30% of held-out data) trained on 70% of the dataset using an Elastic Net multi-class model.
+B) Confusion matrix of predicted versus true labels of samples' cancer types.
+The color of each cell represents the percentage of samples from the cancer type in the rows that are classified into the cancer types on the column.
+Cells that deviate from the red color represent relatively lower
+level of classification accuracy.
+C) Top maui LFs detected per cancer type.
+Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples."
 ```
@@
        plot = Fig, width = 8.27, height = 11.69, units = 'in')
-Legends[[fig_count]] <- "Using MAUI LFs for progression-free survival prediction.
-A) Harrell’s C-index computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs.
-B) Hazard ratios of top LFs picked for each cancer adjusted for clinical factors (tumor stage + age + gender).
-C) Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
+Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type.
+B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs."
 
@@
               plot = Fig, units = 'in', width = 8.27, height = 11.69)
-SuppLegends[[supFig_count]] <- "Survival curves adjusted for tumor stage + age + gender based on a cutpoint of 
-top survival-associated LF per cancer cohort."
+SuppLegends[[supFig_count]] <- "Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
 ```

### 2021-08-26 src/manuscript_figures.Rmd
drop supp. fig2. (bipartite plots about context dependency of subtyping)

@@
 
-bipartite_plots <- unlist(lapply(1:(ncol(df)-1), function(i) {
-  lapply((i+1):ncol(df), function(j) {
-    p <- plot_cluster_comparison(df[,i,drop = F], df[,j,drop = F]) + 
-      theme(plot.margin = unit(c(0.2, 1.5, 0, 1.5), units = 'in'))
-  })
-}), recursive = F)
-
 Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
@@
 The AMI value ranges from 0 to 1 and the higher the value, the more overlap between patient partitions."
-
-Fig <- cowplot::plot_grid(plotlist = bipartite_plots, nrow = 3, labels = 'AUTO')
-supFig_count <- supFig_count + 1
-SuppFigures[[supFig_count]] <- Fig
-fig.name <- paste0('SuppFigure', supFig_count, '.subtype_dependency.bipartite_plots.pdf')
-names(SuppFigures)[supFig_count] <- fig.name
-ggsave(file.path(FigureFolder, fig.name), 
-              plot = Fig, units = 'in', width = 14, height = 14)
-SuppLegends[[supFig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent"
 ```

### 2021-09-07 src/manuscript_figures.Rmd
Update luad vs lusc analysis figures - drop panel E and F (feature contribution to top LFs, cancersea gene set score distributions) - add an analysis of top omics markers associated to top LFs - draw decision tree + heatmap of top markers in panel E

@@
 library(caret)
+library(treeheatr)
+library(partykit)
+
 theme_set(ggpubr::theme_pubclean())
@@
 ens2hgnc <- readRDS(settings$ens2hgnc)
+cpg2hgnc <- readRDS(settings$cpg2hgnc)
 
@@
 }
+
+map_ids <- function(markers, cpg2hgnc, ens2hgnc) {
+  dt <- data.table::data.table('id' = gsub("^.+?\\.", "", markers), 
+                               'type' = sub("(^.+?)\\..+$", "\\1", markers), 
+                               'variable' = markers)
+  do.call(rbind, lapply(names(split(dt, dt$type)), function(x) {
+    message(x)
+    if(x == 'meth') {
+      dt.map <- cpg2hgnc[cpg_site %in% dt[type == x]$id]
+      colnames(dt.map) <- c('id', 'hgnc')
+      return(merge(dt[type == x], dt.map,  by = 'id'))
+    } else if (x == 'cnv') {
+      dt.map <- dt[type == x]
+      dt.map$hgnc <- gsub("^.+?\\.", "", dt.map$variable)
+      return(dt.map)
+    } else if (x == 'gex' || x == 'mut') {
+      dt.map <- ens2hgnc[ref_gene_id %in% dt[type == x]$id]
+      colnames(dt.map) <- c('id', 'hgnc')
+      return(merge(dt[type == x], dt.map,  by = 'id'))
+    }
+  }))
+}
 ```
@@
 ```{r pancancer_prediction, fig.height=9, fig.width=12}
-
 # In this section, we make figures about pan-cancer classification 
@@
 Figures[[fig_count]] <- Fig
-fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
+fig.name <- paste0('Figure', fig_count, '.clinical_factors.pdf')
 names(Figures)[fig_count] <- fig.name
@@
 M <- LFs$maui_factors$nsclc.cnv_gex_meth_mut
-M <- scale(M) #remove_redundant_variables(M, perc = 99)
-ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
+M <- scale(M) 
+ct <- gsub("TCGA-", "", surv[match(rownames(M), bcr_patient_barcode)]$project)
 
 # find type-specific LFs
-dt <- get_factor_specific_variables(M, ct)
-top_vars <- get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 1)
+subtype_lfs <- get_factor_specific_variables(M, ct)
+top_vars <- get_top_by_label(subtype_lfs[padj < 0.05][order(padj)], 'ref_cl', 1)
 
@@
 # top lf for LUSC
-top_lf_lusc <- top_vars[ref_cl == 'TCGA-LUSC']$variable
-top_lf_luad <- top_vars[ref_cl == 'TCGA-LUAD']$variable
+top_lf_lusc <- top_vars[ref_cl == 'LUSC']$variable
+top_lf_luad <- top_vars[ref_cl == 'LUAD']$variable
 df <- cbind(df, 
@@
             aes(x,y,label = text.label), size = 5)
-p5 <- plot_datatype_contribution_by_factor(LFs, FWs, 'maui_factors', 'nsclc', 'cnv_gex_meth_mut', 
-                                     factors = ct, top_lfs = 1, top_features = 500) + 
-  labs(y = 'Relative Contribution\nto top latent factor')
 
-dt <- data.table(M[,top_vars[['variable']]], keep.rownames = T)
-dt <- dt[rowSums(dt[,2:3]) > 0] 
-dt$pr <- surv[match(dt$rn, bcr_patient_barcode)]$project
+# detect subtype specific features that contribute most 
+# to top LFs and make a decision tree plot  
+## get top 10 markers per top LFs per group
+top_vars <- get_top_by_label(subtype_lfs[padj < 0.05][order(padj)], 'ref_cl', 5)
+W <- FWs$maui_factors$nsclc.cnv_gex_meth_mut
+markers <- unlist(lapply(unique(top_vars$variable), function(x) {
+  # get top features per lf and concatenate
+  head(sort(abs(W[,x]), decreasing = T), 10)
+}))
 
-# alternatively, split samples by LF scores, plot gene set scores 
-scores <- import_geneset_scores(geneSetScoresFolder, annotation = 'cancersea')
-S <- data.table(scores[intersect(rownames(scores), rownames(M)),], keep.rownames = T)
+# import assay data and subset for markers 
+A <- data.table::fread(file.path(settings$assay_output$folder, 
+                                 'nsclc.cnv_gex_meth_mut.csv'))
+A.sub <- A[V1 %in% names(markers)]
+#convert to matrix
+A.subM <- as.matrix(data.frame(A.sub[,-1],
+                               row.names = A.sub$V1, 
+                               check.names = F))
+labels <- gsub("TCGA-", "", surv[match(colnames(A.subM), bcr_patient_barcode)]$project)
 
-dt <- merge(dt, S, by = 'rn')
+# further filter top features 
+top_features <- get_factor_specific_variables(t(A.subM), labels)
+# get top 5 per subtype
+top_markers <- get_top_by_label(top_features[padj < 0.05][order(padj)], 
+                                'ref_cl', topN = 5)
+# map ids 
+top_markers <- merge(top_markers, map_ids(top_markers$variable, cpg2hgnc, ens2hgnc), by = 'variable')
 
-mdt <- melt(dt, measure.vars = setdiff(colnames(S), 'rn'))
+# use treeheatr and partykit package to compute/visualize decision tree
+df <- data.frame(t(round(A.subM[top_markers$variable,], 3)))
+colnames(df) <- paste(top_markers$type, top_markers$hgnc, sep = ":")
+df$subtype <- as.factor(labels)
 
-p6 <- ggplot(mdt, aes(x = value, y = variable)) + 
-  ggridges::geom_density_ridges(aes(fill = get(top_lf_lusc) > get(top_lf_luad)), alpha = 0.7) +
-  labs(x = 'CancerSEA gene set scores', y = 'Cancer Cell States') + 
-  guides(fill=guide_legend(title=paste0(top_lf_lusc, " > ",top_lf_luad)))
+set.seed(123)
+inTrain <- caret::createDataPartition(df$subtype, p = 0.6)
+df.train <- df[inTrain[[1]],]
+df.test <- df[-inTrain[[1]],]
+p5 <- treeheatr::heat_tree(
+  x = partykit::ctree(subtype ~ ., data = df.train),
+  data_test = df.test,
+  heat_rel_height = 0.4
+)
 
-Fig <-  cowplot::plot_grid(p1, p2, p3, p4, p5, p6, 
-                         labels = 'AUTO', ncol = 2)
+# compute classification accuracy of using top 10 markers in a random forest model
+rf_stats <- predict_cluster_labels(t(A.subM[top_markers$variable,]), labels, partition = 0.6)
+
+Fig <-  cowplot::plot_grid(cowplot::plot_grid(p1, p2, p3, p4,
+                         labels = 'AUTO', nrow = 2), 
+                         p5, nrow = 2, labels = c('', 'E'))
 
@@
        Fig, units = 'in', 
-       width = 10, height = 12)
+       width = 8.27, height = 11.69)
+
 Legends[[fig_count]] <- "Classification of Non-Small-Cell Lung Adenocarcinoma samples using MAUI latent factors.
@@
 the top most-predictive latent factor for TCGA-LUSC samples. D) t-SNE plot of MAUI LFs colored by the top 
-most predictive latent factor for TCGA-LUAD samples.
-E) Percentage of top 500 contributing features each of the top most predictive latent factors broken down into categories by cancer type and data type.
-F) Cancer Cell State score distribution of samples grouped by whether the sample has a higher score for LUSC-predictive latent factor than the LUAD-predictive latent factor."
+most predictive latent factor for TCGA-LUAD samples. E) Decision tree visualisation of top LF-associated omics markers detected that can be used to distinguish LUSC and LUAD subtypes."
 ```

### 2021-09-13 src/manuscript_figures.Rmd
add jonas tmz gbm analysis figures

@@
 
+```{r tmz_gbm}
+# TMZ resistance in GBM anaysis by Jona.
+# Here I import figure data to have consistent figure styles
+folder <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/tmz_gbm_jona'
+
+# cross-validation results predicting TMZ resistance using 
+# maui LFs and other known markers 
+aucs <- data.table::fread(file.path(folder, 'aucs.csv'))
+aucs$group <- ifelse(grepl('LF', aucs$method), 'maui', 'no-maui')
+aucs$method <- gsub("LF[Ss]", "MAUI LFs", aucs$method)
+colnames(aucs)[2] <- c('AUC')
+Fig <- ggpubr::ggbarplot(aucs, x = 'method', y = 'AUC', add = 'mean_se',
+                  fill = 'group') +
+  theme(axis.title.x = element_blank(), legend.position = 'none', 
+        ) + 
+  scale_fill_brewer(type = 'qual', palette = 3)
+
+fig_count <- fig_count + 1
+Figures[[fig_count]] <- Fig
+fig.name <- paste0('Figure', fig_count, '.tmz_gbm.pdf')
+names(Figures)[fig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name),
+              plot = Fig, units = 'in', width = 10, height = 6)
+Legends[[fig_count]] <- "Classification accuracy quantified as the Area under the ROC (AUC) when predicting TMZ resistance in GBM patients using MAUI LFs alone or in combination with the MGMT expression or promoter methylation values."
+
+# supplementary figure
+topGenes <- data.table::fread(file.path(folder, 'top100_genes.csv'))
+colnames(topGenes)[4] <- 'literature'
+topGenes$label <- 'Other'
+topGenes[MMR == TRUE]$label <- 'MMR Pathway'
+topGenes[literature == TRUE]$label <- 'Literature Support'
+
+supFig <- ggpubr::ggbarplot(topGenes, x = 'V1', y = 'Gene score', 
+                  fill = 'label') +
+  theme(axis.title.y = element_blank(), legend.title = element_blank(),
+        axis.text.y = element_text(size = 8)) +
+  coord_flip()
+
+supFig_count <- supFig_count + 1
+SuppFigures[[supFig_count]] <- supFig
+fig.name <- paste0('SuppFigure', supFig_count, '.tmz_gbm.supp.pdf')
+names(SuppFigures)[supFig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
+              plot = supFig, units = 'in', width = 8.27, height = 11.69)
+SuppLegends[[supFig_count]] <- "Top 100 genes sorted by importance in predicting TMZ resistance in GBM patients."
+```
+
 ```{r print_figures, results='asis', include=TRUE, fig.width=10, fig.height=11}

### 2021-09-13 src/run_mofa.R

@@ -1,5 +1,5 @@
 library(reticulate)
-# run this in interactive mode, turn it off for command-line (I know it is stupid)
+# run this in interactive mode, turn it off for command-line
 # use_python('/home/buyar/.cache/basilisk/1.2.0/MOFA2-1.0.0/mofa_env/bin/python3.7', required = TRUE)
 args <- commandArgs(trailingOnly = TRUE)
 s <- import("sys")

### 2021-09-14 src/common_functions.R
split model building and evaluation for glmnet regression

@@ -348,14 +348,24 @@ predict_numerical <- function(M, y, ...) {
   
 }
 
+evaluate_regression_model <- function(y, y_hat) {
+  # Model performance metrics
+  data.table::data.table(
+    RMSE = RMSE(y_hat, y),
+    Rsquare = R2(y_hat, y),
+    COR = cor(y_hat, y)
+  )
+}
+  
+  
 # elastic net model (cross-validate using 10 )
-predict_numerical_glmnet <- function(M, y) {
+predict_numerical_glmnet <- function(M, y, partition = 0.6) {
   require(glmnet)
   require(caret)
   df <- data.frame(M)
   df$y <- y
   df <- df[!is.na(df$y),]
-  intrain <- caret::createDataPartition(y = df$y, p= 0.6, list = FALSE)
+  intrain <- caret::createDataPartition(y = df$y, p = partition, list = FALSE)
   training <- df[intrain,]
   testing <- df[-intrain,]
   # Build the model using the training set
@@ -367,14 +377,9 @@ predict_numerical_glmnet <- function(M, y) {
   )
   # Best tuning parameter
   model$bestTune
-  x.test <- model.matrix(y ~., testing)[,-1]
-  predictions <- predict(model, x.test)
-  # Model performance metrics
-  data.frame(
-    RMSE = RMSE(predictions, testing$y),
-    Rsquare = R2(predictions, testing$y),
-    COR = cor(predictions, testing$y)
-  )
+  preds_test <- predict(model, model.matrix(y ~., testing)[,-1])
+  return(list('model' = model, 'preds' = data.frame('y' = testing$y, 
+                                                'y_hat' = preds_test)))
 }
 
 
@@ -1095,6 +1100,18 @@ process_clinical_covariates <- function(patientIDs) {
   return(clinicalData)
 }
 
+
+get_tumor_purity_estimates <- function(patientIDs) {
+  dt <- data.table::data.table(TCGAbiolinks::Tumor.purity)
+  dt <- data.table('Sample.ID' = dt$Sample.ID, 
+             'ESTIMATE' = as.numeric(gsub(",", ".", dt$ESTIMATE)), 
+             'ABSOLUTE' = as.numeric(gsub(",", ".", dt$ABSOLUTE)), 
+             'LUMP' = as.numeric(gsub(",", ".", dt$LUMP)), 
+             'IHC' = as.numeric(gsub(",", ".", dt$IHC)), 
[…]

### 2021-09-14 src/manuscript_figures.Rmd
set global fig.width and fig.height to A4 dimensions

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
+knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, 
+                      fig.width = 8.27, fig.height = 11.69)
 knitr::opts_knit$set(root.dir = params$workdir)
@@
 
-```{r pancancer_prediction, fig.height=9, fig.width=12}
+```{r pancancer_prediction}
 # In this section, we make figures about pan-cancer classification 
@@
 
-```{r clinical_survival_factors, fig.width=12, fig.height=8}
+```{r clinical_survival_factors}
 # here we compute LFs associated to clinical factors (e.g. tumor stage, gender etc) and 
@@
 
-```{r msi_stats, fig.width=10, fig.height=8}
+```{r tumor_purity}
+
+```
+
+```{r msi_stats}
 # Predicting MSI status using latent factors
@@
 
-```{r print_figures, results='asis', include=TRUE, fig.width=10, fig.height=11}
+```{r print_figures, results='asis', include=TRUE}
 base_url <- 'https://bimsbstatic.mdc-berlin.de'

### 2021-09-14 src/manuscript_figures.Rmd
add tumor purity prediction analysis figure as supplements

@@
 ```{r tumor_purity} 
+experiments <- grep('TCGA.*cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
+# use maui LFs to predict TCGA IHC/CPE values that represent tumor purity levels 
+cl <- parallel::makeCluster(20)
+parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
+results <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
+    message(experiment)
+    source(settings$utility_script)
+    library(data.table)
+    M <- LFs$maui_factors[[experiment]]
+    tp <- get_tumor_purity_estimates(rownames(M))
+    if(nrow(tp) < 50) {
+      warning("Couldn't find at least 100 samples with tumor purity 
+              scores in experiment ", experiment, " => ", nrow(tp))
+      return(NULL)
+    }
+    # predict IHC
+    y1 <- tp[,mean(IHC), by = bcr_patient_barcode][match(rownames(M), bcr_patient_barcode)]$V1
+    res1 <- predict_numerical_glmnet(M, y1)
+    # predict CPE (consensus estimates)
+    y2 <- tp[,mean(CPE), by = bcr_patient_barcode][match(rownames(M), bcr_patient_barcode)]$V1
+    res2 <- predict_numerical_glmnet(M, y2)
+    return(list('IHC' = res1, 'CPE' = res2))
+})
+parallel::stopCluster(cl)
 
+# summarize stats 
+stats <- do.call(rbind, lapply(names(results), function(x) {
+  message(x)
+  if(is.null(results[[x]])) {
+    return(NULL)
+  }
+  dt <- do.call(rbind, lapply(names(results[[x]]), function(y) {
+    message(y)
+    dt <- evaluate_regression_model(results[[x]][[y]][['preds']]$y, 
+                              results[[x]][[y]][['preds']]$y_hat)
+    dt$purity_method <- y
+    return(dt)
+  }))
+  dt$experiment <- x
+  return(dt)
+}))
+
+stats$project <- unlist(lapply(strsplit(stats$experiment, "\\."), function(x) x[[1]]))
+
+# plot correlation scores for all cohorts 
+p1 <- ggplot(stats[purity_method == 'IHC'][grep('TCGA', project)], 
+       aes(x = reorder(project, COR), y = COR, fill = COR),) + 
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  theme(axis.title.y = element_blank(), legend.position = 'none') + 
+  labs(y = "MAUI vs IHC") + ylim(0, 1)
+
+p2 <- ggplot(stats[purity_method == 'CPE'][grep('TCGA', project)], 
+       aes(x = reorder(project, COR), y = COR, fill = COR)) + 
+  geom_bar(stat = 'identity') +
+  coord_flip() +
+  theme(axis.title.y = element_blank(),  legend.position = 'none') + 
+  labs(y = "MAUI vs CPE") + ylim(0, 1)
+
+# correlation plots for top cancer types
+selected <- stats[purity_method == 'IHC'][grep('TCGA', experiment)][order(COR, decreasing = T)][1:4]$experiment
+plots <- lapply(selected, function(x) {
+  message(x)
+  p <- cowplot::plot_grid(plotlist = lapply(names(results[[x]]), function(y) {
+    message(y)
+    ggpubr::ggscatter(results[[x]][[y]][['preds']], 
+                           x = 'y_hat', y = 'y', 
+                           add = 'reg.line', cor.coef = T, cor.method = 'pearson') + 
+      labs(x = paste('MAUI prediction of', y), y = y, title = gsub("\\..+$", "", x))
+  }))
+  return(p)
+})
+
+p3 <- cowplot::plot_grid(plotlist = plots, ncol = 1)
+
+# supplementary figure
+supFig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2, labels = c('A', 'B')), 
+                             p3, nrow = 2, labels = c('', 'C'), 
+                             rel_heights = c(2, 5))
+
+# print p3 to supplements 
+supFig_count <- supFig_count + 1
+SuppFigures[[supFig_count]] <- supFig
+fig.name <- paste0('SuppFigure', supFig_count, '.tumor_purity.supp.pdf')
+names(SuppFigures)[supFig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
+              plot = supFig, units = 'in', width = 8.27, height = 11.69)
+SuppLegends[[supFig_count]] <- "Predicting tumor purity estimates using MAUI latent factors. 
+A) Correlation of the MAUI predictions of IHC versus the annotated IHC values (immunohistochemistry-based tumor purity estimation). 
+B) Correlation of the MAUI predictions of CPE versus the annotated CPE levels (CPE stands for consensus purity estimated by computational methods: ABSOLUTE, ESTIMATE, and LUMP). 
+C) Scatterplots of MAUI predictions versus the annotated IHC and CPE values for the 
+top best predicted cancer types."
 ```

### 2021-09-14 src/manuscript_figures.Rmd
remove pathway and top feature analysis from immunotherapy analysis figure (panels F and G)

@@
 Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples."
+
 ```
@@
 settings_imtrp <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/immunotherapy_response/Mariathasan_et_al/settings.yaml')
-pathways <- readRDS('/data/local/buyar/arcas/subtyping_paper/data/GeneSets/msigdb.pathways.RDS')
 
@@
   dt <- get_factor_specific_variables(M, colData$binaryResponse)[order(padj)]
-  #dt <- compute_anova(M, colData$binaryResponse)
   dt$tool <- tool
@@
 
-dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')]
 dt$tool <- toupper(gsub("_factors", "", dt$tool))
@@
 
-# top go terms for top features of top LF 
-W <- FWs_imtrp$maui_factors$mariathasan.gex
-feats <- data.table(W[,top_lf,drop=F], keep.rownames = T)[order(abs(get(top_lf)), decreasing = T)][1:100]
-# top pathways for top lf
-top_terms <- compute_geneset_enrichment(feats$rn, pathways)[order(pval)][padj < 0.05][1:5]
-top_terms$term_wr <- stringr::str_wrap(gsub("_", " ", top_terms$term), width = 24)
-p4 <- ggplot(top_terms, aes(x = reorder(term_wr, -pval), y = -log10(padj))) + 
-  geom_bar(stat = 'identity') + 
-  coord_flip() + labs(x = '') 
-colnames(feats)[2] <- 'top_lf'
-p5 <- ggplot(feats[1:10,], aes(x = reorder(rn, abs(top_lf)), y = abs(top_lf))) + 
-  geom_bar(stat = 'identity') + coord_flip() + labs(x = '', y = 'Feature Importance (absolute)')
-
 # correlation of top LF with tumor mutation burden
@@
 dt$tool <- toupper(gsub("_factors", "", dt$tool))
-p6 <- ggplot(dt, aes(x = tool, y = abs(V1))) + 
+p4 <- ggplot(dt, aes(x = tool, y = abs(V1))) + 
   geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
@@
 dt$tool <- toupper(gsub("_factors", "", dt$tool))
-p7 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
+p5 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
              aes(x = ref_cl, y = -log10(padj))) + 
@@
 Fig <- cowplot::plot_grid(
-  cowplot::plot_grid(p1, p6, p7, nrow = 1, labels = 'AUTO'),
-  cowplot::plot_grid(p2$gtable, p3$plot, p4, p5, nrow = 2, labels = c('D', 'E', 'F', 'G')),
+  cowplot::plot_grid(p1, p4, p5, nrow = 1, labels = 'AUTO'),
+  cowplot::plot_grid(p2$gtable, p3$plot, nrow = 1, labels = c('D', 'E')),
   nrow = 2, 
@@
 D) Top MAUI LFs predictive of response to anti-PD-L1 response groups.
-E) Survival stratification of the cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4).
-F) Top pathways enriched among top 100 genes that contribute to the top latent factor (LF44).
-G) Top 10 genes contributing to LF44"
+E) Survival stratification of the cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4)."
 ```

### 2021-09-16 src/common_functions.R
add simplified version of predict_cluster_labels_glmnet

@@ -235,8 +235,34 @@ predict_cluster_labels <- function(M, labels, partition = 0.7, ...) {
   return(auc.dt)
 }
 
+# simplified version of predict_cluster_labels_glmnet
+predict_categorical <- function(df, ref_class) {
+  require(caret)
+  require(ranger)
+  # for each factor level, do a one-vs-all classification with probabilities
+  df$y <- as.factor(ifelse(df$y == ref_class, 'one', 'rest'))
+  set.seed(3456)
+  intrain <- caret::createDataPartition(y = df$y, p= 0.6, list = FALSE)
+  training <- df[intrain,]
+  testing <- df[-intrain,]
+  # Build the model using the training set
+  set.seed(123)
+  require(doParallel)
+  cl <- makePSOCKcluster(5)
+  registerDoParallel(cl)
+  ctrl <- trainControl("repeatedcv", repeats = 5, number = 5, sampling = 'down')
+  model <- train(
+    y ~., data = training, method = "glmnet",
+    trControl = ctrl,
+    tuneLength = 10
+  )
+  stopCluster(cl)
+  return(model)
+}
+
 # returnROC: if set to TRUE, will return a ROC object, when FALSE returns AUC stats 
-predict_cluster_labels_glmnet <- function(M, labels, returnROC = FALSE) {
+predict_cluster_labels_glmnet <- function(M, labels, returnModel = FALSE, returnROC = FALSE, 
+                                          repeats = NULL, sampling = NULL) {
   require(caret)
   require(ranger)
   df <- data.frame(M)
@@ -245,6 +271,7 @@ predict_cluster_labels_glmnet <- function(M, labels, returnROC = FALSE) {
   res <- sapply(simplify = F, levels(labels), function(x) {
     message(x)
     df$y <- as.factor(ifelse(labels == x, 'one', 'rest'))
+    df <- df[!is.na(df$y),]
     set.seed(3456)
     intrain <- caret::createDataPartition(y = df$y, p= 0.6, list = FALSE)
     training <- df[intrain,]
@@ -255,12 +282,16 @@ predict_cluster_labels_glmnet <- function(M, labels, returnROC = FALSE) {
       require(doParallel)
       cl <- makePSOCKcluster(5)
       registerDoParallel(cl)
+      ctrl <- trainControl("repeatedcv", number = 5)
+      if(!is.null(sampling)) {ctrl$sampling <- sampling}
+      if(!is.null(repeats)) {ctrl$repeats <- repeats}
[…]

### 2021-09-16 src/run_maui.sh
use conda to activate env

@@ -1,4 +1,4 @@
-source activate maui
+conda activate maui
 export PYTHONPATH=/home/buyar/.conda/envs/maui/lib/python3.6/site-packages
 assays=$1 # assay matrix file path
 outFile=$2 # path to output file

### 2021-09-16 src/manuscript_figures.Rmd
Update immunotherapy response analysis (add marker combination analysis)  and make cosmetic changes to some figures

@@
 knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, 
-                      fig.width = 8.27, fig.height = 11.69)
+                      fig.width = 9, fig.height = 11)
 knitr::opts_knit$set(root.dir = params$workdir)
@@
 m <- as.matrix(table(res$maui_factors$pred, res$maui_factors$obs))
-p2 <- pheatmap::pheatmap(apply(m, 1, function(x) round(x/sum(x) * 100, 1)), cluster_rows = F, cluster_cols = F, 
-          silent = T, cellwidth = 10, cellheight = 10)
+# compute percentage of correct predictions
+hm <- apply(m, 1, function(x) round(x/sum(x) * 100))
+hm_labels <- hm
+hm_labels[hm_labels < 2] <- '' #don't display very low values
+p2 <- pheatmap::pheatmap(hm, cluster_rows = F, cluster_cols = F, 
+          silent = T, cellwidth = 10, cellheight = 10, color = colorRampPalette(c('white', 'red'))(10), 
+          display_numbers = hm_labels, fontsize = 8)
 
@@
 ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
+dt <- get_factor_specific_variables(M, ct)[padj < 0.05][order(padj)]
+top_vars <- get_top_by_label(dt, 'ref_cl', topN = 5)
 
-p3 <- plot_tsne(M, ct, show.labels = T) + 
+p3 <- plot_tsne(M[,unique(top_vars$variable)], ct, show.labels = T) + 
   theme(legend.position = 'none')
@@
                                            'pancancer', 'cnv_gex_meth_mut', factors = ct, 
-                                           top_lfs = 1, top_features = 100) 
+                                           top_lfs = 1, top_features = 500) 
 p5 <- p5 + labs(y = 'Relative Contribution\nto top latent factor') + 
-  theme(legend.position = 'top', legend.background = element_blank(), 
-        legend.title = element_blank())
+  theme(legend.position = 'right', legend.direction = 'vertical', 
+        legend.background = element_blank(), 
+        legend.title = element_blank(), legend.box.just = 'left')
 
@@
 Fig <- plot_grid(plot_grid(p3, p5, nrow = 1, labels = c('A', 'B'),
-                           rel_widths = c(2,1)), 
-                  plot_grid(p6, p7, nrow = 1, labels = c('C', 'D')), 
+                           rel_widths = c(2,1.5)), 
+                  plot_grid(p6, p7, 
+                            nrow = 1, labels = c('C','D')), 
                   nrow = 2,
@@
               plot = Fig, units = 'in', width = 12, height = 12)
-Legends[[fig_count]] <- " A) tSNE plot of MAUI latent factors learned from pan-cancer dataset  B) Percentage of top feature types contributing to top predictive maui latent factor per cancer type.
+Legends[[fig_count]] <- " A) tSNE plot of MAUI latent factors (using top 5 latent factors enriched per cancer type) learned from pan-cancer dataset  B) Percentage of top feature types contributing to top predictive maui latent factor per cancer type.
 C, D) Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in C and cluster stability (Average Proportion of Non-Overlap (APN) in D.
-Each point represents a clustering score for 132 k-means clustering results (21 cancer types x for 6 different values of k from 3 to 8) for each tool."
+Each point represents a score for an individual clustering experiment in a total of 132 k-means clustering experiments (21 cancer types x for 6 different values of k from 3 to 8) for each tool."
 
 # supplementary figure
-
 supFig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2$gtable, nrow = 1, 
@@
 Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples."
-
 ```
@@
 # Find maui LFs associated to clinical factors and plot them
-
 experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
@@
 dt <- melt(ci, id.vars = 'project')
-dt$variable <- ifelse(dt$variable == 'base', 'Only clinical factors', 'Clinical factors + MAUI LFs')
+dt$variable <- ifelse(dt$variable == 'base', 'Age+gender+stage', 'Age+gender+stage + MAUI LFs')
 
@@
                    y = base, yend = with_lf)) + 
+  geom_hline(yintercept = c(0.5, 0.6, 0.7), linetype = 2) + 
   geom_point(data = dt, aes(x = project, y = value, color = variable), size = 2) +
@@
 Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type.
-B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (tumor stage + age + gender) or using clinical factors along with survival-predictive MAUI LFs."
+B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (age + gender + tumor stage) or using clinical factors along with survival-predictive MAUI LFs."
 
@@
 # TODO: pass paths via args
-settings_imtrp <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/immunotherapy_response/Mariathasan_et_al/settings.yaml')
+projDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/immunotherapy_response/Mariathasan_et_al'
+settings_imtrp <- yaml::read_yaml(file.path(projDir, 'settings.yaml'))
+gex_all <- readRDS(file.path(projDir, 'mariathasan.gex.all_genes.normalised.RDS'))
+human_gene_signatures <- readRDS(file.path(projDir, 'human_gene_signatures.RDS'))
 
@@
 
-# heatmap of top vars 
-# plot maui factors for responders/non-responders
-M <- LFs_imtrp$maui_factors$mariathasan.gex
-top_vars <- unique(dt[tool == 'MAUI'][padj < 0.01]$variable)
-m <- M[,top_vars]
-f <- ifelse(colData$binaryResponse == 'CR/PR', 'Responders', 'Non-responders')
-m <- m[!is.na(f),]
-f <- f[!is.na(f)]
-p2 <- pheatmap::pheatmap(get_basis_matrix(m[,top_vars], f), silent = T)
-
-# survival stratification using the top LF 
-require(survminer)
-require(survival)
-top_lf <- dt[order(padj)][ref_cl == 'Responders'][1,]$variable
-cp <- find_survival_cutpoint(m[,top_lf,drop = F], colData, event_col = 'censOS', duration_col = 'os')
-p3 <- plot_survival(surv = colData, 
-                    partition.df = data.frame('group' = ifelse(m[,top_lf] > cp, 
-                                              paste0(top_lf, ">", round(cp, 1)),
-                                              paste0(top_lf, "<=", round(cp, 1))), 
-                                              row.names = rownames(m)), 
-                    duration_col = 'os', event_col = 'censOS')
-
 # correlation of top LF with tumor mutation burden
@@
 dt$tool <- toupper(gsub("_factors", "", dt$tool))
-p4 <- ggplot(dt, aes(x = tool, y = abs(V1))) + 
+p2 <- ggplot(dt, aes(x = tool, y = abs(V1))) + 
   geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
@@
   labs(y = 'Correlation with Tumor Mutation Burden') + 
-  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
+  theme(legend.position = 'none', axis.title.y = element_blank(), 
         plot.title = element_text(hjust = 0.5)) 
@@
 dt$tool <- toupper(gsub("_factors", "", dt$tool))
-p5 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
+p3 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
              aes(x = ref_cl, y = -log10(padj))) + 
@@
 
+# test comparison of maui LFs and published markers from Mariathasan et al
+# in terms of response prediction
+# published markers from Mariathasan et al (https://www.nature.com/articles/nature25501#Sec4)
+# 1. TMB (tumor neoantigen burden) 2. CD8 effector T cell markers (8 genes)
+# 3. F-TBRS: fibroblast tgf-beta response signature (4 genes see https://www.nature.com/articles/nature25501/figures/3)
+
+# find top maui LFs informative for response prediction 
+M <- scale(LFs_imtrp$maui_factors$mariathasan.gex)
+labels <- colData[match(rownames(M), colData$bcr_patient_barcode),]$binaryResponse
+df <- cbind(data.frame(M), 'y' = labels)
+df <- df[!is.na(df$y),]
+df$y <- as.factor(df$y)
+# build an initial glmnet model using all LFs on training data
+fit <- predict_categorical(df, 'CR/PR') 
+# get subset of top LFs 
+v <- caret::varImp(fit)[['importance']]
+top_lfs <- data.table(v, keep.rownames = T)[order(Overall, decreasing = T)][Overall > 0]$rn
+
+# make a heatmap of top predictive lfs 
+m <- as.matrix(df[,-ncol(df)])
+f <- ifelse(df$y == 'CR/PR', 'Responders', 'Non-responders')
+p4 <- pheatmap::pheatmap(get_basis_matrix(m[,top_lfs], f), silent = T, cellheight = 50, cluster_rows = F)
+
+# survival stratification using the top LF 
+require(survminer)
+require(survival)
+top_lf <- top_lfs[1]
+cp <- find_survival_cutpoint(m[,top_lf,drop = F], colData, event_col = 'censOS', duration_col = 'os')
+p5 <- plot_survival(surv = colData, 
+                    partition.df = data.frame('group' = ifelse(m[,top_lf] > cp, 
+                                              paste0(top_lf, ">", round(cp, 1)),
+                                              paste0(top_lf, "<=", round(cp, 1))), 
+                                              row.names = rownames(m)), 
+                    duration_col = 'os', event_col = 'censOS')
+
+# get the assay data for the studied markers for CD8 effectors + TBRS etc.
+signatures <- list('cd8_Teff' = human_gene_signatures$`CD 8 T effector`, 
+                   'F_TBRS' = c('ACTA2', 'COL4A1', 'TAGLN', 'SH3PXD2A'))
+# make a data frame of marker genes + TMB measures + top MAUI LFs
+DF <- merge(data.frame(t(gex_all[unlist(signatures),])), 
+            data.frame(M[,top_lfs]), by = 'row.names')
+DF$TMB <- colData[match(DF$Row.names, colData$bcr_patient_barcode),]$`Neoantigen burden per MB`
+# response variable
+DF$y <- colData[match(DF$Row.names, colData$bcr_patient_barcode),]$binaryResponse
+# remove samples without response variable
+DF <- DF[!is.na(DF$y),]
+# remove samples without TMB values 
+DF <- DF[!is.na(DF$TMB),]
+DF <- data.frame(DF[,-1], row.names = DF$Row.names)
+DF$y <- as.factor(DF$y)
+
+L <- list('TMB' = 'TMB',
+          'MAUI' = top_lfs, 
+          'CD8_Teff + F_TBRS' = c(signatures$cd8_Teff, signatures$F_TBRS))
+# now predict response using subsets of features 
+# get combination of models
+combs <- c(asplit(combn(names(L), 2), 2),
+           asplit(combn(names(L), 3), 2), 
+           'MAUI',
+           'CD8_Teff + F_TBRS')
+results <- do.call(rbind, lapply(combs, function(x) {
+  comb <- paste0(x, collapse = " + ")
+  message("trying combination:",comb)
+  included <- as.vector(unlist(L[x]))  # variables included in the model
+  message("including variables:",paste(included, collapse = ' '))
+  res <- predict_cluster_labels_glmnet(scale(as.matrix(DF[,intersect(colnames(DF), included),drop = F])), 
+                                DF$y, repeats = 5, sampling = 'down')
+  res$comb <- comb
+  return(res)
+}))
+
+results$comb <- stringr::str_wrap(results$comb, width = 12)
+p6 <- ggplot(results[order(auc)][label == 'CR/PR'], 
+       aes(x = reorder(comb, -auc), y = auc)) + 
+  geom_bar(stat = 'identity', aes(fill = auc)) + 
+  geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), stat = 'identity', 
+                width = 0.1) + 
+  labs(x = 'Marker Combination', y = 'AUC') + 
+  theme(legend.position = 'none', axis.text.x = element_text(angle = 30, hjust = 1))
+
+p5 <- p5$plot + theme(legend.title = element_blank(), legend.direction = 'vertical')
 Fig <- cowplot::plot_grid(
-  cowplot::plot_grid(p1, p4, p5, nrow = 1, labels = 'AUTO'),
-  cowplot::plot_grid(p2$gtable, p3$plot, nrow = 1, labels = c('D', 'E')),
-  nrow = 2, 
-  rel_heights = c(1,2)
+  cowplot::plot_grid(p1, p3, p2, nrow = 1, labels = c('A', 'B', 'C')), 
+  cowplot::plot_grid(p5, p4$gtable, nrow = 1, labels = c('D', 'E')), 
+  cowplot::plot_grid(cowplot::ggdraw(), p6, cowplot::ggdraw(), rel_widths = c(1, 5, 1), nrow = 1, labels = c('', 'F', '')), 
+  nrow = 3
 )
@@
 Legends[[fig_count]] <- "Predicting responders to anti-PD-L1 treatment in a cohort of Metastatic Urothelial
-Cancers (N=348) using MAUI.
+Cancers (N=216) using MAUI.
 A) Comparison of top latent factors predictive of complete/partial response (CR/PR) to the anti-PD-L1.
-B) Comparison of top latent factors with respect to their correlation with tumor mutation burden status of the samples in the cohort C) Comparison of top latent factors predictive of immune phenotypes (inflamed/excluded/desert).
-D) Top MAUI LFs predictive of response to anti-PD-L1 response groups.
-E) Survival stratification of the cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4)."
+B) Comparison of top latent factors predictive of immune phenotypes (inflamed/excluded/desert).
+C)  Comparison of top latent factors with respect to their correlation with tumor mutation burden status of the samples in the cohort D) Survival stratification of the cohort based activation score of top MAUI latent factor E) Average value of top MAUI LFs predictive of response to anti-PD-L1 response groups.
+F) Comparison of marker combinations in terms of accuracy in predicting anti-PD-L1 response. TMB: tumor mutation burden, 
+CD8_Teff: 12 gene signature for CD8+ effector T-cells, F_TBRS: 4-gene pan-fibroblast TGFβ response signature, MAUI: top predictive MAUI LFs as seen in panel E).
+Each bar plot represents prediction accuracy (AUC) when using the corresponding subset of variables in an Elastic Net model."
 ```

### 2021-09-16 src/manuscript_figures.Rmd
move partition comparison supp-figure to the end
update tmz+gbm analysis to include umap and survival plots

@@
 
-```{r}
-# Show subtypes depend on the objective (dependent variable) - they are dynamic, not universal or carved in stone 
-# for a given cancer. There is not just one set of subtypes for a given cancer. Subtypes need to be phrased 
-# along with the dependent variable used to come up with the subtypes. 
-# Patients clustering depend on the objective of stratification
-clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
-project <- 'pan-gastrointestinal'
-datatype <- 'cnv_gex_meth_mut'
-experiment <- paste0(project, '.', datatype)
-
-M <- LFs$maui_factors[[experiment]]
-clin <- process_clinical_covariates(rownames(M))
-clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
-
-pfi_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'PFI', duration_col = 'PFI.time')
-pfi_clust <- get_kmeans_subtypes(M[,unique(pfi_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
-pfi_k <- names(which.min(sapply(colnames(pfi_clust), function(x) plot_survival(surv, pfi_clust[,x,drop = F], 'PFI.time', 'PFI', T))))
-
-os_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'OS', duration_col = 'OS.time')
-os_clust <- get_kmeans_subtypes(M[,unique(os_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
-os_k <- names(which.min(sapply(colnames(os_clust), function(x) plot_survival(surv, os_clust[,x,drop = F], 
-                                                                             'OS.time', 'OS', T))))
-
-# combine with known pan-gi subtype annotation
-df <- merge(pfi_clust[,pfi_k,drop = F], os_clust[,os_k,drop = F], by = 'row.names')
-df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected
-df[is.na(df$subtype),]$subtype <- 'Unavailable'
-df$msi <- MSI[match(df$Row.names, bcr_patient_barcode)]$MSI
-df[is.na(df$msi),]$msi <- 'Unavailable'
-rownames(df) <- df$Row.names
-df$Row.names <- NULL
-colnames(df) <- c('Overall Survival', 'Progression Free Survival', 'Molecular Subtype', 'MSI Status')
-df <- data.frame(apply(df, 2, as.factor), check.names = F)
-df <- df[order(df[,1]),]
-m <- apply(df, 2, function(x) as.numeric(as.factor(x)))
-rownames(m) <- rownames(df)
-
-colors <- structure(1:7, names = c("1", "2", "3", "4", "5", "6", "7"))
-p1 <- Heatmap(t(m), name = "cluster", col = colors, cluster_rows = F, cluster_columns = F,
-              show_column_names = F, row_names_side = 'left',
-              column_title = 'Pan-gastrointestinal Cancer Samples\n(ordered by Overall Survival Clusters)')
-p1 <- grid::grid.grabExpr(ComplexHeatmap::draw(p1))
-# compute pairwise AMI 
-ami <- sapply(colnames(df), function(x) {
-  sapply(colnames(df), function(y) {
-    df <- df[!is.na(df[,x]) & !is.na(df[,y]),]
-    aricode::AMI(df[,x], df[,y])
-  })
-})
-
-p2 <- ggcorrplot::ggcorrplot(ami, legend.title = 'AMI', lab = T, type = 'lower', outline.color = 'white') +
-  scale_fill_gradient(limits = c(0,1), low = 'white', high = 'red') + 
-  guides(fill=guide_legend(title='AMI'))
-
-Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
-supFig_count <- supFig_count + 1
-SuppFigures[[supFig_count]] <- Fig
-fig.name <- paste0('SuppFigure', supFig_count, '.subtype_dependency.pdf')
-names(SuppFigures)[supFig_count] <- fig.name
-ggsave(file.path(FigureFolder, fig.name), 
-              plot = Fig, units = 'in', width = 18, height = 12)
-SuppLegends[[supFig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent.
-A) Heatmap of patient cluster memberships based on different objectives: optimum stratification based on overall survival or progression free survival, molecular homogeneity (TCGA pan-GI subtypes), or micro-satellite instability (MSI) status.
-Columns are ordered first by membership in the optimum Overall Survival cluster membership.
-B) Pairwise Adjusted Mutual Information (AMI) of cluster labels obtained via different patient stratification methods.
-The AMI value ranges from 0 to 1 and the higher the value, the more overlap between patient partitions."
-```
-
 ```{r immunotherapy_response}
@@
 colnames(aucs)[2] <- c('AUC')
-Fig <- ggpubr::ggbarplot(aucs, x = 'method', y = 'AUC', add = 'mean_se',
+
+p1 <- ggpubr::ggbarplot(aucs, x = "method", y = 'AUC', add = 'mean_se',
                   fill = 'group') +
-  theme(axis.title.x = element_blank(), legend.position = 'none', 
-        ) + 
+  coord_flip() + 
+  theme(axis.title.y = element_blank(), legend.position = 'none') + 
   scale_fill_brewer(type = 'qual', palette = 3)
+
+# umap responders
+umap <- data.table::fread(file.path(folder, 'umap_coords_and_response.csv'))
+colnames(umap)[2:3] <- c('UMAP1', 'UMAP2')
+#drop samples without umap coords
+umap <- umap[!is.na(UMAP1)][!is.na(UMAP2)]
+umap$response <- ifelse(umap$response == TRUE, 'responder', 'non-responder')
+umap[is.na(response)]$response <- 'unavailable'
+umap[is.na(response)]$color <- 'gray'
+p2 <- ggscatter(umap, x = 'UMAP1', y = 'UMAP2', color = 'response', size = 2)  + 
+  scale_color_manual(values = c('dodgerblue', 'firebrick', 'gray')) + 
+  theme(legend.title = element_blank())
+
+# survival plots
+gbm_surv <- read.csv(file.path(folder, "survival_time_response_and_predictions.csv"), row.names = 1)
+gbm_surv <- gbm_surv[,c(3, 5:7)]
+panel_titles <- c('observed response', 'MAUI + MGMT', 
+                  'MAUI + MGMT-expression', 'MAUI + MGMT-methylation')
+plots <- sapply(simplify = F, 1:ncol(gbm_surv), function(i) {
+  x <- colnames(gbm_surv)[i]
+  plot_survival(surv, partition.df = gbm_surv[,x,drop = F], duration_col = 'PFI.time', event_col = 'PFI')[['plot']] + 
+    ggtitle(label = panel_titles[i]) + theme(legend.direction = 'vertical', legend.title = element_blank(), 
+                                             legend.position = c(0.5, 0.8))
+})
+
+Fig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2, labels = c('A', 'B')),
+                          cowplot::plot_grid(plotlist = plots, ncol = 2, labels = c('C', 'D', 'E', 'F')), 
+                          nrow = 2, rel_heights = c(1.5, 2))
 
@@
 SuppLegends[[supFig_count]] <- "Top 100 genes sorted by importance in predicting TMZ resistance in GBM patients."
+```
+
+```{r}
+# Show subtypes depend on the objective (dependent variable) - they are dynamic, not universal or carved in stone 
+# for a given cancer. There is not just one set of subtypes for a given cancer. Subtypes need to be phrased 
+# along with the dependent variable used to come up with the subtypes. 
+# Patients clustering depend on the objective of stratification
+clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
+project <- 'pan-gastrointestinal'
+datatype <- 'cnv_gex_meth_mut'
+experiment <- paste0(project, '.', datatype)
+
+M <- LFs$maui_factors[[experiment]]
+clin <- process_clinical_covariates(rownames(M))
+clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
+
+pfi_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'PFI', duration_col = 'PFI.time')
+pfi_clust <- get_kmeans_subtypes(M[,unique(pfi_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
+pfi_k <- names(which.min(sapply(colnames(pfi_clust), function(x) plot_survival(surv, pfi_clust[,x,drop = F], 'PFI.time', 'PFI', T))))
+
+os_vars <- get_prognostic_covariates(surv, clin, M, event_col = 'OS', duration_col = 'OS.time')
+os_clust <- get_kmeans_subtypes(M[,unique(os_vars$variable)], k_vals = 3:8, min_cluster_size = 0)
+os_k <- names(which.min(sapply(colnames(os_clust), function(x) plot_survival(surv, os_clust[,x,drop = F], 
+                                                                             'OS.time', 'OS', T))))
+
+# combine with known pan-gi subtype annotation
+df <- merge(pfi_clust[,pfi_k,drop = F], os_clust[,os_k,drop = F], by = 'row.names')
+df$subtype <- tcga_subtypes[match(df$Row.names, bcr_patient_barcode)]$Subtype_Selected
+df[is.na(df$subtype),]$subtype <- 'Unavailable'
+df$msi <- MSI[match(df$Row.names, bcr_patient_barcode)]$MSI
+df[is.na(df$msi),]$msi <- 'Unavailable'
+rownames(df) <- df$Row.names
+df$Row.names <- NULL
+colnames(df) <- c('Overall Survival', 'Progression Free Survival', 'Molecular Subtype', 'MSI Status')
+df <- data.frame(apply(df, 2, as.factor), check.names = F)
+df <- df[order(df[,1]),]
+m <- apply(df, 2, function(x) as.numeric(as.factor(x)))
+rownames(m) <- rownames(df)
+
+colors <- structure(1:7, names = c("1", "2", "3", "4", "5", "6", "7"))
+p1 <- Heatmap(t(m), name = "cluster", col = colors, cluster_rows = F, cluster_columns = F,
+              show_column_names = F, row_names_side = 'left',
+              column_title = 'Pan-gastrointestinal Cancer Samples\n(ordered by Overall Survival Clusters)')
+p1 <- grid::grid.grabExpr(ComplexHeatmap::draw(p1))
+# compute pairwise AMI 
+ami <- sapply(colnames(df), function(x) {
+  sapply(colnames(df), function(y) {
+    df <- df[!is.na(df[,x]) & !is.na(df[,y]),]
+    aricode::AMI(df[,x], df[,y])
+  })
+})
+
+p2 <- ggcorrplot::ggcorrplot(ami, legend.title = 'AMI', lab = T, type = 'lower', outline.color = 'white') +
+  scale_fill_gradient(limits = c(0,1), low = 'white', high = 'red') + 
+  guides(fill=guide_legend(title='AMI'))
+
+Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
+supFig_count <- supFig_count + 1
+SuppFigures[[supFig_count]] <- Fig
+fig.name <- paste0('SuppFigure', supFig_count, '.subtype_dependency.pdf')
+names(SuppFigures)[supFig_count] <- fig.name
+ggsave(file.path(FigureFolder, fig.name), 
+              plot = Fig, units = 'in', width = 18, height = 12)
+SuppLegends[[supFig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent.
+A) Heatmap of patient cluster memberships based on different objectives: optimum stratification based on overall survival or progression free survival, molecular homogeneity (TCGA pan-GI subtypes), or micro-satellite instability (MSI) status.
+Columns are ordered first by membership in the optimum Overall Survival cluster membership.
+B) Pairwise Adjusted Mutual Information (AMI) of cluster labels obtained via different patient stratification methods.
+The AMI value ranges from 0 to 1 and the higher the value, the more overlap between patient partitions."
 ```

### 2021-09-16 src/manuscript_figures.Rmd
fix panel titles for survival figures

@@
 panel_titles <- c('observed response', 'MAUI + MGMT', 
-                  'MAUI + MGMT-expression', 'MAUI + MGMT-methylation')
+                  'MGMT-expression', 'MGMT-methylation')
 plots <- sapply(simplify = F, 1:ncol(gbm_surv), function(i) {

### 2021-09-16 src/manuscript_figures.Rmd
update main figure legend for tmz-gbm analysis

@@
               plot = Fig, units = 'in', width = 10, height = 6)
-Legends[[fig_count]] <- "Classification accuracy quantified as the Area under the ROC (AUC) when predicting TMZ resistance in GBM patients using MAUI LFs alone or in combination with the MGMT expression or promoter methylation values."
+Legends[[fig_count]] <- "Predicting TMZ resistance in GBM patients.
+A) Area under the ROC (AUC) for 5-fold cross-validation predictions of response to TMZ.
+MAUI latent factors outperform both MGMT expression and MGMT promoter methylation, and the best performance is achieved by combining MAUI latent factors with MGMT expression.
+B) 2D UMAP projection of TCGA-GBM samples based on MAUI latent factors which are predictive of TMZ response.
+Responders and non-responders tend to cluster separately.
+C) Kaplan-Meier curves for progression-free interval for responders and non-responders to TMZ.
+Response defined as progression-free interval of at least six months, non- or partial-response defined as tumor progression within six months of treatment.
+D) Kaplan-Meier curves for progression-free intervals for predicted responders and non-responders using MAUI latent factors combined with MGMT expression.
+E) Kaplan-Meier curves for progression-free intervals for predicted responders and non-responders using MGMT expression.
+F) Kaplan-Meier curves for progression-free intervals for predicted responders and non-responders using MGMT promoter methylation."
 

### 2021-09-17 src/common_functions.R
update tumor purity function to make it more flexible

@@ -1131,7 +1131,7 @@ process_clinical_covariates <- function(patientIDs) {
 }
 
 
-get_tumor_purity_estimates <- function(patientIDs) {
+get_tumor_purity_estimates <- function(patientIDs = NULL) {
   dt <- data.table::data.table(TCGAbiolinks::Tumor.purity)
   dt <- data.table('Sample.ID' = dt$Sample.ID, 
              'ESTIMATE' = as.numeric(gsub(",", ".", dt$ESTIMATE)), 
@@ -1139,7 +1139,11 @@ get_tumor_purity_estimates <- function(patientIDs) {
              'LUMP' = as.numeric(gsub(",", ".", dt$LUMP)), 
              'IHC' = as.numeric(gsub(",", ".", dt$IHC)), 
              'CPE' = as.numeric(gsub(",", ".", dt$CPE)), 
+             'project' = paste0('TCGA-', dt$Cancer.type), 
              'bcr_patient_barcode' = sub("^(TCGA.{8}).+?$", "\\1", dt$Sample.ID))
+  if(is.null(patientIDs)) {
+    return(dt)
+  }
   return(dt[bcr_patient_barcode %in% patientIDs])
 }
 

### 2021-09-17 src/manuscript_figures.Rmd
add tumor purity estimation comparison plot and update figure legend

@@
 
-p2 <- ggplot(stats[purity_method == 'CPE'][grep('TCGA', project)], 
+# compute other tools' correlations with IHC
+tp <- get_tumor_purity_estimates()
+absolute <- tp[!is.na(ABSOLUTE), mean(cor(ABSOLUTE, IHC, use = 'complete.obs')),by = project]
+estimate <- tp[!is.na(ESTIMATE), mean(cor(ESTIMATE, IHC, use = 'complete.obs')),by = project]
+lump <- tp[!is.na(LUMP), mean(cor(LUMP, IHC, use = 'complete.obs')),by = project]
+
+dt <- stats[purity_method == 'IHC']
+colnames(dt)[3] <- 'MAUI'
+dt$ABSOLUTE <- absolute[match(dt$project, project)]$V1
+dt$LUMP <- lump[match(dt$project, project)]$V1
+dt$ESTIMATE <- estimate[match(dt$project, project)]$V1
+
+mdt <- melt(dt, measure.vars = c('MAUI', 'ABSOLUTE', 'ESTIMATE', 'LUMP'), id.vars = 'project')
+p2 <- ggboxplot(mdt, x = 'variable', y = 'value', add = 'jitter', color = 'variable') +
+  theme(legend.position = 'none', axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) + 
+  labs(y = 'Average Correlation with IHC')
+
+p3 <- ggplot(stats[purity_method == 'CPE'][grep('TCGA', project)], 
        aes(x = reorder(project, COR), y = COR, fill = COR)) + 
@@
 # correlation plots for top cancer types
-selected <- stats[purity_method == 'IHC'][grep('TCGA', experiment)][order(COR, decreasing = T)][1:4]$experiment
+selected <- stats[purity_method == 'IHC'][grep('TCGA', experiment)][order(COR, decreasing = T)][1:2]$experiment
 plots <- lapply(selected, function(x) {
@@
 
-p3 <- cowplot::plot_grid(plotlist = plots, ncol = 1)
+p4 <- cowplot::plot_grid(plotlist = plots, ncol = 1)
 
 # supplementary figure
-supFig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2, labels = c('A', 'B')), 
-                             p3, nrow = 2, labels = c('', 'C'), 
-                             rel_heights = c(2, 5))
+supFig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2, p3, labels = c('A', 'B', 'C'), nrow = 1), 
+                             p4, nrow = 2, labels = c('', 'D'))
 
@@
 A) Correlation of the MAUI predictions of IHC versus the annotated IHC values (immunohistochemistry-based tumor purity estimation). 
-B) Correlation of the MAUI predictions of CPE versus the annotated CPE levels (CPE stands for consensus purity estimated by computational methods: ABSOLUTE, ESTIMATE, and LUMP). 
-C) Scatterplots of MAUI predictions versus the annotated IHC and CPE values for the 
-top best predicted cancer types."
+B) Comparison of average correlations between computational estimates of IHC values by different tools. Each point represents the 
+value obtained for each of the TCGA cohorts seen in panel A. ABSOLUTE tool lacks estimates for 6 out of 14 cancer types. 
+C) Correlation of the MAUI predictions of CPE versus the annotated CPE levels (CPE stands for consensus purity estimated by computational methods: ABSOLUTE, ESTIMATE, and LUMP). 
+D) Scatterplots of MAUI predictions versus the annotated IHC and CPE values for the 
+top two best predicted cancer types with respect to annotated IHC values."
 ```

### 2021-09-17 src/manuscript_figures.Rmd
update lung cancer biomarker analysis - use the decision tree to filter candidate markers

@@
 labels <- gsub("TCGA-", "", surv[match(colnames(A.subM), bcr_patient_barcode)]$project)
-
-# further filter top features 
-top_features <- get_factor_specific_variables(t(A.subM), labels)
-# get top 5 per subtype
-top_markers <- get_top_by_label(top_features[padj < 0.05][order(padj)], 
-                                'ref_cl', topN = 5)
-# map ids 
-top_markers <- merge(top_markers, map_ids(top_markers$variable, cpg2hgnc, ens2hgnc), by = 'variable')
-
 # use treeheatr and partykit package to compute/visualize decision tree
+
+top_markers <- map_ids(names(markers), cpg2hgnc, ens2hgnc)
+# get unique table 
+top_markers <- top_markers[match(top_markers[,head(variable, 1), by = c('type', 'hgnc')]$V1, variable)]
+
 df <- data.frame(t(round(A.subM[top_markers$variable,], 3)))
@@
 df.test <- df[-inTrain[[1]],]
+# further filter top features: build initial tree to get subset of most important features
+x = partykit::ctree(subtype ~ ., data = df.train)
+selected <- names(partykit::varimp(x))
+df.train <- df.train[,c(selected, 'subtype')]
+df.test <- df.test[,c(selected, 'subtype')]
 p5 <- treeheatr::heat_tree(
@@
   theme(legend.title = element_blank(), axis.title.y = element_blank(), 
-        plot.title = element_text(hjust = 0.5))
+        plot.title = element_text(hjust = 0.5), 
+        legend.key.size = unit(0.2, units = 'in'))
 

### 2021-09-22 src/common_functions.R
set seed and use repeated cross-validation for regression analysis

@@ -395,6 +395,7 @@ predict_numerical_glmnet <- function(M, y, partition = 0.6) {
   df <- data.frame(M)
   df$y <- y
   df <- df[!is.na(df$y),]
+  set.seed(3456)
   intrain <- caret::createDataPartition(y = df$y, p = partition, list = FALSE)
   training <- df[intrain,]
   testing <- df[-intrain,]
@@ -402,7 +403,8 @@ predict_numerical_glmnet <- function(M, y, partition = 0.6) {
   set.seed(123)
   model <- caret::train(
     y ~., data = training, method = "glmnet",
-    trControl = trainControl("cv", number = 5, allowParallel = F),
+    trControl = trainControl("repeatedcv", repeats = 5, 
+                             number = 5, allowParallel = F),
     tuneLength = 10
   )
   # Best tuning parameter

### 2021-09-22 src/manuscript_figures.Rmd
change figure size in pdf

@@
 ggsave(file.path(FigureFolder, fig.name),
-              plot = Fig, units = 'in', width = 10, height = 6)
+              plot = Fig, units = 'in', width = 9, height = 11)
 Legends[[fig_count]] <- "Predicting TMZ resistance in GBM patients.

### 2021-09-22 src/manuscript_figures.Rmd
update figure legends: replace latent factors with multi-omic fingerprints

@@
                                            'pancancer', 'cnv_gex_meth_mut', factors = ct, 
-                                           top_lfs = 1, top_features = 500) 
+                                           top_lfs = 1, top_features = 100) 
 p5 <- p5 + labs(y = 'Relative Contribution\nto top latent factor') + 
@@
               plot = Fig, units = 'in', width = 12, height = 12)
-Legends[[fig_count]] <- " A) tSNE plot of MAUI latent factors (using top 5 latent factors enriched per cancer type) learned from pan-cancer dataset  B) Percentage of top feature types contributing to top predictive maui latent factor per cancer type.
+Legends[[fig_count]] <- " A) tSNE plot of MAUI multi-omics fingerprints (using top 5 fingerprint enriched per cancer type) learned from pan-cancer dataset  B) Percentage of top 100 features by omics types contributing to top predictive MAUI fingerprint per cancer type.
 C, D) Comparison of unsupervised clustering performances based on cluster homogeneity (Silhouette) in C and cluster stability (Average Proportion of Non-Overlap (APN) in D.
@@
 level of classification accuracy.
-C) Top maui LFs detected per cancer type.
-Size/color/opacity of the points represent the average neural activation value of the LF for the corresponding cancer type samples."
+C) Top MAUI latent factors (multi-omic fingerprints) detected per cancer type.
+Size/color/opacity of the points represent the average neural activation value of the MAUI fingerprint for the corresponding type of tumor samples."
 ```
@@
        plot = Fig, width = 8.27, height = 11.69, units = 'in')
-Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated LFs per cancer type.
-B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (age + gender + tumor stage) or using clinical factors along with survival-predictive MAUI LFs."
+Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated multi-omic fingerprints per cancer type.
+B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (age + gender + tumor stage) or using clinical factors along with survival-predictive MAUI fingerprints"
 
@@
               plot = Fig, units = 'in', width = 8.27, height = 11.69)
-SuppLegends[[supFig_count]] <- "Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive LFs per cancer."
+SuppLegends[[supFig_count]] <- "Kaplan-Meier survival curves adjusted for clinical factors (tumor stage + age + gender) stratified by the neural activation values of top survival-predictive multi-omic fingerprints per cancer."
 ```
@@
 Legends[[fig_count]] <- "Latent factors predictive of MSI status (high: MSI-H, stable: MSI-S, low: MSI-L) in pan-gastrointestinal cancers A) Comparison of top latent factor per MSI status for each tool in terms of adjusted p-values reflecting the differences between MSI subgroups.
-B) ROC curves for prediction accuracy of MSI status using MAUI latent factors C) tSNE plots of samples generated using top MSI-predictive LFs (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI latent factor LF25, which is most predictive of MSI-H samples.
-E) Distribution of MAUI latent factor LF25 across MSI subgroups in the pan-gastrointestinal cancer samples.
-F) Relative abundance of omics feature types among the top 500 features contributing to MAUI latent factor LF25."
+B) ROC curves for prediction accuracy of MSI status using MAUI latent factors (multi-omic fingerprints) C) tSNE plots of samples generated using top MSI-predictive fingerprints (5 LF per MSI class) colored by MSI status D) tSNE plots of samples colored by MAUI fingerprint LF25, which is most predictive of MSI-H samples.
+E) Distribution of MAUI fingerprint LF25 across MSI subgroups in the pan-gastrointestinal cancer samples.
+F) Relative abundance of omics feature types among the top 500 features contributing to MAUI fingerprint LF25."
 ```
@@
 
-Legends[[fig_count]] <- "Classification of Non-Small-Cell Lung Adenocarcinoma samples using MAUI latent factors.
+Legends[[fig_count]] <- "Classification of Non-Small-Cell Lung Adenocarcinoma samples using MAUI multi-omic fingerprints.
 A) ROC curves with AUC values on testing data (40% of the dataset) using an Elastic Net model with 5-fold cross-validation.
-B) t-SNE plot of MAUI LFs colored by cancer types.
-C) t-SNE plot of MAUI LFs colored by
-the top most-predictive latent factor for TCGA-LUSC samples. D) t-SNE plot of MAUI LFs colored by the top 
-most predictive latent factor for TCGA-LUAD samples. E) Decision tree visualisation of top LF-associated omics markers detected that can be used to distinguish LUSC and LUAD subtypes."
+B) t-SNE plot of MAUI fingerprints colored by cancer types.
+C) t-SNE plot of MAUI fingerprints colored by
+the top most-predictive fingerprint for TCGA-LUSC samples. D) t-SNE plot of MAUI fingerprints colored by the top 
+most predictive fingerprint for TCGA-LUAD samples. E) Decision tree visualisation of top MAUI fingerprint-associated omics markers detected that can be used to distinguish LUSC and LUAD subtypes."
 ```
@@
 B) Comparison of top latent factors predictive of immune phenotypes (inflamed/excluded/desert).
-C)  Comparison of top latent factors with respect to their correlation with tumor mutation burden status of the samples in the cohort D) Survival stratification of the cohort based activation score of top MAUI latent factor E) Average value of top MAUI LFs predictive of response to anti-PD-L1 response groups.
+C)  Comparison of top latent factors with respect to their correlation with tumor mutation burden status of the samples in the cohort D) Survival stratification of the cohort based activation score of top MAUI latent factor (fingerprint) E) Average value of top MAUI latent factors (fingerprints) predictive of response to anti-PD-L1 response groups.
 F) Comparison of marker combinations in terms of accuracy in predicting anti-PD-L1 response. TMB: tumor mutation burden, 
-CD8_Teff: 12 gene signature for CD8+ effector T-cells, F_TBRS: 4-gene pan-fibroblast TGFβ response signature, MAUI: top predictive MAUI LFs as seen in panel E).
+CD8_Teff: 12 gene signature for CD8+ effector T-cells, F_TBRS: 4-gene pan-fibroblast TGF-beta response signature, MAUI: top predictive MAUI fingerprints as seen in panel E).
 Each bar plot represents prediction accuracy (AUC) when using the corresponding subset of variables in an Elastic Net model."
@@
 Legends[[fig_count]] <- "Predicting TMZ resistance in GBM patients.
-A) Area under the ROC (AUC) for 5-fold cross-validation predictions of response to TMZ.
-MAUI latent factors outperform both MGMT expression and MGMT promoter methylation, and the best performance is achieved by combining MAUI latent factors with MGMT expression.
-B) 2D UMAP projection of TCGA-GBM samples based on MAUI latent factors which are predictive of TMZ response.
+A) Area under the ROC curve (AUC) for 5-fold cross-validation predictions of response to TMZ.
+MAUI multi-omic fingerprints outperform both MGMT expression and MGMT promoter methylation, and the best performance is achieved by combining MAUI fingerprints with MGMT expression.
+B) 2D UMAP projection of TCGA-GBM samples based on MAUI multi-omic fingerprints which are predictive of TMZ response.
 Responders and non-responders tend to cluster separately.
@@
 Response defined as progression-free interval of at least six months, non- or partial-response defined as tumor progression within six months of treatment.
-D) Kaplan-Meier curves for progression-free intervals for predicted responders and non-responders using MAUI latent factors combined with MGMT expression.
+D) Kaplan-Meier curves for progression-free intervals for predicted responders and non-responders using MAUI multi-omic fingerprints combined with MGMT expression.
 E) Kaplan-Meier curves for progression-free intervals for predicted responders and non-responders using MGMT expression.

### 2021-09-24 src/manuscript_figures.Rmd
add a panel about LGG clinical feature associated latent factors

@@
 library(partykit)
-
 theme_set(ggpubr::theme_pubclean())
+
 geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
@@
   scale_color_gradient(low = 'blue', high = 'red') + 
-  labs(x = '', y = 'clinical factor')
+  theme(axis.title = element_blank())
+
+# make example plots for selected clinical features
+pr <- 'TCGA-LGG' 
+datatype <- 'cnv_gex_meth_mut'
+M <- LFs$maui_factors[[paste0(pr, '.', datatype)]]
+clin <- process_clinical_covariates(rownames(M))
+dt.pr <- dt[padj < 0.05][project == pr][order(padj)]
+# select top 3 clinical features for the project
+plots <- lapply(head(dt.pr$label, 3), function(x) {
+  # get lf, clinical feature and plot distribution 
+  f <- gsub("[\n ]", "_", x)
+  lf <- dt.pr[label == x]$variable
+  df <- cbind(data.frame(M[,lf,drop = F]), clin[,f,drop=F])
+  df <- df[!is.na(df[,f]),]
+  ggpubr::ggboxplot(df, y = lf, x = f, 
+                    color = f, add = 'jitter') + 
+    labs(x = x) + 
+    theme(legend.position = 'none', axis.text.x = element_text(angle = 45, hjust = 1)) 
+})
+
+p2 <- cowplot::plot_grid(plotlist = plots, nrow = 1, align = 'h')
 
@@
 dt <- melt(ci, id.vars = 'project')
-dt$variable <- ifelse(dt$variable == 'base', 'Age+gender+stage', 'Age+gender+stage + MAUI LFs')
-
-p2 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
+dt$variable <- ifelse(dt$variable == 'base', 'Age+gender+stage', 'Age+gender+stage + MAUI fingerprints')
+
+p3 <- ggplot(ci, aes(x = reorder(project, with_lf))) + 
   geom_segment(aes(x = reorder(project, with_lf), xend = project, 
@@
 
-Fig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2, nrow = 2, labels = c('A', 'B')))
+Fig <- cowplot::plot_grid(p1, 
+                          cowplot::plot_grid(p2, p3, 
+                                             nrow = 1, 
+                                             labels = c('B', 'C'), 
+                                             rel_widths = c(2, 2)), 
+                          labels = c('A', ''), 
+                          nrow = 2)
 
@@
 Legends[[fig_count]] <- "A) Top significant (padj < 0.05) clinical factor-associated multi-omic fingerprints per cancer type.
-B) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (age + gender + tumor stage) or using clinical factors along with survival-predictive MAUI fingerprints"
+B) MAUI fingerprints enriched for subgroups of different clinical features of TCGA-LGG samples.
+LF45 is enriched among samples histologically categorized as Oligodendrogliomas; LF54 is enriched among donors of male gender; LF65 is enriched among samples with histological grade of G3.
+C) Harrell’s C-index for progression-free survival prediction accuracy computed using only clinical factors (age + gender + tumor stage) or using clinical factors along with survival-predictive MAUI fingerprints"
 

### 2021-09-27 src/compile_figures.sh
add script for compiling figures

@@ -1,0 +1,7 @@
+RSCRIPT=$1
+RMD=$2 # path the Rmd file
+settings_file=$3 # settings file used to run the pipeline
+figures_folder=$4  # where to export pdf figures
+workdir=$5 # where to save docx file
+
+${RSCRIPT} -e "library(rmarkdown); rmarkdown::render(\"${RMD}\", output_dir = \"${workdir}\", params = list(settings_file = \"${settings_file}\", figures_folder = \"${figures_folder}\", workdir = \"${workdir}\"))"

### 2021-09-27 src/score_genesets_tcga.R
remove script from repo - unneeded

@@ -1,141 +1,0 @@
-# use ssGSEA approach to score each TCGA sample (using gene expression datasets)
-# 
-library(singscore)
-library(data.table)
-source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
-geneSetFolder <- '/data/local/buyar/arcas/subtyping_paper/data/GeneSets'
-ens2hgnc <- readRDS('/data/local/buyar/arcas/subtyping_paper/data/ens2hgnc.RDS') #id mapping 
-GDC_data_dir <- '/data/local/buyar/pancancer_diagnostics/data/GDCdata_prepared' # path to TCGA data 
-outDir <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
-
-# f: path to gmt format file 
-read_msigdb <- function(f) {
-  geneSets <- lapply(readLines(f), function(x) {
-    unlist(strsplit(x, "\t"))
-  })
-  #add names
-  names(geneSets) <- unlist(lapply(geneSets, function(x) x[1]))
-  #remove first two 
-  geneSets <- lapply(geneSets, function(x) x[-c(1:2)])
-  return(geneSets)
-}
-# d: path to folder that contains cancerSEA cell state signature files 
-read_cancersea <- function(d) {
-  files <- dir(d, '.txt$', full.names = T)
-  geneSets <- lapply(files, 
-                               function(f) {
-                                 dt <- data.table::fread(f)
-                                 return(unique(dt$EnsembleID))
-                               })
-  names(geneSets) <- gsub(".txt$", "", basename(files))
-  return(geneSets)
-}
-
-# f: path to file containing immune gene sets used in xCell package 
-read_xCell_geneSets <- function(f, group_by_cell_type = FALSE) {
-  l <- readLines(f)
-  geneSets <- lapply(l[-1], function(x) {
-    res <- unlist(strsplit(gsub(" ", "_", x), "\t"))
-    return(setdiff(res[-c(1:2)], ''))
-  })
-  
-  names(geneSets) <- unlist(lapply(l[-1], function(x) 
-    gsub(" ", "_", unlist(strsplit(x, "\t"))[1])))
-  
-  # there are multiple sources for cell types, which can be grouped together 
-  if(group_by_cell_type == TRUE) {
-    dt <- data.table('name' = names(geneSets), 
[…]

### 2021-09-27 src/download.tcga.R
add script for omics data download from tcga

@@ -1,0 +1,89 @@
+# use TCGAbiolinks package to download open-access data from GDC.
+# See vignettes for TCGAbiolinks package:
+# http://bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html
+# download: http://bioconductor.org/packages/release/bioc/vignettes/TCGAbiolinks/inst/doc/download_prepare.html
+
+library(TCGAbiolinks)
+
+args <- commandArgs(trailingOnly = TRUE)
+
+#workdir for downloading GDCdata
+downloaddir <- args[1]
+#output folder for writing prepared data after downloads
+outDir <- args[2]
+
+pbapply::pboptions(type = 'txt')
+
+get_Data <- function(project, outDir, dataType, ...) {
+  if(dataType == 'mut') {
+    outFile <- file.path(outDir, 
+                         paste0(project, 
+                                '.maf'))
+    if(!file.exists(outFile)) {
+      maf <- GDCquery_Maf(tumor = gsub('TCGA-', '', project), 
+                            pipelines = 'muse', save.csv = FALSE)
+      write.table(x = maf, file = outFile, 
+                  sep = '\t', 
+                  row.names = FALSE, 
+                  quote = FALSE)
+    }
+    return(outFile)
+  } else if (dataType == 'clinical') {
+    outFile <- file.path(outDir, paste0(project, '.clinical.tsv'))
+    if(!file.exists(outFile)) {
+      clinical <- GDCquery_clinic(project = project, type = "clinical")
+      write.table(x = clinical, file = outFile, sep = '\t', quote = FALSE, row.names = FALSE)
+    }
+    return(outFile)
+  }else {
+    outFile <- file.path(outDir, 
+                         paste(project, dataType, 'RDS', sep = '.'))
+    if(!file.exists(outFile)) {
+      query <- GDCquery(project = project, ...)
+      GDCdownload(query = query)
+      dat <- GDCprepare(query = query)
+      saveRDS(object = dat, file = outFile)
+    }
+    return(outFile)
+  } 
+}
+
+projects <- TCGAbiolinks:::getGDCprojects()$project_id
+projects <- projects[grepl('^TCGA',projects,perl=T)]
+
+setwd(downloaddir)
+
+cl <- parallel::makeCluster(2, outfile = 'download.tcga.log')
+parallel::clusterExport(cl = cl, varlist = c('get_Data', 'outDir'))
+
+parallel::parLapply(cl = cl, X = projects[1:2], fun = function(pr) {
+  require("TCGAbiolinks")
[…]

### 2021-09-27 src/download.tcga.MSI.R
add script for MSI data download from TCGA

@@ -1,0 +1,40 @@
+# import MSI data for TCGA patients 
+# see: https://bioconductor.org/packages/release/bioc/vignettes/TCGAbiolinks/inst/doc/clinical.html#Microsatellite_data
+library(TCGAbiolinks)
+
+args <- commandArgs(trailingOnly = TRUE)
+
+downloaddir <- args[1]
+outDir <- args[2]
+setwd(downloaddir)
+
+projects <- TCGAbiolinks:::getGDCprojects()$project_id
+projects <- projects[grepl('^TCGA',projects,perl=T)]
+
+
+cl <- parallel::makeCluster(20, outfile = 'download.tcga.MSI.log')
+parallel::clusterExport(cl = cl, varlist = c('outDir'))
+parallel::parLapply(cl = cl, X = projects, fun = function(pr) {
+  require("TCGAbiolinks")
+  message(date()," => Processing project:",pr)
+  query <- tryCatch (
+    {
+      GDCquery(project = pr, 
+                    data.category = "Other",
+                    legacy = TRUE,
+                    access = "open",
+                    data.type = "Auxiliary test")  
+    }, error = function(cond) {
+      return(NULL)
+    }
+  )
+  if(!is.null(query)) {
+    GDCdownload(query)
+    msi_results <- GDCprepare_clinic(query, "msi")
+    write.table(x = msi_results, quote = F, sep = '\t',
+                file = file.path(outDir, paste0(pr, ".MSI.tsv")))
+  }
+})
+parallel::stopCluster(cl)
+
+

### 2021-09-27 src/download.tcga.firehose.R
add script for CNV data download from Broad Firehose

@@ -1,0 +1,65 @@
+# Use RTCGAtoolbox package to download processed TCGA datasets such as CNV - GISTIC
+
+library(RTCGAToolbox)
+library(TCGAbiolinks)
+
+args <- commandArgs(trailingOnly = TRUE)
+
+pbapply::pboptions(type = 'txt')
+
+projects <- TCGAbiolinks:::getGDCprojects()$project_id
+projects <- projects[grepl('^TCGA',projects,perl=T)]
+
+#download location for firehose data
+downloadDir <- args[1] 
+preparedDataDir <- args[2] 
+
+# download TCGA CNV data processed using GISTIC served at BROAD
+lastDate <- RTCGAToolbox::getFirehoseAnalyzeDates(last = 1)
+cl <- parallel::makeCluster(11)
+parallel::clusterExport(cl, varlist = c('lastDate', 'projects', 'downloadDir'))
+pbapply::pblapply(X = gsub('TCGA-', '', projects), 
+                  FUN = function(pr) {
+                    message("\n", date(), " => Downloading data for project: ",pr)
+                    if(!file.exists(file.path(downloadDir, 
+                                             paste(lastDate, pr, 
+                                                   'all_data_by_genes.txt', 
+                                                   sep = '-')))) {
+                      g <- RTCGAToolbox::getFirehoseData(dataset = pr,
+                                                         gistic2Date = lastDate,
+                                                         GISTIC = TRUE,
+                                                         destdir = downloadDir)
+                    } else {
+                      message("Data for project ",pr," already downloaded!")
+                    }
+                  })
+parallel::stopCluster(cl)
+
+#NOTE 05.12.2018: SKCM and LAML didn't return the tables for 
+#all_data_by_genes.txt, only returned clinical data. 
+
+# process each file and save as RDS files 
+assayFiles <- dir(downloadDir, 'all_data_by_genes.txt$', full.names = T)
+gistic.assays <- pbapply::pbsapply(simplify = FALSE,  
+                          X = assayFiles, function(f) {
+                            dt <- data.table::fread(f, header = T, sep = '\t')
+                            dt <- dt[,-c('Locus ID', 'Cytoband')]
+                            M <- as.matrix(dt[,-1])
+                            rownames(M) <- dt$`Gene Symbol`
+                            return(M)
+                          })
+names(gistic.assays) <- paste0("TCGA-", 
[…]

### 2021-09-27 src/prepare_data.immther.R
add script to prepare anti-pd-l1 data processing

@@ -1,0 +1,48 @@
+# prepare data for maui 
+# download data from Mariathasan paper
+# from http://research-pub.gene.com/IMvigor210CoreBiologies/packageVersions/IMvigor210CoreBiologies_0.1.13.tar.gz
+args <- commandArgs(trailingOnly = TRUE)
+
+sourceFolder <- args[1] #path to source code of IMvigor210CoreBiologies
+geneSetFile <- args[2] # path to file with gene names to be included in the analysis
+
+library(SummarizedExperiment)
+# load Rdata objects from the downloaded package
+load(file.path(sourceFolder, 'data', 'cds.RData'))
+
+counts <- counts(cds) # raw count table
+colData <- pData(cds) # sample metadata 
+entrez2hdnc <- data.table(fData(cds)) # gene name mapping from entrez to HGNC symbols
+
+normCounts <- sapply(1:nrow(colData), function(i) {
+  counts[,i]/colData$sizeFactor[i]
+})
+colnames(normCounts) <- colnames(counts)
+
+# filter for most variable hallmark + xCell gene sets:
+genes <- unique(readLines(geneSetFile)) 
+rownames(normCounts) <- entrez2hdnc[match(rownames(normCounts), entrez_id)]$Symbol
+normCounts <- normCounts[rownames(normCounts) %in% genes,]
+which(table(rownames(normCounts)) > 1)
+# drop row with non-unique names
+normCounts <- normCounts[-match(names(which(table(rownames(normCounts)) > 1)), 
+                                rownames(normCounts)),]
+# pick top 5000 genes by sd
+normCounts <- normCounts[names(sort(apply(normCounts, 1, sd), decreasing = T)[1:5000]),]
+
+# scale and save count table
+dir.create(file.path(getwd(), 'assays_immther'))
+out <- file.path(getwd(), 'assays_immther', "mariathasan.gex")
+write.table(x = scale(log(normCounts+1)), 
+            file = paste0(out, '.csv'), quote = F, row.names = T, sep = ',')
+
+saveRDS(list('gex' = scale(log(normCounts+1))), paste0(out, '.RDS')) 
+write.table(x = colData, 
+            file = file.path(getwd(), 'mariathasan.colData.csv'), 
+            quote = F, row.names = T, sep = ',')
+
+
+
+
+
+

### 2021-09-27 src/prepare_data.immther.R
update script to print all files under assays_immther

@@ -27,18 +27,25 @@ which(table(rownames(normCounts)) > 1)
 # drop row with non-unique names
 normCounts <- normCounts[-match(names(which(table(rownames(normCounts)) > 1)), 
                                 rownames(normCounts)),]
+
+
+# create folder to save processed data
+dir.create(file.path(getwd(), 'assays_immther'))
+
+# Save all normalized genes as RDS
+saveRDS(scale(log(normCounts+1)), file.path(getwd(), 'assays_immther', 'mariathasan.gex.all_genes.normalised.RDS'))
+
 # pick top 5000 genes by sd
 normCounts <- normCounts[names(sort(apply(normCounts, 1, sd), decreasing = T)[1:5000]),]
 
-# scale and save count table
-dir.create(file.path(getwd(), 'assays_immther'))
+# scale and save count table for analysis by dimension reducction tools
 out <- file.path(getwd(), 'assays_immther', "mariathasan.gex")
 write.table(x = scale(log(normCounts+1)), 
             file = paste0(out, '.csv'), quote = F, row.names = T, sep = ',')
 
 saveRDS(list('gex' = scale(log(normCounts+1))), paste0(out, '.RDS')) 
 write.table(x = colData, 
-            file = file.path(getwd(), 'mariathasan.colData.csv'), 
+            file = file.path(getwd(), 'assays_immther', 'mariathasan.colData.csv'), 
             quote = F, row.names = T, sep = ',')
 
 

### 2021-09-28 src/download.tcga.R
download all projects

@@ -53,10 +53,10 @@ projects <- projects[grepl('^TCGA',projects,perl=T)]
 
 setwd(downloaddir)
 
-cl <- parallel::makeCluster(2, outfile = 'download.tcga.log')
+cl <- parallel::makeCluster(10, outfile = 'download.tcga.log')
 parallel::clusterExport(cl = cl, varlist = c('get_Data', 'outDir'))
 
-parallel::parLapply(cl = cl, X = projects[1:2], fun = function(pr) {
+parallel::parLapply(cl = cl, X = projects, fun = function(pr) {
   require("TCGAbiolinks")
   message(date()," => Processing project:",pr)
 

### 2021-09-29 src/prepare_data.immther.R
save data and assays in separate folder

@@ -3,6 +3,8 @@
 # from http://research-pub.gene.com/IMvigor210CoreBiologies/packageVersions/IMvigor210CoreBiologies_0.1.13.tar.gz
 args <- commandArgs(trailingOnly = TRUE)
 
+library(data.table)
+
 sourceFolder <- args[1] #path to source code of IMvigor210CoreBiologies
 geneSetFile <- args[2] # path to file with gene names to be included in the analysis
 
@@ -31,9 +33,10 @@ normCounts <- normCounts[-match(names(which(table(rownames(normCounts)) > 1)),
 
 # create folder to save processed data
 dir.create(file.path(getwd(), 'assays_immther'))
+dir.create(file.path(getwd(), 'data_immther'))
 
 # Save all normalized genes as RDS
-saveRDS(scale(log(normCounts+1)), file.path(getwd(), 'assays_immther', 'mariathasan.gex.all_genes.normalised.RDS'))
+saveRDS(scale(log(normCounts+1)), file.path(getwd(), 'data_immther', 'mariathasan.gex.all_genes.normalised.RDS'))
 
 # pick top 5000 genes by sd
 normCounts <- normCounts[names(sort(apply(normCounts, 1, sd), decreasing = T)[1:5000]),]
@@ -45,11 +48,8 @@ write.table(x = scale(log(normCounts+1)),
 
 saveRDS(list('gex' = scale(log(normCounts+1))), paste0(out, '.RDS')) 
 write.table(x = colData, 
-            file = file.path(getwd(), 'assays_immther', 'mariathasan.colData.csv'), 
+            file = file.path(getwd(), 'data_immther', 'mariathasan.colData.csv'), 
             quote = F, row.names = T, sep = ',')
-
-
-
 
 
 

### 2021-09-29 src/prepare_data.immther.R
change file names for assays/output

@@ -32,23 +32,23 @@ normCounts <- normCounts[-match(names(which(table(rownames(normCounts)) > 1)),
 
 
 # create folder to save processed data
-dir.create(file.path(getwd(), 'assays_immther'))
-dir.create(file.path(getwd(), 'data_immther'))
+dir.create(file.path(getwd(), 'assays'))
+dir.create(file.path(getwd(), 'data'))
 
 # Save all normalized genes as RDS
-saveRDS(scale(log(normCounts+1)), file.path(getwd(), 'data_immther', 'mariathasan.gex.all_genes.normalised.RDS'))
+saveRDS(scale(log(normCounts+1)), file.path(getwd(), 'data', 'mariathasan.gex.all_genes.normalised.RDS'))
 
 # pick top 5000 genes by sd
 normCounts <- normCounts[names(sort(apply(normCounts, 1, sd), decreasing = T)[1:5000]),]
 
 # scale and save count table for analysis by dimension reducction tools
-out <- file.path(getwd(), 'assays_immther', "mariathasan.gex")
+out <- file.path(getwd(), 'assays', "mariathasan.gex")
 write.table(x = scale(log(normCounts+1)), 
             file = paste0(out, '.csv'), quote = F, row.names = T, sep = ',')
 
 saveRDS(list('gex' = scale(log(normCounts+1))), paste0(out, '.RDS')) 
 write.table(x = colData, 
-            file = file.path(getwd(), 'data_immther', 'mariathasan.colData.csv'), 
+            file = file.path(getwd(), 'data', 'mariathasan.colData.csv'), 
             quote = F, row.names = T, sep = ',')
 
 

### 2021-09-29 src/manuscript_figures.Rmd
use parameters to pass experiment-related settings/projdir files as parameters. use R script to compile

@@
 params:
-  settings_file: ''
-  figures_folder: '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas/test'
+  settings_file: 'settings.yaml' # settings file for pancancer analysis
+  immther_projdir: 'immunotherapy_analysis' # settings file for immunotherapy analysis 
+  tmz_gbm_figure_data_folder: 'data/tmz_gbm_figure_data' # path to data folder for tmz-gbm analysis 
+  figures_folder: 'figures' 
   workdir: '.'
@@
 
-```{r setup, include=FALSE}
+```{r knitr_setup, include=FALSE}
 knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE, 
-                      fig.width = 9, fig.height = 11)
+                      fig.width = 9, fig.height = 11, eval = TRUE)
 knitr::opts_knit$set(root.dir = params$workdir)
+```
+
+```{r setup}
 settings <- yaml::read_yaml(params$settings_file)
-message(settings$GDCdataDir)
 FigureFolder <- params$figures_folder
@@
 theme_set(ggpubr::theme_pubclean())
-
-geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
-
 inputDir <- settings$pipeline_output
@@
                       function(x) gsub(' ', '', unlist(strsplit(x, ','))))
+
 tcga_subtypes <- process_tcga_subtypes()
@@
 # immunotherapy response prediction analysis: 
-# TODO: pass paths via args
-projDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/immunotherapy_response/Mariathasan_et_al'
-settings_imtrp <- yaml::read_yaml(file.path(projDir, 'settings.yaml'))
-gex_all <- readRDS(file.path(projDir, 'mariathasan.gex.all_genes.normalised.RDS'))
-human_gene_signatures <- readRDS(file.path(projDir, 'human_gene_signatures.RDS'))
-
-LFs_imtrp <- import_LFs(c('maui_factors', 'mofa_factors', 'pca_factors'), settings_imtrp$pipeline_output)
-FWs_imtrp <- import_feature_weights(c('maui_factors'), settings_imtrp$pipeline_output)
-colData <- data.table::fread(settings_imtrp$clin)
+projDir <- params$immther_projdir 
+settings_imtrp <- yaml::read_yaml(file.path(projDir, 'settings_immther.yaml'))
+gex_all <- readRDS(file.path(projDir, 'data', 'mariathasan.gex.all_genes.normalised.RDS'))
+human_gene_signatures <- readRDS(file.path(projDir, 'data', 'human_gene_signatures.RDS'))
+
+LFs_imtrp <- import_LFs(c('maui_factors', 'mofa_factors', 'pca_factors'), 
+                        file.path(projDir, 'output'))
+FWs_imtrp <- import_feature_weights(c('maui_factors'), file.path(projDir, 'output'))
+colData <- data.table::fread(file.path(projDir, 'data', 'mariathasan.colData.csv'))
 colnames(colData)[1] <- 'bcr_patient_barcode'
@@
 # Here I import figure data to have consistent figure styles
-folder <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/tmz_gbm_jona'
+folder <- params$tmz_gbm_figure_data_folder
 

### 2021-09-29 src/compile_figures.R

@@ -1,0 +1,17 @@
+library(rmarkdown)
+
+args = commandArgs(trailingOnly = TRUE)
+
+RMD=args[1] # path the Rmd file
+settings_file=args[2] # settings file used to run the pipeline
+immther_projdir=args[3] # path to folder with immunotherapy analysis
+tmz_gbm_figure_data_folder=args[4] # path to folder with tmz-gbm figure data
+figures_folder=args[5]  # where to export pdf figures
+workdir=args[6] # where to save docx file
+
+rmarkdown::render(input = RMD, output_dir = workdir, 
+                  params = list('settings_file' = settings_file, 
+                                'immther_projdir' = immther_projdir, 
+                                'tmz_gbm_figure_data_folder' = tmz_gbm_figure_data_folder, 
+                                'figures_folder' = figures_folder,
+                                'workdir' = workdir))

### 2021-09-29 src/compile_figures.sh

@@ -1,7 +1,0 @@
-RSCRIPT=$1
-RMD=$2 # path the Rmd file
-settings_file=$3 # settings file used to run the pipeline
-figures_folder=$4  # where to export pdf figures
-workdir=$5 # where to save docx file
-
-${RSCRIPT} -e "library(rmarkdown); rmarkdown::render(\"${RMD}\", output_dir = \"${workdir}\", params = list(settings_file = \"${settings_file}\", figures_folder = \"${figures_folder}\", workdir = \"${workdir}\"))"

### 2021-09-29 src/common_functions.R
remove unused function

@@ -92,23 +92,6 @@ fit_RSF_clin <- function(clin, M, ntree = 1000) {
   return(list('fit_with' = fit1, 'fit_wo' = fit2, 'C_with' = Cindex1, 'C_wo' = Cindex2))
 }
 
-# this is wrong. 
-# 1. combat cannot handle multiple batch factors at the same time.
-# 2. it is not a good idea to sequentially remove batch effects unless they are orthogonal. 
-correct_batch_effects_wrongly <- function(M, batch.df, method = 'limma') {
-  M.c <- M #corrected matrix
-  for (b in colnames(batch.df)) {
-    batch_factors <- as.character(batch.df[colnames(M.c), b])
-    if(method == 'limma') {
-      M.c <- limma::removeBatchEffect(x = M.c, batch = batch_factors)
-    } else if (method == 'combat') {
-      M.c <- sva::ComBat(dat = M.c, batch = batch_factors)
-    }
-  }
-  return(M.c)
-}
-
-
 # Use RUVseq to correct for batch effects while preserving tumor/normal status
 # counts: raw count matrix
 # colData: batch table including at least one column that contains the factor of interest to be preserved
